### 2022-04-06 hw/1/homework_1_turnin.tex
add homework and stuff

\documentclass[a4paper]{article}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{paralist}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{paralist}
\usepackage[svgname]{xcolor}
\usepackage{enumerate}
\usepackage{array}
\usepackage{times}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{environ}
\usepackage{times}
\usepackage{textcomp}
\usepackage{caption}
\usepackage[square,numbers]{natbib}

\urlstyle{rm}

\setlength\parindent{0pt} % Removes all indentation from paragraphs

% TO SHOW SOLUTIONS, include following (else comment out):
\newenvironment{soln}{
    \leavevmode\color{blue}\ignorespaces
}{}

\hypersetup{
%    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

\geometry{
  top=1in,            % <-- you want to adjust this
  inner=1in,
  outer=1in,
  bottom=1in,
  headheight=3em,       % <-- and this
  headsep=2em,          % <-- and this
  footskip=3em,
}

\title{Assignment 1} % Title
\author{
Derek Paulsen \\
}

\date{}

\newcommand{\ind}[1]{\mathds{1}[#1]}
\def \eps{\epsilon}
\begin{document}
\maketitle

\section{Problem 1}
\subsection{Part A}
In order to describe a single hash function we need to store three numbers $a,b$ and $p$, where,
$p > |U|, 1 \geq a < p, 0 \geq b < p$, therefore we need to at most $\log_2(|U|)$ bits to store 
each number. Hence a single hash function requires $O(\log_2(|U|))$ bits. We need $n+1$ hash functions 
to describe the two level hashing hence we need $O(n \log_2(|U|))$ to describe the perfect two level 
hashing function.

\subsection{Part B}
Let the items that we are trying to hash be $S = (s_1,...,s_n)$. (the items of $S$ are inserted in order).
Suppose we have a single array of size $O(n^{1.5})$. 
Let $X_i = \ind{s_i \text{  collides with another element in the array}}$, we wish to compute $E[X_i]$.
Let $h$ be a random hash function $h(x) \rightarrow \{1,...,n^{1.5}\}$

\begin{align*}
	E[X_i] = E[\sum_{j=1}^{i-1} \ind{h(s_i) = h(s_j)}]\\
	E[X_i] = \sum_{j=1}^{i-1} E[\ind{h(s_i) = h(s_j)}]\\
	E[X_i] = \sum_{j=1}^{i-1} Pr[h(s_i) = h(s_j)]\\
	E[X_i] = \sum_{j=1}^{i-1} 1/n^{1.5}\\
	E[X_i] =  (i-1)/n^{1.5}\\
\end{align*}

suppose we have a two arrays of size $O(n^{1.5})$. The probability of a collision in this both arrays is,
Let $Y_i = \ind{s_i \text{  collides with another element in both arrays}}$, we wish to compute $E[Y_i]$.
We note that each array has at most $i-1$ elements when inserting the $i$th element, hence we have 
\begin{align*}
	E[Y_i] \leq E[X_i]^2\\
	E[Y_i] \leq (i-1)^2/n^{3}\\
\end{align*}

Let $Y = \sum_{i=1}^{n} = Y_i$, the number of collisions. We want $E[Y]$
\begin{align*}
	E[Y_i] &= E[\sum_{i=1}^{n} Y_i]\\
	E[Y_i] &= \sum_{i=1}^{n} E[Y_i]\\
	E[Y_i] &\leq \sum_{i=1}^{n} \frac{(i-1)^2}{n^3}\\
	E[Y_i] &\leq \frac{n(n-1)(2n-1)}{6n^3}\\
	E[Y_i] &\leq \frac{n^3 - 3n^2 + n}{6n^3}\\
	E[Y_i] &= O(1)\\
\end{align*}

Therefore the expected number of collisions is $O(1)$
\subsection{Part C}

We note that our work above only assumes that $x,y \in U, x\neq y, Pr[h(x) = h(y)] = 1/n^{1.5}$, which is 
satisfied by a 2-universal hash function by definition. Therefore the expected number of collisions
is still $O(1)$ when the hash functions 2-universal hash functions.
\subsection{Part D}

The algorithm procedes as follows,

\begin{enumerate}
		\item Hash all elements $s\in S$ into the first array with collisions
		\item remove elements from the first array such that each bucket has at most one element, let $C$ be the set of elements that were removed
		\item hash the elements $C$ into the second array 
		\item repeat step 3 until a perfect hashing has been found for the elements in $C$
\end{enumerate}

Let $X = \sum_{i=1}^n X_i$, we wish to compute $E[X]$ the number of expected collisions.
\begin{align*}
	E[X] &= E[\sum_{i=1}^n X_i]\\
	E[X] &= \sum_{i=1}^n E[X_i]\\
	E[X] &= \sum_{i=1}^n (i-1)/n^{1.5}\\
	E[X] &= \sum_{i=1}^n \frac{(n-1)(n-2)}{n^{1.5}}\\
	E[X] &= \frac{n^2 - 3n + 2}{n^{1.5}}\\
	E[X] &\leq \sqrt{n}
\end{align*}

We now compute the probability of collision in step 3,
using the union bound we get
\begin{align*}
	Pr[collision] &\leq \sum_{i=1}^{\sqrt{n}} X_i\\
	Pr[collision] &\leq  \frac{(\sqrt{n}-1)(\sqrt{n}-2)}{n^{1.5}}\\
	Pr[collision] &\leq \frac{n - 3\sqrt{n} + 2}{n^{1.5}}\\
	Pr[collision] &\leq 1/\sqrt{n}
\end{align*}

Therefore in expectation we will need try one hash function in step 3 to get a perfect hashing for $C$.

Step 1 requires $n$ hash functions evaluations, and step 3 requires $\sqrt{n}$ hash functions evaluations in expectation, hence the 
number of hash function evaluations to get a perfect hashing with the above algorithm is 
$O(n)$.

\subsection{Part E}
% TODO

\section{Problem 2}
\subsection{Part A}

There are $n$ choose $k$ ways to have $k$ balls in the first bucket and then $(n-1)^{n-k}$ ways to place the rest of the balls. There are $n^n$ total 
ways to place all the balls, hence we want,

\begin{align*}
	\Pr[\text{bin 1 has $k$ balls}] \geq 1/\sqrt{n}\\
	\frac{{n \choose k}(n-1)^{n-k}}{n^n} \geq 1/\sqrt{n}\\
\end{align*}
Using the fact that $(\frac{n}{k})^ k \leq {n \choose k}$

\begin{align*}
	\frac{(\frac{n}{k})^k(n-1)^{n-k}}{n^n} \geq 1/\sqrt{n}\\
	\frac{1}{k^k} \geq \frac{(\frac{n}{n-1})^{n-k}}{\sqrt n}\\
\end{align*}

using this bound we get $(1 + \frac{1}{n-1})^{n} \leq e \implies (1 + \frac{1}{n-1})^{n-k} \leq e$
\begin{align*}
	\frac{1}{k^k} \geq \frac{e}{\sqrt n}\\
	k^k \leq \frac{\sqrt n}{e}\\
\end{align*}

We now set $k = \log(n) / e\log\log(n)$,
\begin{align*}
	(\log n / e\log\log n)^k \leq \frac{\sqrt n}{e}\\
	k(\log\log n  - \log\log\log n - 1) \leq .5 \log n - 1\\
	\frac{1}{e}\log n - k(\log\log\log n + 1) \leq .5 \log n - 1\\
	True
\end{align*}

Therefore we have that the probability of bin 1 having $k$ balls is greater than or equal to $1/\sqrt{n}$

\subsection{Part B}

I don't have a formal proof for this, however here is the idea,
Let $a_i$ be the event that bin $i$ has $k$ balls.
Let $b_i$ be the event that at least one bin $i-1,...,1$ has $k$ balls.
We then have the probability that bin $i$ has $k$ given that bin $i-1,...,1$ dont have $k$ is,

\begin{align*}
	Pr(a_i | \bar b_{i-1}) &= \frac{Pr(\bar b_{i-1} | a_i)Pr(a_i)}{1 - Pr (b_{i-1})}\\
\end{align*}

We then note that $Pr(b_i)$ is strictly increasing while $Pr(a_i)$ is constant, 
therefore, $Pr(a_i | \bar b_{i-1})$ is strictly increasing.

We then have the probability that none of the bins have $k$ is,

\begin{align*}
	Pr(\bigcap_{i=1}^n \bar a_i ) &= \prod_{i=1}^n Pr(\bar a_i | \bigcap_{j=1}^{i-1} \bar a_j)\\
	Pr(\bigcap_{i=1}^n \bar a_i ) &= \prod_{i=1}^n 1 - Pr(a_i | \bigcap_{j=1}^{i-1} \bar a_j)\\
	Pr(\bigcap_{i=1}^n \bar a_i ) &= \prod_{i=1}^n 1 - Pr(a_i | \bar b_{i-1})\\
	Pr(\bigcap_{i=1}^n \bar a_i ) &\leq \prod_{i=1}^n 1 - 1/\sqrt{n}\\
	Pr(\bigcap_{i=1}^n \bar a_i ) &\leq (1 - 1/\sqrt{n})^n\\
	Pr(\bigcap_{i=1}^n \bar a_i ) &\leq (1 - 1/\sqrt{n})^n\\
	Pr(\bigcap_{i=1}^n \bar a_i ) &\leq ((1 - 1/\sqrt{n})^{\sqrt{n}})^{\sqrt{n}}\\
	Pr(\bigcap_{i=1}^n \bar a_i ) &\leq 1/e^{\sqrt{n}}\\
\end{align*}
We then use that fact that $e^{\sqrt{n}} \geq n$ 
\begin{align*}
	Pr(\bigcap_{i=1}^n \bar a_i ) &\leq 1/n\\
\end{align*}

Therefore we have that the probability of some bin having $k$ is greater than or equal to $1 - 1/n$.

\section{Problem 3}

Suppose that we have the a sample of $t$ numbers from the unordered list. Let $p = k/n$, i.e. the 
percentile of the rank that we are trying to estimate. Our estimator is the number with 
rank $tp$ in our sample.

Proof.

We begin by noting that our esimator is incorrect if and only if we have $tp$ elements which are 
less than $(1-\eps)k$ or we have $t(1-p)$ elements which are greater $(1+\eps)k$. 
Let $S = \{s_i,...,s_t\}$ be a random sample with replacement of size $t$ from $x_1,...x_n$.
Let $L_i = \ind{rank(s_i) < (1-\eps)k}$ and $L = \sum_{i=1}^{t} L_i$. We then compute the expectation,

\begin{align*}
	E[L] &= E[\sum_{i=1}^{t} L_i] \\
			&= \sum_{i=1}^{t} E[L_i]\\
			&= \sum_{i=1}^{t} (1-\eps)p\\
			&=  t(1-\eps)p\\
\end{align*}

Additionally, let 
$U_i = \ind{rank(s_i) > (1+\eps)k}$ and $U = \sum_{i=1}^{t} U_i$. We then compute the expectation,

\begin{align*}
	E[U] &= E[\sum_{i=1}^{t} U_i] \\
			&= \sum_{i=1}^{t} E[U_i]\\
			&= \sum_{i=1}^{t} (1-(1+\eps)p)\\
			&=  t(1-(1+\eps)p)\\
\end{align*}

We then have\\
$$Pr[(1-\eps)k > rank(k) \lor rank(k) < (1=\eps)k] \leq Pr[L \geq tp] +  Pr[U \geq t(1-p)] $$

We now compute bounds on the probabilities. We note that $(1 + \frac{\eps}{1-\eps})t(1-\eps)p = tp$

\begin{align*}
	Pr[L \geq (1 + \frac{\eps}{1-\eps})t(1-\eps)p] &\leq e^{ \frac{-	(\frac{\eps}{1-\eps})^2t(1-\eps)p}{2 + \frac{\eps}{1-\eps}}}\\
\end{align*}
We then set this to $\delta$

\begin{align*}
	\delta &= e^{ \frac{-	(\frac{\eps}{1-\eps})^2t(1-\eps)p}{2 + \frac{\eps}{1-\eps}}}\\
	log(1/\delta) &= t\frac{(\frac{\eps}{1-\eps})^2(1-\eps)p}{2 + \frac{\eps}{1-\eps}}\\
	 \frac{log(1/\delta)(2 + \frac{\eps}{1-\eps})}{(\frac{\eps}{1-\eps})^2(1-\eps)p}&= t\\
	 \frac{log(1/\delta)(2 + \frac{\eps}{1-\eps})(1-\eps)}{\eps^2p}&= t\\
	 \frac{log(1/\delta)(2(1-\eps) + \eps)  }{\eps^2p}&= t\\
	 \frac{log(1/\delta)(2-\eps)  }{\eps^2p}&= t\\
\end{align*}

we then do the same for the other element,
We now compute bounds on the probabilities. We note that $(1 + \frac{\eps p}{1-(1+\eps)p})t(1 - (1+\eps)p) = t(1-p)$

\begin{align*}
	Pr[U \geq (1 + \frac{\eps p}{1-(1+\eps)p})t(1 - (1+\eps)p)] &\leq e^{ \frac{-	(\frac{\eps p}{1-(1+\eps)p})^2t(1-(1+\eps)p)}{2 + \frac{\eps p}{1-(1+\eps)p}}}\\
\end{align*}

We then set this to $\delta$

\begin{align*}
	\delta &= e^{ \frac{-	(\frac{\eps p}{1-(1+\eps)p})^2t(1-(1+\eps)p)}{2 + (\frac{\eps p}{1-(1+\eps)p})}}\\
	log(1/\delta) &=  \frac{	(\frac{\eps p}{1-(1+\eps)p})^2t(1-(1+\eps)p)}{2 + (\frac{\eps p}{1-(1+\eps)p})}\\
	\frac{log(1/\delta) (2 + \frac{\eps p}{1-(1+\eps)p})}{	(\frac{\eps p}{1-(1+\eps)p})^2(1-(1+\eps)p)} &= t  \\
	\frac{log(1/\delta) (2 + \frac{\eps p}{1-(1+\eps)p})(1-(1+\eps)p)}{	\eps^2 p^2} &= t\\
	\frac{log(1/\delta) (2 - 2p - \eps p)}{	\eps^2 p^2} &= t\\
\end{align*}

Now combine the two probabilities,

\begin{align*}
	t &= \frac{log(1/\delta) (2 - 2p - \eps p)}{	\eps^2 p^2} + \frac{log(1/\delta)(2-\eps)}{\eps^2p}\\
	 t &= log(1/\delta) (\frac{ (2 - 2p - \eps p)}{	\eps^2 p^2} + \frac{(2-\eps)}{\eps^2p})\\
	 t &=  \frac{ log(1/\delta)2(1 - \eps p ) }{	\eps^2 p^2}\\
	 t &=  \frac{ log(1/\delta)2(1 - \eps \frac{k}{n} )n^2 }{	\eps^2 k^2}\\
	 t &=  \frac{ log(1/\delta)2(n^2 - \eps nk ) }{	\eps^2 k^2}\\
\end{align*}

Therefore we need $ O(\frac{ log(1/\delta)2(n^2 - \eps nk ) }{	\eps^2 k^2})$ queries to distingush

\section{Problem 4}
\subsection{Part A}
Suppose that we we randomly create a set $Q \subseteq \{1,...,n\}$ where each element has $1/k$ chance of 
being included in the set. We first consider 
$Pr[Q \cap S = \emptyset]$. Because of how we constructed the set, for each element $s \in S$, 
$Pr[s \notin Q] = 1 - 1/k$, therefore $Pr[Q \cap S = \emptyset] = (1 - 1/k)^{ |S| }$. 
Let $$X_i = \ind{Q \cap S = \emptyset} \text{  and  } X = \sum_{i=1}^{t} X_i  \text{  and  } p = (1-1/k) = \frac{k-1}{k}$$

We note that,
\begin{align*}
	E[X] &= E[\sum_{i=1}^{t} X_i] \\
			&= \sum_{i=1}^{t} E[X_i]\\
			&= \sum_{i=1}^{t} (1 - 1/k)^{ |S| }\\
	E[X]	&= t (1 - 1/k)^{ |S| }\\
	E[X]	&= t p^{ |S| }\\
\end{align*}

Additionally,
\begin{align*}
	|S| \leq k &\implies E[X] \geq t p^{ k } \\
	|S| \geq (1+\eps)k &\implies E[X] \leq t p^{ (1+\eps)k } 
\end{align*}

When $X \leq (1-\eps/2)E[X]$ we predict that $|S| \geq (1+\eps)k$. Suppose that $|S| \leq k$ 
we know that the $\mu = E[X] \geq t p^{ k }$. Therefore, using the Chernoff bound,
$$Pr[X \leq (1-\eps)\mu] \leq e^{-\frac{\eps^2\mu}{4(2 + \eps})}$$

now if we set our probability of failure to be $\delta$,
\begin{align*}
	\delta &= e^{-\frac{\eps^2\mu}{4(2 + \eps)}}\\
	log(\delta) &= -\frac{\eps^2\mu}{2 + \eps}\\
	log(1/\delta) &= \frac{\eps^2\mu}{4(2 + \eps)}\\
	\frac{log(1/\delta)4(2 + \eps)}{\eps^2} &= \mu\\
	\frac{log(1/\delta)4(2 + \eps)}{\eps^2} &= tp^k\\
\end{align*}

note that for $1 \geq k, p^k = (\frac{k-1}{k})^k \leq 1$
\begin{align*}
	\frac{log(1/\delta)4(2 + \eps)}{\eps^2} &= t\\
	O(\frac{log(1/\delta)}{\eps^2}) &= t\\
\end{align*}

Now assume that $|S| \geq (1+\eps)k$, 
First note that 
$$ (1 + ((1-\eps/2)p^{-k\eps} - 1))p^{(1+\eps)k} =  (1-\eps/2)p^k $$

The probability of making an error is then,

$$
Pr[X \geq (1 + ((1-\eps/2)p^{-k\eps} - 1))\mu] \leq e^{\frac{((1-\eps/2)p^{-k\eps} - 1))^2 \mu}{2 + ((1-\eps/2)p^{-k\eps} - 1))}}
$$

We set $\mu = tp^{(1+\eps)k}$ since the closer the true $\mu$ is to the assumed $\mu = tp^k$ the more
likely we are to make an error,
setting this to $\delta$ we get

\begin{align*}
	\delta &= e^{\frac{-((1-\eps/2)p^{-k\eps} - 1))^2 \mu}{2 + ((1-\eps/2)p^{-k\eps} - 1))}} \\
	log(1/\delta) &= \frac{((1-\eps/2)p^{-k\eps} - 1))^2 \mu}{2 + ((1-\eps/2)p^{-k\eps} - 1))}\\
	log(1/\delta)(2 + ((1-\eps/2)p^{-k\eps} - 1))) &= ((1-\eps/2)p^{-k\eps} - 1))^2 \mu\\
	\frac{log(1/\delta)(2 + ((1-\eps/2)p^{-k\eps} - 1)))}{((1-\eps/2)p^{-k\eps} - 1))^2} &=  \mu\\
	\frac{log(1/\delta)(2 + ((1-\eps/2)p^{-k\eps} - 1)))}{((1-\eps/2)p^{-k\eps} - 1))^2} &=  tp^{(1+\eps)k}\\
	\frac{log(1/\delta)(2 + ((1-\eps/2)p^{-k\eps} - 1)))}{((1-\eps/2)p^{-k\eps} - 1))^2 p^{(1+\eps)k}} &=  t\\
	O(\frac{log(1/\delta)}{\eps^2}) &= t\\
\end{align*}

Therefore, in order to make the probability of error $1-\delta$ we only need $t = O(\frac{log(1/\delta)}{\eps^2})$
queries.

\subsection{Part B}

Let $F(k, k')$ be the estimator from part A, where the $F(k, k') = 0$ if the size of $S$ is predicted to 
be less than or equal to $k$ and  $F(k, k') = 1$ if the size of $S$ is predicted to be greater than 
or equal to $k'$ (eqivalent to $(1+\eps)k$). Our algorithm is as follows,

\begin{enumerate}
	\item $l \gets 1$
	\item $u \gets n$
	\item while $l > (1+\eps)u$
	\begin{enumerate}
		\item // The quartiles for the current range
		\item $q_1 \gets  l + (u - l) / 4$, $q_2 \gets  l + (u - l) / 2$, $q_3 \gets  l + 3(u - l) / 4$
		\item If $F(q_1, q_2) \neq F(q_2, q_3)$ then $l \gets q_1, u \gets q_3$ // Take the second and third quartile
		\item If $F(q_1, q_2) = F(q_2, q_3) = 1$ then $l \gets q_2, u \gets u$ // Take the bottom half
		\item If $F(q_1, q_2) = F(q_2, q_3) = 0$ then $l \gets l, u \gets q_2$ // Take the top half
	\end{enumerate}
	\item Return $l$
\end{enumerate}

\subsubsection{Probability}

Assuming $l \leq |S| \leq u$ we have 4 cases, one for each quartile.
If $|S|$ is in the first quartile, then with probability $(1-\delta)^2$ we have $F(q_1, q_2) = F(q_2, q_3) = 0$,
in which case $|S|$ stays in the range. Similar reasoning can be applied if $|S|$ is in the fourth quartile.
If $|S|$ is in the second quartile, then $F(q_2, q_3)$ is reliable but $F(q_1, q_2)$ is not. However, as long as 
$F(q_2, q_3)$ is correct, $|S|$ stays in the range. Symmetric reasoning is applied for when $|S|$ is in the 
third quartile. Therefore the probability of success for a given iteration is at least $(1-\delta)^2$.

In order to succeed we must have every iteration succeed, hence the probability that the algorithm succeeds is
at least $(1-\delta)^{2 \log_2 n}$.

\subsubsection{Runtime}

We note that the size of the interval we consider halves, therefore, for any $\eps$ we perform at most $O(\log n)$ iterations.
Each iteration takes a constant number of estimator evaluations therefore the number of $O(\log n)$ number of queries. We want our
probability of success to be at least $2/3$ hence we get,

\begin{align*}
	(1-\delta)^{2 \log n} &\geq 2/3\\
	\log (1-\delta) (2 \log n) &\geq \log (2/3)\\
	\log (1-\delta)  &\geq \frac{\log (2/3)}{(2 \log n)}\\
	1-\delta  &\geq (2/3)^{1/2 \log n}\\
	\delta  &\leq 1-(2/3)^{1/2 \log n}\\
	\log(1/\delta)  &\geq -\log(1-(2/3)^{1/2 \log n})\\
	\log(1/\delta)  &= O(\log \log x)
\end{align*}

Since the number of queries per estimator evaluation is $O(\log(1/\delta)/\eps^2)$  the number of queries
per iteration is $O(\log \log x)$. Therefore our runtime complexity is $O(\log n \log \log n)$.

\subsection{Part C}

We can randomly sample a unique element from $S$ as follows,

\begin{itemize}
	\item $Q =\{1,...,n\}$ 
	\item While $|Q| > 1$
	\begin{itemize}
			\item $Q' =$ random subset of $Q$ containing each element with probability $1/2$
			\item While $Q' \cap S = \emptyset$
			\begin{itemize}
				\item $Q' =$ random subset of $Q$ containing each element with probability $1/2$
			\end{itemize}
			\item $Q = Q'$
	\end{itemize}
	\item Return the single element in $Q$
\end{itemize}

\subsubsection{Runtime}

First we note that $Q$ always contains at least one element in $S$. Because the way $Q'$ is constructed,
$Pr[Q' \cap S = \emptyset] \leq 1/2$ at every step of the algorithm. Therefore we have that each iteration
in the worst case (where $S$ only has one element) we expect two queries to be executed. We then note that 
the size of $Q$ halves in expectation each iteration, therefore there are $O(\log_2 n)$ expected iterations
required. Since the expected number of queries per iteration is constant, the runtime of the algorithm 
is $O(\log_2 n)$

\subsubsection{Sampling Randomness}

Suppose our algorithm runs for $h$ steps before $|Q \cap S| = 1$  at which point the output
of the algorithm is determined and with the single element. We note that 
each element in each iteration has $1/2$ probability of being include in the next iteration, therefore 
the probability of each element being in $Q$ after $h$ iterations is $1/2^h$. Since all elements have equal probability
of being in the final set, the sample is uniformly random.

%\bibliographystyle{abbrvnat}
%\bibliography{references}
\end{document}

### 2022-04-06 hw/3/homework_3_turnin.tex
add homework and stuff

\documentclass[a4paper]{article}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{paralist}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{paralist}
\usepackage[svgname]{xcolor}
\usepackage{enumerate}
\usepackage{array}
\usepackage{times}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{environ}
\usepackage{times}
\usepackage{textcomp}
\usepackage{caption}
\usepackage[square,numbers]{natbib}

\urlstyle{rm}

\setlength\parindent{0pt} % Removes all indentation from paragraphs

% TO SHOW SOLUTIONS, include following (else comment out):
\newenvironment{soln}{
    \leavevmode\color{blue}\ignorespaces
}{}

\hypersetup{
%    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

\geometry{
  top=1in,            % <-- you want to adjust this
  inner=1in,
  outer=1in,
  bottom=1in,
  headheight=3em,       % <-- and this
  headsep=2em,          % <-- and this
  footskip=3em,
}

\title{Assignment 2} % Title
\author{
Derek Paulsen \\
}

\date{}

\newcommand{\ind}[1]{\mathds{1}[#1]}
\newcommand{\indeg}[1]{\delta_{in}(#1)}
\newcommand{\outdeg}[1]{\delta_{out}(#1)}
\def \eps{\epsilon}
\begin{document}
\maketitle

\section{Problem 1}
\subsection{Part 1}

We can update the flow by finding a path in the residual graph
from $s$ to $u$ and then finding a path from $v$ to 
$t$, when then push an extra unit of flow through this path if it 
exists. Find the path from $s$ to $u$ and from $v$ to 
$t$ can be done using breadth first search each requiring $O(m)$ time, meaning that
this is operation overall takes $O(m)$ time.

\subsection{Part 2}

Define a potential $p(u) = $ cheapest path from $s$ to $u$ before increasing the 
cost of $(u,v)$, because the initial flow is optimal, this potential will 
result in $\epsilon \geq 0$.
Additionally, since the costs are all integral, we know that all 
reduced costs are multiples of 1 (i.e. they are integral). Because all 
reduced costs are a multiple of the increment, in any edge with reduced cost 
$c_p(x,y) > 0$ included in a cycle will make the cycle's cost non-negative. 
Since the previous flow was optimal we only need to consider cycles that 
decrease the flow over the edge which has it's cost incremented. This means that
we only consider paths from $u$ to $v$ where all the edges have $c_p(x,y) = 0$.
Now run max flow from $u$ to $v$ using only edges 
with $c_p(x,y) = 0$.\\
\subsubsection*{Runtime}

$p$ can be computed with Bellman Ford in $O(mn)$ time, finding the edges with $c_p(u,v) = 0$ is $O(m)$, 
running max flow is $O(mn)$, therefore the whole procedure runs in $O(mn)$ time.

\subsubsection*{Algorithm}
Below is our algorithm, note that we omit the capacities for ease of exposition, however
the maxflow algorithm step runs with respect to the capacities of the original graph.
\begin{itemize}
	\item IncrementCost$(V, E, f, \hat{c}, (u,v), inc)$
	\item compute $p(w)$ = the cheapest path from $s$ to $w$
	\item $E' \gets \{(w,z) | (w,z) \in E, \hat{c}(w,z) + p(w) - p(z) = 0\}$
	\item $\hat{c}(u,v) \gets \hat{c}(u,v) + inc$ // increment the cost the edge
	\item Run maxflow from $u$ to $v$ over the graph $(V, E')$ with starting flow $f$ to get $f'$
	\item Return $f', \hat{c}$
\end{itemize}

\subsection{Part 3}

Let $c(u,v)$ be the true cost of an edge and $\hat{c}(u,v)$ be the current cost of the edge in 
the graph during algorithm execution.
Our algorithm works as follows. We begin by running maxflow over the input
graph and then setting the cost of every edge to be 0. We then iteratively increment the 
costs on the edges by decreasing powers of two mantaining the invariant that
the costs of all edges are always a multiple of the cost increment. Because the increment is 
always a multiple of the current costs of the edges we can 
use the algorithm from part 2 to correctly update the flow in $O(mn)$ time. The algorithm the terminates 
once our increment becomes $< 1$ at which point $\forall (u,v)\in E, c(u,v) = \hat{c}(u,v)$ since all of 
our costs are integral. Because at every step the flow  is optimal relative to $\hat{c}$, $c(u,v) = \hat{c}(u,v)$
implies that the cost is optimal to $c$, hence we terminate with 
a min cost maximum flow for the original input graph.

\begin{itemize}
	\item CostScaling$(V, E, c)$
	\item $\beta \gets \max_{(u,v) \in E}(\lfloor\log_2(c(u,v))\rfloor)$ // the largest power of two that is less than or equal to the maximum cost
	\item $\hat{c}(u,v) = 0$ // all edge costs are zero to start
	\item $f \gets $ maxflow from $s$ to $t$ over $(V, E)$
	\item While $\beta \geq 0$
	\begin{itemize}
		\item For each $(u,v) \in E$ where $\hat{c}(u,v) + 2^\beta \leq c(u,v)$ // each edge that can be incremented with the current increment
		\begin{itemize}
			\item $f, \hat{c} \gets $ IncrementCost$(V, E, f, \hat{c}, (u,v), 2^\beta)$
		\end{itemize}
		\item $\beta \gets \beta - 1$ //halve increment
	\end{itemize}
\end{itemize}

\subsubsection*{Runtime}

The outerloop runs $O(\log_2(C))$ and the inner loop in the worse case 
makes $m$ calls to the procedure from part 2. Hence we make $O(m\log C)$ calls
to the procedure from part 2. This results in an overall runtime of $O(m^2n\log C)$.

\section{Problem 2}

\subsection{Part 1}
We start with a high level overview of our algorithm.
We begin by running a max flow algorithm over the graph $G$. We then 
check if the flow is feasible (i.e. satisfies all of the lower bounds for each 
edge). If the flow is not feasible then we find cycles in the resdiual graph 
and push flow around the cycles in the satisfy the lower bounds of each edge. We call 
this initial flow feasible flow $f$.
We then update the capacity on each edge to be $f(e) - l(e)$ (the amount of slack that we 
have in that edge) and reverse all edges in the graph to get a new graph $G'$.
Then run maxflow from $t$ to $s$ on the updated graph to get a new maxflow $f'$. Finally substract this 
flow from the original flow to get $f - f' = f^*$ which is the final minflow.

First we define a $\gamma((u,v), f)$ to be the maximum amount of flow that can be passed through
an edge in the residual graph $H$, such that the resulting flow doesn't violate the constraints. Formally, this is 
$$
\gamma((u,v), f) = \begin{cases}
	u(e) - f(e) \text{  if  } (u,v) \in E \text{ (a forward edge in the graph)}\\
	f(e) - l(e) \text{  if  } (v,u) \in E \text{ (a backward edge in the graph)}
\end{cases}
$$

We then define $\Gamma(C, f)$ where $C$ is a cycle in the residual graph $H$ with flow $f$ to be 
$$
\Gamma(C, f) = \min_{(u,v) \in C}\{\gamma((u,v), f)\}
$$

\begin{itemize}
	\item Run any maxflow algorithm over $G$ with capacities = $u$ to get flow $f$
	\item While $f$ is not feasible (there are edges which violate the lower bounds)
	\begin{itemize}
		\item $e \gets $ an edge which has $f(e) < l(e)$
		\item find an cycle $C$ such that $e \in C, \Gamma(C, f) > 0$
		\item push $\Gamma(C, f)$ units of flow around $C$ and update $f$
	\end{itemize}
	\item Update the capacity of each edge $e \in E$ to be $f(e) - l(e)$ and reverse all of the edges in the graph, to get a new graph $G'$
	\item run max flow from $t$ to $s$ on $G'$ to obtain $f'$
	\item $f^* \gets f - f'$
	\item return $f^*$
\end{itemize}
\subsection{Part 2}

Let $P$ be the set of all $s-t$ paths and cycles which have no repeating edges.
Let $x_p$ be the flow running through path or cycle $p \in P$.
Let $$
A \in \mathds{Z}^{ |E| \times |P| } = A_{e,p} = \begin{cases} 1 \text{  if  } e \in p\\ 0 \text{  otherwise}\end{cases}
$$

Consider the following LP,
\begin{align*}
	\min_x \quad &\sum_{p\in P} x_p\\
	\text{  subject to} \quad\\
	Ax &\leq u\\
	Ax &\geq l\\
	x &\geq 0
\end{align*}

This is the min flow formulated as an LP. We can then rearrgage,

\begin{align*}
	\min_x \quad &\sum_{p\in P} x_p\\
	\text{  subject to} \quad\\
	-Ax &\geq -u\\
	Ax &\geq l\\
	x &\geq 0
\end{align*}

and take the dual,

\begin{align*}
	\max_{y, z} \quad & y^Tl - z^Tu\\
	\text{  subject to} \quad\\
	y^TA - y_i^TA &\leq 1\\
	y, y_i &\geq 0
\end{align*}

We now claim that $y, z$ define the forward and backward edges in the max $s-t$ cut
of the graph $G$.
\subsubsection*{Lemma : $y, z$ are integer}
We first note that the solution to the dual is integer because all entries of
$A$ are 0 or 1, hence for any basis $B$ of $A$, $det(B) \in \{-1, 0, 1\}$. 
By Cramer's rule, any basic feasible solution to the dual will have integer entries.

\subsubsection*{Lemma : $y, z$ are disjoint and 0,1 for some optimal solution}
Next we note that $y^TA - z^TA \leq 1 \implies y - z \leq 1$.
Taking a look at the objective of the dual we note that $u \geq l, \forall i = 1,...,|P|, h > 0, l_i(h + 1) - u_ih \leq  l_i$
this implies there exists an optimal solution where $y, z \leq 1$ and they are disjoint because
for any feasible solution where $\exists i, yi > 1, z_i > 0$, we could obtain a new feasible solution 
with objective value at least as good by setting $y_i = y_i - z_i$ and $z_i = 0$.

\subsubsection*{Meaning of the constriants}
A path $p \in P$. For any $s-t$ cut of $G$, a path must cross the cut 
an odd number of times. $y^TA - y_i^TA \leq 1$ simply means that the cut must cut the 
backward and forward edges. This is because if a path $p$ crosses a cut $(S, V\setminus S)$,  $2k+1$ times, then 
there must be $k+1$ forward edges and $k$ backward edges.

%Now consider a cycle in $P$, we know that
%the cycle must cross any cut and even number of times, hence 
\subsubsection*{Conclusion}

Finally, we observe that the objective of the dual is equal to the capacity of the maximum cut, by 
strong duality we have that if the primal has an optimal solution so does the dual and their respective
objective values are equal. Therefore we have that the min flow is equal to the max cut.

\subsection{Part 3}

We reduce the problem to a min flow problem by defining a graph $G$. We begin by defining our 
vertexes 
\begin{align*}
	V &= \{s, t\}\ \text{(source and sink)} \\ 
	&\cup \{u_i | i=1,...n\} \text{(start of class)}\\
	&\cup \{v_i | i=1,...n\} \text{(end of class)}\\
\end{align*}

Next we define the edges, 
\begin{align*}
	E &= \{(s,u_i) | i=1,...n\} \text{(source to start of each class, new student attends a class)}\\
	  &\cup \{(v_i, t) | i=1,...n\}  \text{(end of each class to the sink, student goes home)}\\
	  &\cup \{(u_i, v_i) | i=1,...n\} \text{(one edge per class, class attendance)}\\
	  &\cup \{(v_i, u_j) | i,j=1,...n, b_i + r_{i,j} \leq a_j\} \text{(an edge from the end of a class to another if there is time to get there)}
\end{align*}

We then define the lower bounds to be

$$
l : E \rightarrow \mathds{Z}^+ \quad
l(e) = \begin{cases}
	1 \text{  if  } e = (u_i, v_i)\\
	0 \text{  otherwise }
\end{cases}
$$
finally we define all upper bounds to be $n$,
$$
u : E \rightarrow \mathds{Z}^+ \quad
u(e) = n
$$

Each unit of flow to this problem corresponds to a single student, if that
unit of flow crosses edges $(u_i, v_i)$ then that student attends lecture $i$. 
The student can then either attend another lecture if time permits corresponding 
to following some edge in $\{(v_i, u_j) | i,j=1,...n, b_i + r_{i,j} \leq a_j\}$ or can 
go home (following the edge of the end of the lecture to the sink, $t$). Since the lower bound on 
all the edges corresponding to lectures have lower bound $1$ any valid solution to this min flow 
problem corresponds to a valid assignment of students to lectures. Finally, we minimize the flow 
which corresponds to the minimizing the number of students required to cover all lectures.

\section{Problem 3}

\subsection{Part 1}

Suppose we construct a bipartite graph $G = \{E, U, V\}$ where for each student $u \in U$ there 
is an edge $(u,v)$ if the student wishes to meet with faculty $v \in V$. By Konig's Theorm
we have that the maximum matching in this graph is equal in size to the minimum vertex cover. We
note that $|E| = d|V|$ and that each vertex in the graph covers at most $d$ edges, therefore the 
size of the minimum vertex cover is at least $|V|$ therefore the maximum matching is at least $|V|$,
hence there exists a perfect matching in the graph. A perfect matching corresponds to assigning
each student to meet with a single faculty member, this concludes the proof.

\subsection{Part 2}

Suppose we do a single round of matching where each student meets with a single faculty member. At the
end of this round we remove the edges that were used in the matching. This decreases the number of
edges by $|V|$ meaning that the degree of each student and faculty decreases by 1. However now we can apply the same proof as above, 
there are $(d-1)|V|$ edges in the graph and each vertex covers at most $d-1$ edges therefore the minimum
vertex cover is still of size $|V|$ hence there exists a perfect matching. After $d$ rounds 
of matching we have removed $d|V| = |E|$ edges meaning that all students have meet with all the faculty that 
they wanted to.

\subsection{Part 3}

Consider adding 'fake' edges such that each student has out degree $s$ and each professor has 
in degree $s$. From part 2 we know that we can complete all assignments in 
$s$ steps. At each step if a 'fake' edge is choosen then the time slot for the corresponding 
professor and student is free for that round. This produces a schedule where all desired meetings take
place is $s$ time slots, which completes the proof.

\section{Problem 4}
For this we first define our min cost flow problem.

Let $\indeg{v}$ be the number of edges entering vertex $v$ and $\outdeg{v}$ be the number of edges leaving vertex $v$.
Let $G = \{V, E\}$ be the input graph.
First construct two sets $X = \{v \in V | \indeg{v} > \outdeg{v}\}$ and 
$Y = \{v \in V | \indeg{v} < \outdeg{v}\}$.

Now we construct a graph, $G' = \{E', X, Y, s, t\}$ and define a cost function $c : E' \rightarrow \mathds{R}$ and a 
capacity function $u : E' \rightarrow \mathds{R}$.

We define $E' = X \times Y \cup \{(s,x) | x \in X\} \cup \{(y,t) | y \in Y\}$.
We then define our capacity function, 
$$
u(x,y) = 
\begin{cases} 
	\indeg{y} - \outdeg{y} \text{ if } x = s\\
	\outdeg{x} - \indeg{x} \text{ if } y = t\\
	\indeg{x} - \outdeg{x} \text{ else }\\
\end{cases}
$$

We then define our cost function,
$$
c(x,y) = 
\begin{cases} 
	0 \text{ if } x = s \lor y = t\\
	\text{cheapest path from $(x,y)$ in $G$ else}\\ 
\end{cases}
$$

We then run any min cost max flow algorithm over this graph. The resulting min cost flow $f^*$ is 
then used to modify $G$ by adding $f^*(x,y)$ 'fake' edges between $x,y$ in the the graph $G$. We then use 
a standard algorithm for finding the Eulerian path in the modified graph $G$ to give us a path
$P$. We then modify this path by replacing any 'fake' edge $(x,y)$ that we added in the last step with 
the shortest path between the $x$ and $y$ (which we computed during the construction of $G'$).

\subsubsection*{Optimality}

Suppose that for our input graph we must add at least $n$ edges to the graph to 
create a graph with an Eulerian path. Now suppose that we have a different modification to 
the graph which adds $k > 0, n+2k$ edges to the graph while still allow for an Eulerian graph.
By construction we there is some way to remove $2k$ edges to get a strictly better path, hence 
any optimal solution will only add the minimum number of edges required to make the graph Eulerian.

Now consider our solution, we note that we add the minimum number of edges to the graph to make it 
Eulerian. By construction we add these 'fake' edges in the cheapest possible way, hence 
our solution produces an optimal Eulerian path through the graph.

\end{document}

### 2022-04-06 hw/4/homework_4_turnin.tex
add homework and stuff

\documentclass[a4paper]{article}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{paralist}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{paralist}
\usepackage[svgname]{xcolor}
\usepackage{enumerate}
\usepackage{array}
\usepackage{times}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{environ}
\usepackage{times}
\usepackage{textcomp}
\usepackage{caption}
\usepackage[square,numbers]{natbib}

\urlstyle{rm}

\setlength\parindent{0pt} % Removes all indentation from paragraphs

% TO SHOW SOLUTIONS, include following (else comment out):
\newenvironment{soln}{
    \leavevmode\color{blue}\ignorespaces
}{}

\hypersetup{
%    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

\geometry{
  top=1in,            % <-- you want to adjust this
  inner=1in,
  outer=1in,
  bottom=1in,
  headheight=3em,       % <-- and this
  headsep=2em,          % <-- and this
  footskip=3em,
}

\title{Assignment 4} % Title
\author{
Derek Paulsen \\
}

\date{}

\newcommand{\ind}[1]{\mathds{1}[#1]}
\newcommand{\indeg}[1]{\delta_{in}(#1)}
\newcommand{\outdeg}[1]{\delta_{out}(#1)}
\def \eps{\epsilon}
\begin{document}
\maketitle

\section{Problem 1}
\subsection{Part A}
The first constraint has that all nodes
have a flow conservation constraint, i.e. each node must have net zero flow. 
the second constraint states that the total flow over all edges must sum to 1. 
Now consider a feasible solution to this LP. Since we must have positive flow 
on some of the edges and every node has a flow conservation constraint, any feasible 
solution must be a circulation to be feasible.

As the hint suggests. We can decompose any circulation into a set of cycles each with non-zero flow. 
Assume for contradiction that we have a \textit{unique} optimal solution $x^*$ which is composed of 
cycles $\{C_1, ..., C_k\}$, each contributing a total of $\{h_1, ..., h_k\}$ flow to the solution for some $k > 1$.

$$
h_j = \sum_{(u,v)\in C_j} x_{u,v}
$$

The cost of each of the cycles is,
$$
\sum_{(u,v)\in C_j} \frac{c(u,v)}{ |C_j| } \times h_j
$$

WLOG we assume that $C_1$ is the cycle with the minimum average cost. We then have 
$$
\sum_{(u,v)\in C_1} \frac{c(u,v)}{ |C_1| } \leq  \sum_{(u,v)\in C_j} \frac{c(u,v)}{ |C_j| }, \quad \forall j=2,...,k
$$

Since all flows are positive we then have, 
$$
\sum_{(u,v)\in C_1} \frac{c(u,v)}{ |C_1| } \times h_j \leq  \sum_{(u,v)\in C_j} \frac{c(u,v)}{ |C_j| } \times h_j , \quad \forall j=2,...,k
$$

This implies that we could obtain a solution with cost less than or equal to $x^*$, 
by pushing $\sum_{i=2}^k h_i / |C_1|$ flow around cycle $C_1$ and still have a 
feasible solution. This contradicts our assumption that $x^*$ is a unique optimal 
solution, therefore there exists an optimal solution which only contains a single cycle. 
The cost of this optimal solution is then 
$$
\sum_{(u,v)\in C_j} \frac{c(u,v)}{ |C_j| } \times 1
$$

Which is just the value of the minimum mean cost cycle. This completes the proof.

\subsection{Part B}
We first rewrite the LP in standard notation for flow LP's.

Let 
$$
A \in \mathds{R}^{|V| \times |E|} = \begin{cases} -1 \text{ if } (i,j) \in E\\ 1 \text{ if } (j,i) \in E\\ 0 \text{ otherwise}\end{cases}
$$

We can then write our LP as, 
\begin{align*}
	\min_x \quad &c^Tx\\
	\text{subject to} \quad Ax &= 0\\
		   \mathds{1}^Tx &= 1\\
						x &\geq 0\\
\end{align*}

We can then write down the dual,

\begin{align*}
	\max_{p, \epsilon} \quad &\epsilon\\
	\text{subject to} \quad p^TA + \mathds{1}^T\epsilon \leq c^T\\
\end{align*}

We now note that the dual defines a potential $p$, which maximizes 
the the minimum of $c(u,v) + p(u) - p(v)$

\begin{align*}
	p^TA + \mathds{1}^T \epsilon &\leq c^T\\
	\mathds{1}^T \epsilon &\leq c^T - p^TA\\
\end{align*}

Which implies,

$$
\epsilon = \min_{(u,v) \in E} c(u,v) + p(u) - p(v) = \min_{(u,v) \in E} c_p(u,v)
$$

Which is exactly the definition of $\epsilon$ from class,
this completes the proof.

\subsection{Part C}
From strong duality, we have that for primal and dual optimal soluions $x^*$ and $p^*, \epsilon^*$,

$$
c^Tx^* = \epsilon^*
$$

Since, the primal's optimal cost is the mimimum mean cost cycle, this means that 
$\epsilon$ is also equal to the mimimum mean cost cycle.

\section{Problem 2}

Consider the following LP
\begin{align*}
	\max_\pi \quad & \pi^T 0 \\
	\text{subject to} \quad \pi^TP &= \pi^T\\
		   \pi^T\mathds{1} &= 1\\
						\pi &\geq 0\\
\end{align*}

We can equivalently rewrite this as, 
\begin{align*}
	\max_\pi \quad & \pi^T 0 \\
	\text{subject to} \quad \pi^T(P - I) &= 0\\
		   \pi^T\mathds{1} &= 1\\
						\pi &\geq 0\\
\end{align*}

Now consider the dual,

\begin{align*}
	\min_{x, y} \quad &  0^Tx + y  \\
	\text{subject to} \quad (P - I)x + \mathds{1}y \geq  0\\
\end{align*}

Because the rows of $P$ sum to 1, and all entries are non-negative, we have that

$$
Px \leq \max_{i = 1,...,n} x_i
$$

This implies that for some element of $(P-I)x$ is $\leq 0$ because $(P-I)x = Px - x$. 
By the constraint of dual $(P - I)x + \mathds{1}y \geq  0$, this 
implies that $y \geq 0$. meaning that the problem is bounded. Additionally, 
there is a trivial feasible solution to the problem of $x = (0,...,0), y = 0$. 
Because the dual is bounded and feasible, this implies that it has a optimal solution, 
by strong duality we have that the primal must also have an optimal solution. Finally,
this implies that there exists $\pi$ that statifies that contraints of the primal, since 
$P$ was arbitrary we have that 
$$
\forall \text{ stochastic matrices } P, \exists \pi\quad  \pi^TP = \pi^T \land \sum_{i=1}^{n} \pi_i = 1
$$

Which is the statement we wanted to prove.

\section{Problem 3}

Consider the following LP,
\begin{align*}
	\min_x \quad  \max_{y,z} x^TAy + &x^TBz\\
	\text{subject to} \quad x,y,z &\in \Delta_p
\end{align*}

We then note that the objective is equalvalent to minimizing the sum 
of the largest element of $x^TA$ and $x^TB$, since the player 2 and 3 will 
just choose to play the strategy which takes the max of these elements.

We can introduce auxillary variables to rewrite,

\begin{align*}
	\min_{x, v_1, v_2} \quad  v_1 + v_2\\
	\text{subject to} \quad x^TA &\leq \mathds{1}^Tv_1\\
							x^TB &\leq \mathds{1}^Tv_2\\
							\sum_{i=1}^nx_i &=1 \\
							x &\geq 0\\
\end{align*}

Rearrange,

\begin{align*}
	\min_{x, v_1, v_2} \quad  v_1 + v_2\\
	\text{subject to} \quad x^TA - v_1 &\leq 0 \\
							x^TB - v_2 &\leq 0 \\
							\sum_{i=1}^nx_i &=1 \\
							x &\geq 0\\
\end{align*}

\begin{align*}
	\min_{x, v_1, v_2} \quad  v_1 + v_2\\
	\text{subject to} \quad -x^TA + v_1 &\geq 0 \\
							-x^TB + v_2 &\geq 0 \\
							\sum_{i=1}^nx_i &=1 \\
							x &\geq 0\\
\end{align*}

We then take the dual and get, 
\begin{align*}
	\max_{y,z,v_3} \quad  v_3\\
	\text{subject to} \quad -Ay - Bz + v_3 &\leq 0 \\
							\sum_{i=1}^ny_i &=1 \\
							\sum_{i=1}^nz_i &=1 \\
							y,z &\geq 0\\
\end{align*}

simplify

\begin{align*}
	\max_{y,z,v_3} \quad  v_3\\
	\text{subject to} \quad Ay + Bz  &\geq v_3 \\
							\sum_{i=1}^ny_i &=1 \\
							\sum_{i=1}^nz_i &=1 \\
							y,z &\geq 0\\
\end{align*}

We now note however that this is equivalent to, 
\begin{align*}
	\max_{y,z} \quad \min_{x} x^TAy + &x^TBz\\
	\text{subject to} \quad x,y,z &\in \Delta_p
\end{align*}

By strong duality we then have that the above dual has 
the same optimal value as the primal,

\begin{align*}
	\min_x \quad  \max_{y,z} x^TAy + &x^TBz\\
	\text{subject to} \quad x,y,z &\in \Delta_p
\end{align*}

Furthermore, both the dual and primal have trivial feasible solutions
which are any combination of the unit vectors in $\mathds{R}^n$. By strong duality 
since both problems are feasible they both have an optimal solution which is the same 
value, which is the definition of a Nash equilibrium. This completes the proof.

\section{Problem 4}
\subsection{Part A}
We note that the constraints of the LP relaxation are simply a statement of Hall's Theorem, 
however there is an equivalent formulation of this LP which has fewer constraints, which is as follows,

Let $y_{i,j,t}$ be the fraction of time that client $i$ stays are property $j$ during 
time slot $t$. 
Let $s_{i,j,t}$ be the preference for client $i$ for property $j$ during 
time slot $t$, which is 0 if client doesn't want to stay at the property or 1 if they do. 
We can then construct the LP as follows,

\begin{align}
	\min_{x} \sum_{j=1}^m x_j\\
	\text{subject to}\\
	\sum_{i=1}^n y_{i,j,t} &\leq x_j \quad \forall j\in [m], t \in [T]\\
	y_{i,j,t} &\leq s_{i,j,t} \quad \forall i\in [n], j\in [m], t \in [T]\\
	\sum_{j=1}^m y_{i,j,t} &= 1 \quad \forall i\in [n], t \in [T]\\
		0 \leq x &\leq 1\\
		0 \leq y &\leq 1
\end{align}

The above LP has a polynomial number of constraints, hence we can easily apply the 
ellipsoid method to get a polynomial runtime to find a solution.

\subsection{Part B}

Consider the set of $k$ clients that are not yet matched for some week $t$, from the constraints of 
the LP we have that,
$$\sum_{j\in \bigcup_{i \ in S} S_{i,t}} x_j \geq |S|$$

Let $N$ be the set of clients which are reachable via an alternating path
in the residual graph of the current matching starting from an unmatched client.
Again from the constraints we have that

$$\sum_{j\in \bigcup_{i \ in N} S_{i,t}} x_j \geq |N|$$

If we then remove all the locations currently used in the matching we get all 
of locations, which if they were added, would increase the size of the matching 
by 1. We now note that there are $|N| - k$ locations used in the current matching, each 
corresponding $x_j$ is at most 1 hence we get, 
$$\sum_{j\in \bigcup_{i \ in N} S_{i,t} \setminus \{\text{properties used in the matching}\}} x_j \geq |N| - |N| + k$$
$$\sum_{j\in \bigcup_{i \ in N} S_{i,t} \setminus \{\text{properties used in the matching}\}} x_j \geq k$$

By construction the probability that one of these properties is bought is,
$$ (\sum_{j\in \bigcup_{i \ in N} S_{i,t}  \setminus \{\text{properties used in the matching}\}} x_j) / OPT_{LP} \geq  k / OPT_{LP}$$
Therefore the probability that the matching increases by one is at least $k / OPT_{LP}$.

\subsection{Part C}

We begin with the probability that client $i$ is matched for a partiucular round,

$$Pr[\text{client $i$ is matched for week $t$ on round $r$}] \geq 1/OPT_{LP}$$

We then get, 
$$Pr[\text{client $i$ is not matched for week $t$ on round $r$}] \leq (1 - 1/OPT_{LP})$$

Which means that after $r$ rounds of adding properties we have,

$$Pr[\text{client $i$ is unmatched for week $t$ after $r$ rounds}] \leq (1 - 1/OPT_{LP})^r$$

We can then union bound to get the probability that the client is unmatched for any week

$$Pr[\text{client $i$ is unmatched for any week after $r$ rounds}] \leq T(1 - 1/OPT_{LP})^r$$

Finally, we union bound one more time to get probability that any client in unmatched,

$$Pr[\text{any client is unmatched for any week after $r$ rounds}] \leq nT(1 - 1/OPT_{LP})^r$$

We can then derive a value of $r$ that will give us a solution with high probability,

\begin{align*}
	nT(1 - 1/OPT_{LP})^r \leq nT e^{-r/OPT_{LP}}\\
\end{align*}

If we set $r = 4 OPT_{LP}\log n \log T$, we then compute the probability that 
any client is unmatched, for the first case, if $n > T$, 
\begin{align*}
	nT e^{-r/OPT_{LP}} &= nT \frac{1}{n^{4\log T}}\\
					&\leq n^2 \frac{1}{n^{4\log T}}\\
					&\leq  \frac{1}{n^{4\log T - 2}}\\
					&\leq  \frac{1}{n^{4\log T - 2}}\\
					&\leq  \frac{1}{n^2}\\
\end{align*}

If $T > n$,

\begin{align*}
	nT e^{-r/OPT_{LP}} &= nT \frac{1}{T^{4\log n}}\\
					&\leq T^2 \frac{1}{T^{4\log n}}\\
					&\leq  \frac{1}{T^2}\\
\end{align*}

We now note that, $r = 4OPT_{LP} \log n \log T \leq 4OPT\log n \log T$.

Hence we get our final algorithm which solves the LP relaxation in polynomial time, 
then random adds properties until we have perfect matchings for all weeks. We note that checking 
perfect matchings for each week is polynomial time using any maximum matching algorithm. Finally, we have shown that we 
terminate with high probability after $4OPT\log n \log T = O(OPT\log n \log T)$ rounds, hence 
our solution is $O(\log n  \log T)$ approximate and runs in polynomial time.

\end{document}

### 2022-04-06 proj/optimizer.py
add file

@@ -1,0 +1,47 @@
+from abc import ABC, abstractmethod, abstractproperty
+import pandas as pd
+
+
+class Optimizer(ABC):
+    
+    @abstractproperty
+    def results(self):
+        pass
+
+    @abstractmethod
+    def optimze(self, constraints):
+        pass
+    
+    @staticmethod
+    def create_results(constraints, weights):
+
+        satisfied = constraints.mul(weights).sub(1.0).le(0.0)
+
+
+
+    @staticmethod
+    def format_constraints(constraints):
+        # begins as (id2, matching tuple, id1)
+        constraints.index = pd.MultiIndex.from_tuples(list(map(eval, constraints.index)))
+        constraints.columns = pd.MultiIndex.from_tuples(list(map(eval, constraints.columns)))
+        return constraints
+    
+    @staticmethod
+    def _truncate_topk(grp, k):
+        if len(grp) <= k:
+            return grp
+
+        # take the topk based on the sum of the scores
+        take = grp.sum(axis=1)\
+                    .sort_values()\
+                    .tail(k)\
+                    .index
+
+        return grp.loc[take]
+
+
+    @staticmethod
+    def truncate_topk(constraints, k):
+        return constraints.grouby(level=0)\
+                .apply(lambda x : Optimizer._truncate_topk(x, k))
+    

### 2022-04-06 proj/print_stats.py
add data

@@ -1,0 +1,13 @@
+from pathlib import Path
+import pandas as pd
+from optimizer import Optimizer
+
+
+data = Path('./data')
+
+
+for f in data.glob('*.parquet'):
+    df = pd.read_parquet(f)
+    print(f.name)
+    s = df.groupby(level=0).count()
+    print(s)

### 2022-04-06 proj/optimizer.py
add stuff

@@ -1,4 +1,5 @@
 from abc import ABC, abstractmethod, abstractproperty
+import numpy as np
 import pandas as pd
 
 
@@ -13,11 +14,16 @@ class Optimizer(ABC):
         pass
     
     @staticmethod
-    def create_results(constraints, weights):
-
-        satisfied = constraints.mul(weights).sub(1.0).le(0.0)
-
-
+    def create_results(constraints, weights, method_name):
+        violated = constraints.mul(weights).sum(axis=1).sub(1.0).gt(0.0)
+        hist = violated.groupby(level=0).sum().value_counts()
+        
+        return {
+                'method_name' : method_name,
+                'total_violated' : violated.sum(),
+                'hist' : hist.to_dict(),
+                'max_violated' : np.max(hist.index.values)
+        }
 
     @staticmethod
     def format_constraints(constraints):

### 2022-04-06 proj/print_stats.py

@@ -7,7 +7,7 @@ data = Path('./data')
 
 
 for f in data.glob('*.parquet'):
-    df = pd.read_parquet(f)
+    df = Optimizer.format_constraints(pd.read_parquet(f))
     print(f.name)
-    s = df.groupby(level=0).count()
+    s = df.groupby(level=0).count().max(axis=1).describe()
     print(s)

### 2022-04-06 proj/run.py

@@ -1,0 +1,53 @@
+from argparse import ArgumentParser
+import json
+from pathlib import Path
+from datetime import datetime
+from torch_optimizer import TorchOptimizer
+from optimizer import Optimizer
+from pprint import pformat 
+
+DATA_DIR = Path('./data')
+datasets = list(DATA_DIR.glob('*.parquet'))
+
+argp = ArgumentParser()
+
+argp.add_argument('--method', required=True, choices=['torch', 'MILP', 'LP'])
+argp.add_argument('--k', required=False, default=50, type=int)
+argp.add_argument('--out_file', default='out.json')
+
+
+
+def get_optimizer(t):
+    if t == 'torch':
+        return TorchOptimizer()
+    else:
+        raise ValueError(t)
+
+def run(opt, f, k):
+    const = Optimizer.format_constraints(f)
+    const = Optimizer.truncate_topk(const, k)
+    w = opt.optimize(const)
+
+    res = opt.results.copy()
+    res['dataset'] = f.stem
+    res['dataset_path'] = str(f.absolute())
+    res['time_ran'] = str(datetime.now())
+    print(pformat(res))
+
+    return res
+
+
+def main(args):
+    if args.k < 1:
+        raise ValueError(args.k)
+
+    opt = get_optimizer(args.method)
+    with open(args.out, 'a') as ofs:
+        for ds in datasets:
+            res = run(opt, ds, args.k)
+            ofs.write(res.json.dumps(res)) 
+            ofs.write('\n')
+
+
+if __name__ == '__main__':
+    main(argp.parse_args())

### 2022-04-06 proj/torch_optimizer.py

@@ -1,20 +1,12 @@
-import pulp
-from utils import get_index_name, read_table, Timer, get_logger, init_spark, invoke_task
-from pyspark import SparkContext
-import operator
-import multiprocessing
-import pyspark.sql.functions as F
-from joblib import delayed
-
-
 import numpy as np
-from tqdm import tqdm
-from es.query_generator import QuerySpec
 import pandas as pd
-from copy import deepcopy
 import sys
 import torch
+from optimizer import Optimizer
 from torch import nn
+from utils import get_logger, Timer
+
+
 sys.path.append('.')
 pd.set_option('display.width', 150)
 
@@ -54,26 +46,35 @@ class BoostModel(nn.Module):
 
 
 
-class TorchOptimizer:
+class TorchOptimizer(Optimizer):
 
-    def __init__(self, rounds=50, round_duration=10, topk=50, restarts=10):
-        self._topk = topk
+    def __init__(self, iters=50, step_interval=10, restarts=10):
         self._restarts = restarts
-        self._rounds = rounds
+        self._iters = iters
         # number iterations per round (gradient updates)
-        self._round_duration = round_duration
-        self.results_ = {
-                'topk' : topk, 
-                'rounds' : rounds, 
-                'round_duration' : round_duration,
-                'total_grad_updates' : rounds * round_duration,
+        self._step_interval = step_interval
+        self._results = {
+                'iters' : iters, 
+                'step_interval' : step_interval,
+                'total_grad_updates' : iters * step_interval,
                 'restarts' : restarts
         }
     
-    
+    @property
+    def results(self):
+        return self._results
 
     def optimize(self, constraints):
-        pass
+        timer = Timer()
+        cands = [self._optimize(constraints) for i in range(self._restarts)]
+
+        self._results['opt_time'] = timer.get_interval()
+        # sort by the number of violated constraints
+        cands.sort(k=lambda x : x[1])
+        weights = cands[0][0]
+        self._results.update(Optimizer.create_results(constraints, weights, 'torch'))
+
+        return weights
 
 
     def _optimize(self, constraints):
@@ -85,24 +86,36 @@ class TorchOptimizer:
         
         X = torch.Tensor(constraints.values)
         label = -torch.ones(len(constraints), dtype=torch.float32)
+
+        niter = self._iters 
+
+        weights = [None] * niter
+        cvs = np.zeros(niter, dtype=np.int32)
+
+        for i in range(niter):
[…]

### 2022-04-06 proj/utils.py

@@ -1,0 +1,34 @@
+import sys
+import logging
+import time
+
+logging.basicConfig(
+        stream=sys.stderr,
+        format='[%(filename)s:%(lineno)s - %(funcName)s() ] %(asctime)-15s : %(message)s',
+)
+logger = logging.getLogger(__name__)
+
+class Timer:
+
+    def __init__(self):
+        self.start_time = time.time()
+        self.last_interval = time.time()
+
+    def get_interval(self):
+        t = time.time()
+        interval = t - self.last_interval
+        self.last_interval = t
+        return interval
+
+    def get_total(self):
+        return time.time() - self.start_time
+
+    def set_start_time(self):
+        self.start_time = time.time()
+
+
+def get_logger(name, level=logging.DEBUG):
+    logger = logging.getLogger(name)
+    logger.setLevel(level)
+
+    return logger

### 2022-04-06 proj/run.py
torch running

@@ -5,6 +5,10 @@ from datetime import datetime
 from torch_optimizer import TorchOptimizer
 from optimizer import Optimizer
 from pprint import pformat 
+from utils import get_logger
+
+
+log = get_logger(__name__)
 
 DATA_DIR = Path('./data')
 datasets = list(DATA_DIR.glob('*.parquet'))
@@ -13,7 +17,7 @@ argp = ArgumentParser()
 
 argp.add_argument('--method', required=True, choices=['torch', 'MILP', 'LP'])
 argp.add_argument('--k', required=False, default=50, type=int)
-argp.add_argument('--out_file', default='out.json')
+argp.add_argument('--out', default='out.json')
 
 
 
@@ -24,7 +28,8 @@ def get_optimizer(t):
         raise ValueError(t)
 
 def run(opt, f, k):
-    const = Optimizer.format_constraints(f)
+    log.info(f'running {f}')
+    const = Optimizer.read_constraints(f)
     const = Optimizer.truncate_topk(const, k)
     w = opt.optimize(const)
 
@@ -32,7 +37,7 @@ def run(opt, f, k):
     res['dataset'] = f.stem
     res['dataset_path'] = str(f.absolute())
     res['time_ran'] = str(datetime.now())
-    print(pformat(res))
+    log.info(f'\n{pformat(res)}')
 
     return res
 
@@ -45,7 +50,7 @@ def main(args):
     with open(args.out, 'a') as ofs:
         for ds in datasets:
             res = run(opt, ds, args.k)
-            ofs.write(res.json.dumps(res)) 
+            ofs.write(json.dumps(res)) 
             ofs.write('\n')
 
 

### 2022-04-06 proj/linprog_optimizer.py
add MILP

@@ -1,16 +1,8 @@
-from utils import  get_index_name, read_table, Timer, get_logger, init_spark, invoke_task
-from pyspark import SparkContext
-import operator
+from utils import  get_logger, Timer
 import multiprocessing
-import pyspark.sql.functions as F
-from joblib import delayed
+from optimizer import Optimizer
 
-
-import numpy as np
-from tqdm import tqdm
-from es.query_generator import QuerySpec
 import pandas as pd
-from copy import deepcopy
 import sys
 sys.path.append('.')
 pd.set_option('display.width', 150)
@@ -19,112 +11,37 @@ import pulp
 N_PROCS=28
 log = get_logger(__name__)
 
-class LinProgOptimizer:
-    pass
-    COST_TYPES = {
-            'max', 
-            'sum'
-    }
+class LinProgOptimizer(Optimizer):
 
-    def __init__(self, index, problem_cost='sum', query_cost='max', topk=25):
-
-        self._problem_cost = problem_cost
-        if self._problem_cost not in LinProgOptimizer.COST_TYPES:
-            raise ValueError(f'problem_cost must be one of the following {LinProgOptimizer.COST_TYPES}')
-
-        self._query_cost = query_cost
-        if self._query_cost not in LinProgOptimizer.COST_TYPES:
-            raise ValueError(f'query_cost must be one of the following {LinProgOptimizer.COST_TYPES}')
-
-        # the topk query results to take for model creation
-        self._topk = topk
+    def __init__(self, relaxation):
+        self._relaxed=relaxation
         # solver config
         self._solver_name = 'PULP_CBC_CMD'
         self._solver_threads = multiprocessing.cpu_count() # use all threads on the solver
-        self._solver_time_limit_secs = 600 # 10 mins
+        # no timelimit currently
+        #self._solver_time_limit_secs = 600 # 10 mins
         self._solver_presolve = True 
 
-        self.index = deepcopy(index)
-        self.index.to_spark()
-        init_spark()
+        self._results = {
+                'solver' : self._solver_name,
+                'relaxed' : self._relaxed,
+        }
 
 
-    def _gen_fvs_row(self, search_doc, res_ids, is_match, query_spec):
-
-        self.index.init()
-        clauses = self.index.query_gen.generate_query_clauses(search_doc, query_spec)
-        fvs = self.index.score_docs(res_ids.tolist(), clauses)
[…]

### 2022-04-06 proj/run.py

@@ -3,6 +3,7 @@ import json
 from pathlib import Path
 from datetime import datetime
 from torch_optimizer import TorchOptimizer
+from linprog_optimizer import LinProgOptimizer
 from optimizer import Optimizer
 from pprint import pformat 
 
@@ -20,6 +21,8 @@ argp.add_argument('--out_file', default='out.json')
 def get_optimizer(t):
     if t == 'torch':
         return TorchOptimizer()
+    elif t == 'MILP':
+        return LinProgOptimizer(False)
     else:
         raise ValueError(t)
 

### 2022-04-07 proj/linprog_optimizer.py
bug in optimizer

@@ -16,10 +16,10 @@ class LinProgOptimizer(Optimizer):
     def __init__(self, relaxation):
         self._relaxed=relaxation
         # solver config
-        self._solver_name = 'PULP_CBC_CMD'
+        self._solver_name = 'GUROBI_CMD'
         self._solver_threads = multiprocessing.cpu_count() # use all threads on the solver
         # no timelimit currently
-        #self._solver_time_limit_secs = 600 # 10 mins
+        self._solver_time_limit_secs = 3600 # 60 mins
         self._solver_presolve = True 
 
         self._results = {
@@ -27,21 +27,24 @@ class LinProgOptimizer(Optimizer):
                 'relaxed' : self._relaxed,
         }
 
+    
+    @property
+    def results(self):
+        return self._results
 
     def get_solver(self):
         return pulp.get_solver(
                 self._solver_name,
                 threads=self._solver_threads,
-                presolve=self._solver_presolve
+                timeLimit=self._solver_time_limit_secs,
         )
-                #timeLimit=self._solver_time_limit_secs,
     
     def _solve_problem(self, problem):
         timer = Timer()
         problem.solve(self.get_solver())
         return timer.get_interval()
 
-    def  _opt_mip(self, constraints):
+    def  _opt(self, constraints, relaxed):
         # the variables corresponding to the boost values in the 
         # query spec, this is the the main optimization target
         boost_vars = pd.Series(data=[
@@ -49,10 +52,16 @@ class LinProgOptimizer(Optimizer):
             ],
             index=constraints.columns
         )
+        
+        # switch between strict binary and relaxed LP
+        if relaxed:
+            kwargs = {'lowBound' : 0.0, 'upBound' : 1.0}
+        else:
+            kwargs = {'cat' : pulp.const.LpBinary}
 
         # binary vars used to indicate that the constraint has been violated
         constraint_indicator_vars = pd.Series(data=[
-            pulp.LpVariable(f'y{i}_{j}_{k}', cat=pulp.const.LpBinary) for i,j,k in constraints.index
+            pulp.LpVariable(f'y{i}_{j}_{k}', **kwargs) for i,j,k in constraints.index
             ],
             index=constraints.index
         )
@@ -102,15 +111,16 @@ class LinProgOptimizer(Optimizer):
 
 
     def optimize(self, constraints):
-        if self._relaxed:
-            raise NotImplementedError()
-        else:
[…]

### 2022-04-07 proj/optimizer.py

@@ -17,12 +17,13 @@ class Optimizer(ABC):
     def create_results(constraints, weights, method_name):
         violated = constraints.mul(weights).sum(axis=1).sub(1.0).gt(0.0)
         hist = violated.groupby(level=0).sum().value_counts()
-        
+        weights.index = [str(x) for x in weights.index]
         return {
                 'method_name' : method_name,
                 'total_violated' : int(violated.sum()),
                 'hist' : hist.to_dict(),
-                'max_violated' : int(np.max(hist.index.values))
+                'max_violated' : int(np.max(hist.index.values)),
+                'boost_weights' : weights.to_dict()
         }
 
     @staticmethod
@@ -43,12 +44,17 @@ class Optimizer(ABC):
                     .sort_values()\
                     .tail(k)\
                     .index
-
+        
         return grp.loc[take]
 
 
     @staticmethod
     def truncate_topk(constraints, k):
-        return constraints.groupby(level=0)\
+        c = constraints.groupby(level=0)\
                 .apply(lambda x : Optimizer._truncate_topk(x, k))
+
+        if len(c.index.names) == 4:
+            c.index = c.index.droplevel(0)
+        return c
+
     

### 2022-04-07 proj/print_stats.py

@@ -7,7 +7,6 @@ data = Path('./data')
 
 
 for f in data.glob('*.parquet'):
-    df = Optimizer.format_constraints(pd.read_parquet(f))
+    df = Optimizer.read_constraints(f)
     print(f.name)
-    s = df.groupby(level=0).count().max(axis=1).describe()
-    print(s)
+    print(df)

### 2022-04-07 proj/run.py

@@ -17,7 +17,7 @@ datasets = list(DATA_DIR.glob('*.parquet'))
 argp = ArgumentParser()
 
 argp.add_argument('--method', required=True, choices=['torch', 'MILP', 'LP'])
-argp.add_argument('--k', required=False, default=50, type=int)
+argp.add_argument('--k', required=False, default=25, type=int)
 argp.add_argument('--out', default='out.json')
 
 
@@ -25,6 +25,8 @@ argp.add_argument('--out', default='out.json')
 def get_optimizer(t):
     if t == 'torch':
         return TorchOptimizer()
+    elif t == 'LP':
+        return LinProgOptimizer(True)
     elif t == 'MILP':
         return LinProgOptimizer(False)
     else:

### 2022-04-07 proj/linprog_optimizer.py
update parameters, add more logging

@@ -113,10 +113,7 @@ class LinProgOptimizer(Optimizer):
     def optimize(self, constraints):
 
         boost_map = self._opt(constraints, self._relaxed)
-
-        boost_map = boost_map.dropna()
-        boost_map = boost_map[boost_map.gt(0)]
-        boost_map /= boost_map.min()
+        boost_map = self.post_process_boost_map(boost_map)
 
         method = 'LP' if self._relaxed else 'MILP'
         self._results.update(Optimizer.create_results(constraints, boost_map, method))

### 2022-04-07 proj/optimizer.py

@@ -57,4 +57,7 @@ class Optimizer(ABC):
             c.index = c.index.droplevel(0)
         return c
 
-    
+    @staticmethod
+    def post_process_boost_map(boost_map):
+        boost_map = boost_map.where(boost_map > 0).dropna()
+        return boost_map / boost_map.min()

### 2022-04-07 proj/torch_optimizer.py

@@ -48,7 +48,7 @@ class BoostModel(nn.Module):
 
 class TorchOptimizer(Optimizer):
 
-    def __init__(self, iters=500, step_interval=10, restarts=10):
+    def __init__(self, iters=500, step_interval=10, restarts=100):
         self._restarts = restarts
         self._iters = iters
         # number iterations per round (gradient updates)
@@ -71,7 +71,7 @@ class TorchOptimizer(Optimizer):
         self._results['opt_time'] = timer.get_interval()
         # sort by the number of violated constraints
         cands.sort(key=lambda x : x[1])
-        weights = cands[0][0]
+        weights = self.post_process_boost_map(cands[0][0])
         self._results.update(Optimizer.create_results(constraints, weights, 'torch'))
 
         return weights

### 2022-04-18 proj/linprog_optimizer.py
add more logging

@@ -1,6 +1,7 @@
 from utils import  get_logger, Timer
 import multiprocessing
 from optimizer import Optimizer
+from tempfile import mkstemp
 
 import pandas as pd
 import sys
@@ -10,6 +11,28 @@ import pulp
 
 N_PROCS=28
 log = get_logger(__name__)
+
+
+def parse_log_line(l):
+    toks = [x for x in l.split() if x[0].isdigit()]
+    if len(toks) != 9:
+        return None
+    # e.g. 120s
+    t = int(toks[-1][:-1])
+    # typically a float
+    soln_val = float(toks[5])
+
+    return (t, soln_val)
+
+
+def parse_gurobi_log(f):
+    with open(f) as ifs:
+        res = list(filter(lambda x : x is not None, map(parse_log_line, ifs)))
+        df = pd.DataFrame(res, columns=['time', 'obj_val'])
+
+    return df
+
+
 
 class LinProgOptimizer(Optimizer):
 
@@ -21,10 +44,12 @@ class LinProgOptimizer(Optimizer):
         # no timelimit currently
         self._solver_time_limit_secs = 3600 # 60 mins
         self._solver_presolve = True 
+        self._log_file = mkstemp(prefix='gurobi.log.')[1]
 
         self._results = {
                 'solver' : self._solver_name,
                 'relaxed' : self._relaxed,
+                'log_file' : self._log_file
         }
 
     
@@ -37,6 +62,7 @@ class LinProgOptimizer(Optimizer):
                 self._solver_name,
                 threads=self._solver_threads,
                 timeLimit=self._solver_time_limit_secs,
+                logPath=self._log_file
         )
     
     def _solve_problem(self, problem):
@@ -114,9 +140,11 @@ class LinProgOptimizer(Optimizer):
 
         boost_map = self._opt(constraints, self._relaxed)
         boost_map = self.post_process_boost_map(boost_map)
+        
+        time_series = parse_gurobi_log(self._log_file)
 
         method = 'LP' if self._relaxed else 'MILP'
-        self._results.update(Optimizer.create_results(constraints, boost_map, method))
+        self._results.update(Optimizer.create_results(constraints, boost_map, method, time_series))
 
         return boost_map
 

### 2022-04-18 proj/optimizer.py

@@ -14,7 +14,7 @@ class Optimizer(ABC):
         pass
     
     @staticmethod
-    def create_results(constraints, weights, method_name):
+    def create_results(constraints, weights, method_name , time_series):
         violated = constraints.mul(weights).sum(axis=1).sub(1.0).gt(0.0)
         hist = violated.groupby(level=0).sum().value_counts()
         weights.index = [str(x) for x in weights.index]
@@ -23,7 +23,8 @@ class Optimizer(ABC):
                 'total_violated' : int(violated.sum()),
                 'hist' : hist.to_dict(),
                 'max_violated' : int(np.max(hist.index.values)),
-                'boost_weights' : weights.to_dict()
+                'boost_weights' : weights.to_dict(),
+                'time_series' : time_series.to_json(index=False)
         }
 
     @staticmethod

### 2022-04-18 proj/run.py

@@ -53,7 +53,7 @@ def main(args):
 
     opt = get_optimizer(args.method)
     with open(args.out, 'a') as ofs:
-        for ds in datasets:
+        for ds in datasets[1:]:
             res = run(opt, ds, args.k)
             ofs.write(json.dumps(res)) 
             ofs.write('\n')

### 2022-04-18 proj/torch_optimizer.py

@@ -48,8 +48,8 @@ class BoostModel(nn.Module):
 
 class TorchOptimizer(Optimizer):
 
-    def __init__(self, iters=500, step_interval=10, restarts=100):
-        self._restarts = restarts
+    def __init__(self, iters=500, step_interval=10, timeout=600):
+        self._timeout = timeout
         self._iters = iters
         # number iterations per round (gradient updates)
         self._step_interval = step_interval
@@ -57,7 +57,7 @@ class TorchOptimizer(Optimizer):
                 'iters' : iters, 
                 'step_interval' : step_interval,
                 'total_grad_updates' : iters,
-                'restarts' : restarts
+                'timeout' : self._timeout
         }
     
     @property
@@ -66,26 +66,43 @@ class TorchOptimizer(Optimizer):
 
     def optimize(self, constraints):
         timer = Timer()
-        cands = [self._optimize(constraints) for i in range(self._restarts)]
-
+        X = torch.Tensor(constraints.values)
+        label = -torch.ones(len(constraints), dtype=torch.float32)
+        cands = []
+        while timer.get_total() < self._timeout:
+            w, obj_val = self._optimize(constraints.columns, X, label)
+            cands.append((w, obj_val, timer.get_total()))
+        
         self._results['opt_time'] = timer.get_interval()
-        # sort by the number of violated constraints
-        cands.sort(key=lambda x : x[1])
-        weights = self.post_process_boost_map(cands[0][0])
-        self._results.update(Optimizer.create_results(constraints, weights, 'torch'))
+        cands_df = pd.DataFrame(cands, columns=['weights', 'obj_val', 'time']).sort_values('time')
+
+        best_row = cands_df.sort_values(obj_val)\
+                        .iloc[0]
+    
+        boost_map = pd.Series(
+                np.maximum(best_row['weights'], 0.0),
+                index=constraints.columns
+        )
+        boost_map = boost_map[boost_map.gt(0)]
+        boost_map /= boost_map.min()
+        weights = self.post_process_boost_map(boost_map)
+
+        time_series = cands_df[['time', 'obj_val']]
+        # take the running best
+        time_series['obj_val'] = np.maximum.accumulate(time_series['obj_val'].values)
+
+        self._results.update(Optimizer.create_results(constraints, weights, 'torch', time_series))
 
         return weights
 
 
-    def _optimize(self, constraints):
[…]

### 2022-04-18 proj/linprog_optimizer.py
change timeouts

@@ -42,7 +42,8 @@ class LinProgOptimizer(Optimizer):
         self._solver_name = 'GUROBI_CMD'
         self._solver_threads = multiprocessing.cpu_count() # use all threads on the solver
         # no timelimit currently
-        self._solver_time_limit_secs = 3600 # 60 mins
+        self._solver_time_limit_secs = 100 
+        #self._solver_time_limit_secs = 3600 # 60 mins
         self._solver_presolve = True 
         self._log_file = mkstemp(prefix='gurobi.log.')[1]
 

### 2022-04-18 proj/torch_optimizer.py

@@ -48,7 +48,7 @@ class BoostModel(nn.Module):
 
 class TorchOptimizer(Optimizer):
 
-    def __init__(self, iters=500, step_interval=10, timeout=600):
+    def __init__(self, iters=500, step_interval=10, timeout=100):
         self._timeout = timeout
         self._iters = iters
         # number iterations per round (gradient updates)

### 2022-04-18 proj/optimizer.py
bug fixes

@@ -24,7 +24,7 @@ class Optimizer(ABC):
                 'hist' : hist.to_dict(),
                 'max_violated' : int(np.max(hist.index.values)),
                 'boost_weights' : weights.to_dict(),
-                'time_series' : time_series.to_json(index=False)
+                'time_series' : time_series.to_json(orient='table', index=False)
         }
 
     @staticmethod

### 2022-04-18 proj/run.py

@@ -53,7 +53,7 @@ def main(args):
 
     opt = get_optimizer(args.method)
     with open(args.out, 'a') as ofs:
-        for ds in datasets[1:]:
+        for ds in datasets:
             res = run(opt, ds, args.k)
             ofs.write(json.dumps(res)) 
             ofs.write('\n')

### 2022-04-18 proj/torch_optimizer.py

@@ -76,7 +76,7 @@ class TorchOptimizer(Optimizer):
         self._results['opt_time'] = timer.get_interval()
         cands_df = pd.DataFrame(cands, columns=['weights', 'obj_val', 'time']).sort_values('time')
 
-        best_row = cands_df.sort_values(obj_val)\
+        best_row = cands_df.sort_values('obj_val')\
                         .iloc[0]
     
         boost_map = pd.Series(
@@ -89,7 +89,7 @@ class TorchOptimizer(Optimizer):
 
         time_series = cands_df[['time', 'obj_val']]
         # take the running best
-        time_series['obj_val'] = np.maximum.accumulate(time_series['obj_val'].values)
+        time_series['obj_val'] = np.minimum.accumulate(time_series['obj_val'].values)
 
         self._results.update(Optimizer.create_results(constraints, weights, 'torch', time_series))
 

### 2022-04-18 proj/optimizer.py
change output format

@@ -24,7 +24,7 @@ class Optimizer(ABC):
                 'hist' : hist.to_dict(),
                 'max_violated' : int(np.max(hist.index.values)),
                 'boost_weights' : weights.to_dict(),
-                'time_series' : time_series.to_json(orient='table', index=False)
+                'time_series' : time_series.to_json(orient='records')
         }
 
     @staticmethod

### 2022-04-18 proj/optimizer.py
meh

@@ -24,7 +24,7 @@ class Optimizer(ABC):
                 'hist' : hist.to_dict(),
                 'max_violated' : int(np.max(hist.index.values)),
                 'boost_weights' : weights.to_dict(),
-                'time_series' : time_series.to_json(orient='records')
+                'time_series' : time_series.to_dict(orient='list')
         }
 
     @staticmethod

### 2022-04-18 proj/linprog_optimizer.py
change timeout

@@ -42,7 +42,7 @@ class LinProgOptimizer(Optimizer):
         self._solver_name = 'GUROBI_CMD'
         self._solver_threads = multiprocessing.cpu_count() # use all threads on the solver
         # no timelimit currently
-        self._solver_time_limit_secs = 100 
+        self._solver_time_limit_secs = Optimizer.TIMEOUT
         #self._solver_time_limit_secs = 3600 # 60 mins
         self._solver_presolve = True 
         self._log_file = mkstemp(prefix='gurobi.log.')[1]

### 2022-04-18 proj/optimizer.py

@@ -4,6 +4,8 @@ import pandas as pd
 
 
 class Optimizer(ABC):
+
+    TIMEOUT=2700 # 24hrs of cpu time
     
     @abstractproperty
     def results(self):

### 2022-04-18 proj/torch_optimizer.py

@@ -49,7 +49,7 @@ class BoostModel(nn.Module):
 class TorchOptimizer(Optimizer):
 
     def __init__(self, iters=500, step_interval=10, timeout=100):
-        self._timeout = timeout
+        self._timeout = Optimizer.TIMEOUT
         self._iters = iters
         # number iterations per round (gradient updates)
         self._step_interval = step_interval

### 2022-04-19 proj/linprog_optimizer.py
add experimental results

@@ -15,7 +15,7 @@ log = get_logger(__name__)
 
 def parse_log_line(l):
     toks = [x for x in l.split() if x[0].isdigit()]
-    if len(toks) != 9:
+    if len(toks) not in {9, 10}:
         return None
     # e.g. 120s
     t = int(toks[-1][:-1])

### 2022-04-19 hw/5/homework_5_turnin.tex
add new stuff

\documentclass[a4paper]{article}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{paralist}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{paralist}
\usepackage[svgname]{xcolor}
\usepackage{enumerate}
\usepackage{array}
\usepackage{times}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{environ}
\usepackage{times}
\usepackage{textcomp}
\usepackage{caption}
\usepackage[square,numbers]{natbib}

\urlstyle{rm}

\setlength\parindent{0pt} % Removes all indentation from paragraphs

% TO SHOW SOLUTIONS, include following (else comment out):
\newenvironment{soln}{
    \leavevmode\color{blue}\ignorespaces
}{}

\hypersetup{
%    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

\geometry{
  top=1in,            % <-- you want to adjust this
  inner=1in,
  outer=1in,
  bottom=1in,
  headheight=3em,       % <-- and this
  headsep=2em,          % <-- and this
  footskip=3em,
}

\title{Assignment 5} % Title
\author{
Derek Paulsen \\
}

\date{}

\newcommand{\ind}[1]{\mathds{1}[#1]}
\newcommand{\indeg}[1]{\delta_{in}(#1)}
\newcommand{\outdeg}[1]{\delta_{out}(#1)}
\def \eps{\epsilon}
\begin{document}
\maketitle

\section{Problem 1}
\subsection{Part A}

Let $S$ be the optimal solution to the knapsack instance, consider the value of this solution,

$$
value(S) = \sum_{i\in S} v_i
$$

If we round the values down we get a new value of 
$$
value(S') = \sum_{i\in S} \left\lfloor \frac{v_i}{V\epsilon/n} \right\rfloor V\epsilon/n
$$

by rounding, we decrease the value of each object by at most $V\epsilon/n$, hence we get,
$$
value(S') \geq  \sum_{i\in S}  v_i -(V\epsilon/n)
$$

We then show that 
\begin{align*}
	\sum_{i\in S}  v_i -(V\epsilon/n) &\geq  (1-\epsilon)\sum_{i \in S} v_i\\
	-|S|(V\epsilon/n) + \sum_{i\in S}  v_i  &\geq  (1-\epsilon)\sum_{i \in S} v_i\\
	-|S|(V\epsilon/n) &\geq  -\epsilon\sum_{i \in S} v_i\\
	|S|(V/n) &\leq  \sum_{i \in S} v_i\\
\end{align*}

We now note that $|S|/n \leq 1$, 
\begin{align*}
	V &\leq  \sum_{i \in S} v_i\\
\end{align*}

From the setup of the problem, we have that $V$ is the maximum value object which has weight at most $W$,
hence any optimal solution must have value at least $V$ since we could easily obtain a solution 
of value $V$ by taking said item. This means that indeed, $V \leq  \sum_{i \in S} v_i$ which implies, 
rounding the solution decreased the value by at most $(1-\epsilon)$.

\subsection{Part B}

Let $X = \{(v_i, w_i) | i\in [1,n]\}$ the set of possible items to take. We first 
round the value of each item as described above for some $\epsilon \in (0,1)$. 
Assume we obtain $v_1', ..., v_k'$ distinct values, where $\forall i<j, v_i' < v_j'$. Finally, 
let $X_i$ be the set of all items with value $v_i'$ around rounding. Our algorithm is then,

\begin{itemize}
	\item $A \gets $ an $k\times n$ matrix keeping track of the items used, the total value and weight 
	\item For $i \in [1,k]$
	\begin{itemize}
		\item $y \gets argmin_{x\in X_i}((weight(x))$ 
		\item $Y \gets X_i \setminus \{y\}$
		\item For $j \in [1,n]$
		\begin{itemize}
			\item $b \gets $ the maximum value set of items that we have already computed in $A$ which has total weight less than $W - weight(y)$ and at most $i-1$ items
			\item $c \gets $ the maximum value set of items that we have already computed in $A$ and at most $i$ items
			\item If $value(b) + value(y) = value(c)$ then $A_{ij} \gets $ the combination with the least weight
			\item If $value(b) + value(y) > value(c)$ then $A_{ij} \gets b \cup \{y\}$ and $y \gets argmin_{y\in Y}((weight(y))$ and $Y \gets Y \setminus \{y\}$
			\item If $value(b) + value(y) < value(c)$ then $A_{ij} \gets c$
		\end{itemize}
	\end{itemize}
\end{itemize}

The idea of this algorithm is that at we compute the best set of at most $i$ items only using 
items which have value at most $v_i$.

\section{Problem 2}
\subsection{Part A}
We solve this problem as follows. Suppose for $i\in [1,k]$ we have $c_i > 0$ items of size $s_i$. 
Let $x \in \mathds{Z}_+^k$ where $x_i = $ the number of items of size $s_i$ in the packing.
We then define $f(x)$ to be the optimal packing for items specified by $x$.

\begin{itemize}
	\item For $i \in [1,n]$
	\begin{itemize}
		\item For $x \in \{x | \sum_{j=1}^k x_j = i \land x_j \leq c_j\}$ // all combinations of $i$ items
		\begin{itemize}
			\item Compute $f(x)$ by taking at pair of previously compute optimal packings and merging them together
		\end{itemize}
	\end{itemize}
\end{itemize}

We note that we compute the optimal packing for all combinations of $x$. There are $(\Pi_{i=1}^k (s_i + 1)) - 1 = O(n^k)$ different 
combinations of sizes to compute. Computing each packing requires examining at most $O(n^k)$ different combinations and merging 
previous solutions takes $O(n)$ hence the runtime of the algorithm is $O(n^{2k+1})$ which is polynomial.

\subsection{Part B}

Suppose we have an empty bin which we pack by sorting the elements into decreasing order of size and then 
adding the elements until we cannot fit the next one. Additionally, assume that the first element is of
size $l$. We note that the amount of unused space in the bin is less than or equal to $l$ since all of the elements in 
the stream we were trying to pack were of size less than or equal to $l$. Now suppose we have 
packed all the items of size $ > \epsilon$ into $B$ bins, leaving only items of size $ \leq \epsilon$ left. We then proceed to
three steps. First, we sort the items into order of decreasing size. Next,
for each of the existing $B$ bins, we greedily items from the sorted list until the next item doesn't fit,
at which point we move to the next bin and repeat the process. Finally, we pack the remaining items into 
bins in the same fasion until we have exhausted all of the bins. We now consider how many bins we have used by the end of the procedure.

Let $A = \sum_{i=1}^n a_i$.
After the second step of the algorithm above, we have two cases. In the first case, we have exhausted all 
of the items we needed to pack in which case we have packed all of the items into $B$ bins. In the 
second case, we havn't exhausted the items, we now consider this case.
From the lemma above, after out second step each of the $B$ bins has at most $\epsilon$ wasted space, 
hence each bin has at least $(1-\epsilon)$ stuff packed into it. This means that we have at most 
$A - \frac{B}{1-\epsilon}$ stuff left to pack into bins, again apply the lemma above we know that 
we can pack all of these items into at most,

$$
\left\lceil\frac{A - \frac{B}{1-\epsilon}}{1-\epsilon}\right\rceil
$$

bins. Additionally by our assumption we have that $A - \frac{B}{1-\epsilon} > 0$, we now want to show,
\begin{align*}
	\left\lceil\frac{A - \frac{B}{1-\epsilon}}{1-\epsilon}\right\rceil &\leq 1+(1+2\epsilon)B^*\\
\end{align*}

Remove the ceil, 
\begin{align*}
	B+1+\frac{A - \frac{B}{1-\epsilon}}{1-\epsilon} &\leq 1+(1+2\epsilon)B^*\\
	B+\frac{A - \frac{B}{1-\epsilon}}{1-\epsilon} &\leq (1+2\epsilon)B^*\\
	 B(1-\epsilon)+A - \frac{B}{1-\epsilon} &\leq (1-\epsilon)(1+2\epsilon)B^*\\
	 A - B\frac{\epsilon(2-\epsilon)}{1-\epsilon} &\leq (1-\epsilon)(1+2\epsilon)B^*\\
\end{align*}

$A \leq B^*$,

\begin{align*}
	B^* - B\frac{\epsilon(2-\epsilon)}{1-\epsilon} &\leq (1-\epsilon)(1+2\epsilon)B^*\\
	 B^* - B\frac{\epsilon(2-\epsilon)}{1-\epsilon} &\leq  (1+ \epsilon -2\epsilon^2)B^*\\
	  -    B\frac{\epsilon(2-\epsilon)}{1-\epsilon} &\leq  (\epsilon -2\epsilon^2)B^*\\
	  -    B\frac{\epsilon(2-\epsilon)}{1-\epsilon} &\leq  \epsilon(1 -2\epsilon)B^*\\
	  -    B\frac{(2-\epsilon)}{1-\epsilon} &\leq   (1 -2\epsilon)B^*\\
	  -    B\frac{(2-\epsilon)}{1-\epsilon} &\leq   (1 -2\epsilon)B^*\\
	  -    B(2-\epsilon) &\leq   (1-\epsilon)(1 -2\epsilon)B^*\\
\end{align*}

We now note that the right hand side of the equation is strictly positive while the 
left hand side is strictly negative, hence the inequality holds. Hence in both cases we 
use $\max\{B, 1+(1+2\epsilon)B^*\}$ bins to pack all of the items.

Finally, the sorting of the items takes $O(n\log n)$ and the inserting them takes $O(n)$ hence 
the procedure takes $O(n \log n)$ time which is polynomial in $n$.

\subsection{Part C}

Consider the round instances, $\forall i<j, x\in S_i, y \in S_j, original\_size(x) \geq rounded\_size(y)$. Now consider 
the original packing, we could set aside all of the items in $S_1$ and then place each item from 
$S_2$ in the spot used for $S_1$. We could then repeat this process for each set, place the items
from the set in spaces used by the previous set in the original packing and still end up with a valid packing for the 
items in $S_2, ...S_{n/k}$. We are then left with $n/k$ items in $S_1$ which trivially fit into $n/k$ bins (one bin 
per item) hence we used the original packing + at most $n/k$ extra bins for the rounded instance.

\subsection{Part D}
Let $S$ be the set of items of size greater than $\epsilon/2$. 
Let $B$ be the optimal packing for $S$. From the question above we have that 
after rounding and packing the large items we have used at most $B + 1/2|S|\epsilon^2$ for $k = 1/2\epsilon^2$. From part two we get that 
we can pack the remaining items into $\max\{B + 1/2|S|\epsilon^2, 1+(1+\epsilon)B^*\}$ bins. We now show that,

\begin{align*}
	B + 1/2|S|\epsilon^2 \leq 1+(1+\epsilon)B^*\\
\end{align*}

$B \leq B^*$,
\begin{align*}
	B^* + 1/2|S|\epsilon^2 \leq 1+(1+\epsilon)B^*\\
	1/2|S|\epsilon^2 \leq 1 + \epsilon B^*\\
	1/2|S|\epsilon \leq B^*\\
\end{align*}

$\sum a_i \leq B^*$,
\begin{align*}
	1/2|S|\epsilon \leq \sum a_i\\
	1/2|S|\epsilon \leq \sum_{i\in S} a_i\\
\end{align*}

We now note that this must be true since all items in $S$ are at least size 
$\epsilon/2$. Hence $B + 1/2|S|\epsilon^2 \leq 1+(1+\epsilon)B^*$. Which implies that 
we can pack the remaining items and end up with, $1+(1+\epsilon)B^*$ bins.

Finally, from part one we show that we can pack the items the large items into bins in 
$O(n^{2k+1}$ time and in $O(n)$ time for the small bins, hence the algorithm runs in polynomial time.

\section{Problem 3}

\subsection{Part A}

Consider a sequence of prices where the price is $B$ until time $T$ at which point all subsequent prices becomes $0$.
Clearly after time $T$ the algorithm should buy the computer (no one is going to pay you to take a new computer right?).
This of course is the exact same problem as the ski rental problem. You pay one dollar per minute (day in the ski rental) 
and at any point you can purchase the computer for $B$ dollars (equivalent to buying the skis) and stop. After time $T$, 
any algorithm will stop which is equivalent to quitting skiing.

\subsection{Part B}

My calculus is very bad, however the idea behind that if you use the skiing problem and have set the probability of
the price dropping to zero (stopping skiing) to an exponential distribution, any deterministic stopping 
rule will have expected competitive ratio of $e/(e-1)$. Since no deterministic algorithm can do better than 
$e/(e-1)$ no randomized algorithm can do better than $e/(e-1)$, hence no algorithm can do better than $e/(e-1)$.

\subsection{Part C}

Here is a cow.
\begin{verbatim}
 ___________________________________
/ I've forgotten all the calculus i \
\ learned in my undergrad           /
 -----------------------------------
        \   ^__^
         \  (xx)\_______
            (__)\       )\/\
             U  ||----w |
                ||     ||
\end{verbatim}

\section{Problem 4}
\subsection{Part A}

As in lecture we wish to show that for each step,

\begin{equation}
\Delta ALG(\sigma) + \Delta\phi \leq k \Delta OPT(\sigma)
\end{equation}

again we break each request into two stages, in stage one OPT moves, 
and stage two DC moves. Now consider the stage 1, where OPT moves distance $d$ to service the 
request, consider the changes of each

\begin{itemize}
		\item $\Delta OPT = d$, since OPT moved $d$
		\item $\Delta DC = 0$, since we didn't move yet
		\item $\Delta\phi_1 \leq kd$, since in the worse case, the cost of $M(x,y)$ increased by $d$
		\item $\Delta\phi_2 = 0$, since we didn't move yet
\end{itemize}

summing these terms together we get 
$$
0 + kd  + 0 \leq kd
$$

hence for stage one the inequality holds.

We now consider stage two. Let $d_i$ be the distance 
that server $i$ moves. WLOG we assume that $\{d_1, ..., d_k\}$ is 
sorted in increasing order, that is, server $k$ moves the furthest and 
actually goes to the request, server $1$ moves the least distance. For ease of exposition,
we break this movement into $k$ substeps, at step $i$, all servers $>= i$ move $(d_i - d_{i-1})$,
where $d_0 = 0$. Now consider the pairwise distances. For when a single server moves it 
moves $(d_i - d_{i-1})$ units closer to the servers it doesn't cover, while moving away from 
the servers that it covers. We note that each server is only every covered by a single server
that is moving hence at each substep, the pairwise distances increase by at most $i(d_i - d_{i-1})$,
hence we get,
\begin{align*}
	(d_i - d_{i-1}) &= \text{ the distance covered by the servers}\\
	(k-i-1) &= \text{ the number of servers that are moving}\\
	(k-1) &= \text{ the number of servers that the moving server is moving towards}\\
	-2i &= \text{ compensation for the servers that are currently covered, times 2 because the previous term counts the distance as decreasing}
\end{align*}

We know consider 
each component, 
\begin{itemize}
		\item $\Delta OPT = 0$, since OPT didn't move
		\item $\Delta DC = \sum_{i=1}^k d_i$, the sum of the distances that we moved,
		\item $\Delta\phi_1 \leq k(\sum_{i=1}^{k-1} d_i - d_k)$, since in the worse case, all servers except for server $k$ moved away from its matched counterpart
		\item $\Delta\phi_2 = \sum_{i=1}^k((k-i+1)(k-1) - 2i)(d_i - d_{i-1})$
\end{itemize}

we then wish to show,
\begin{align*}
	\sum_{i=1}^k d_i + k(\sum_{i=1}^{k-1} d_i - d_k) - \sum_{i=1}^k((k-i+1)(k-1) - 2(i-1))(d_i - d_{i-1}) \leq 0k\\
		\sum_{i=1}^k d_i + k(\sum_{i=1}^{k-1} d_i - d_k) - \sum_{i=1}^k(k^2-ki-i-3)(d_i - d_{i-1}) \leq 0\\
		\sum_{i=1}^k d_i + k(\sum_{i=1}^{k-1} d_i - d_k) - (k^2-3)d_k + \sum_{i=1}^{k}(-ki-i)(d_i - d_{i-1}) \leq 0\\
		\sum_{i=1}^k d_i + k(\sum_{i=1}^{k-1} d_i - d_k) - (k^2-3)d_k - (k+1)\sum_{i=1}^{k}d_i  \leq 0\\
		k(\sum_{i=1}^{k-1} d_i - d_k) - (k^2-3)d_k - k\sum_{i=1}^{k}d_i  \leq 0\\
		-2kd_k - (k^2-3)d_k  \leq 0\\
		True
\end{align*}

Hence for stage two the inequality hold for stage two. This implies that

\begin{equation}
\Delta ALG(\sigma) + \Delta\phi \leq k \Delta OPT(\sigma)
\end{equation}

For all steps, which implies that the algorithm is k-competitive.

\subsection{Part B}

Consider a star graph, where each edge corresponds to a single page and all edges are length 1. 
Suppose we start with all the servers in the center point of the graph and all requests
are at the end points of the edges (away from the center). Suppose we apply algorithm from above, when a request comes in
at the end of one of the edges of the graph,
all servers technically cover each other, hence only a single server is sent to service the request. 
This page is then 'marked' because the server won't leave move from this point until $k$ unique requests have 
come in.  For subsequent requests there are three scenarios.

\begin{enumerate}
		\item The request is where a server currently, in which case no server moves, a cache hit.
		\item The request is not where a server currently is, and there is at least one server in the middle of the graph, at which point a random server will
			choosen to service the request, marking the page
		\item The request is not where a server currently is, and there arn't any servers in the middle of the graph, at which point all servers will move
			to the middle of the graph 'unmarking' all of the pages (and evicting them) and then moving a random server to service the request, marking the new page.
\end{enumerate}

We can see that this algorithm evicts pages unnecessarily, hence any Marking algorithm is at least as good as this algorithm. Since this 
algorithm is k-competitive, any marking algorithm must also be k-competitive.

\end{document}

### 2022-04-19 hw/scribe/lecture_20.tex
add new stuff

\documentclass[11pt]{article}

\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{fullpage}
\usepackage{epsfig}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{theorem}
\usepackage{tabu,multirow}
\usepackage{graphicx}
\usepackage{color} 
\makeatletter
\setlength{\textwidth}{6.5in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{0.25in}
\setlength{\textheight}{8.25in}
\setlength{\headheight}{0pt}
\setlength{\headsep}{0pt}
\setlength{\marginparwidth}{59pt}

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt plus 1pt}
\setlength{\theorempreskipamount}{5pt plus 1pt}
\setlength{\theorempostskipamount}{0pt}
\setlength{\abovedisplayskip}{8pt plus 3pt minus 6pt}

\renewcommand{\section}{\@startsection{section}{1}{0mm}%
                                 {2ex plus -1ex minus -.2ex}%
                                 {1.3ex plus .2ex}%
                                 {\normalfont\Large\bfseries}}%
\renewcommand{\subsection}{\@startsection{subsection}{2}{0mm}%
                                   {1ex plus -1ex minus -.2ex}%
                                   {1ex plus .2ex}%
                                   {\normalfont\large\bfseries}}%
\renewcommand{\subsubsection}{\@startsection{subsubsection}{3}{0mm}%
                                   {1ex plus -1ex minus -.2ex}%
                                   {1ex plus .2ex}%
                                   {\normalfont\normalsize\bfseries}}
\renewcommand\paragraph{\@startsection{paragraph}{4}{0mm}%
                                  {1ex \@plus1ex \@minus.2ex}%
                                  {-1em}%
                                  {\normalfont\normalsize\bfseries}}
\renewcommand\subparagraph{\@startsection{subparagraph}{5}{\parindent}%
                                     {2.0ex \@plus1ex \@minus .2ex}%
                                     {-1em}%
                                    {\normalfont\normalsize\bfseries}}
\makeatother

\newenvironment{proof}{{\bf Proof:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofof}[1]{{\bf Proof of #1:  }}{\hfill\rule{2mm}{2mm}}
\newenvironment{proofofnobox}[1]{{\bf#1:  }}{}
\newenvironment{example}{{\bf Example:  }}{\hfill\rule{2mm}{2mm}}
\renewcommand{\thesection}{\lecnum.\arabic{section}}

\renewcommand{\theequation}{\thesection.\arabic{equation}}
\renewcommand{\thefigure}{\thesection.\arabic{figure}}

\newtheorem{fact}{Fact}[section]
\newtheorem{lemma}[fact]{Lemma}
\newtheorem{theorem}[fact]{Theorem}
\newtheorem{definition}[fact]{Definition}
\newtheorem{corollary}[fact]{Corollary}
\newtheorem{proposition}[fact]{Proposition}
\newtheorem{claim}[fact]{Claim}
\newtheorem{exercise}[fact]{Exercise}

% math notation
\newcommand{\R}{\ensuremath{\mathbb R}}
\newcommand{\Z}{\ensuremath{\mathbb Z}}
\newcommand{\N}{\ensuremath{\mathbb N}}
\newcommand{\F}{\ensuremath{\mathcal F}}
\newcommand{\SymGrp}{\ensuremath{\mathfrak S}}
\newcommand{\pr}{{\bf {\rm Pr}}}

\newcommand{\CCP}{\ensuremath{\sf P}}
\newcommand{\CCNP}{\ensuremath{\sf NP}}
\newcommand{\C}{\ensuremath{\mathcal C}}

\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\ceil}[1]{\ensuremath{\left\lceil#1\right\rceil}}
\newcommand{\floor}[1]{\ensuremath{\left\lfloor#1\right\rfloor}}
\newcommand{\poly}{\operatorname{poly}}
\newcommand{\polylog}{\operatorname{polylog}}

% some abbreviations
\newcommand{\e}{\varepsilon}
\newcommand{\half}{\ensuremath{\frac{1}{2}}}
\newcommand{\junk}[1]{}
\newcommand{\sse}{\subseteq}
\newcommand{\union}{\cup}
\newcommand{\meet}{\wedge}

\newcommand{\prob}[1]{\ensuremath{\text{{\bf Pr}$\left[#1\right]$}}}
\newcommand{\expct}[1]{\ensuremath{\text{{\bf E}$\left[#1\right]$}}}
\newcommand{\Event}{{\mathcal E}}

\newcommand{\mnote}[1]{\normalmarginpar \marginpar{\tiny #1}}
\newcommand{\DC}{\text{DC}}
\newcommand{\OPT}{\text{OPT}}

\newcommand{\headings}[4]{
\noindent
\fbox{\parbox{\textwidth}{
	{\bf CS787: Advanced Algorithms} \hfill {{\bf Scribe:} #3}\\
\vspace{0.02in}
{{\bf Lecture #1:} #2} \hfill {{\bf Date:} #4}
}}
\vspace{0.2in}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Document begins here %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% Fill in lecture number, topic, your name, and date of lecture below. %%

\newcommand{\lecnum}{20}
\headings{\lecnum}{k-Server Problem Analysis; Convex Body Chasing}{Derek Paulsen}{4/7/22}
\section{k-server problem, continued}
Last time, we discussed the k-server problem,
\begin{itemize}
	\item You have k-servers that are in a metric space (e.g. a line, 2-d euclidean space)
	\item Requests arrive online each at some point in the space
	\item Each request must be addressed by sending a single server to the request
	\item Minimize the total distance the servers move to service the requests
\end{itemize}

The naive greedy algorithm of sending the closest server to the request to this problem is not competitive. 
In this lecture we discuss the line metric, analyze the double coverage algorithm, and show that it is k-competitive.
Recall the double coverage algorithm,

\subsection{Double Coverage Algorithm}
\begin{algorithm}
	\caption{Double Coverage Algorithm}
\begin{algorithmic}[1]
	\State $x \gets $ the current positions of the servers
	\State $r \gets $\text{ the next request}
	\If{$r \not\in [\min(x), \max(x)]$}
		\State move the closest server to $r$
	\Else
		\State $i \gets $ the closest server to the right of $r$
		\State $j \gets $ the closest server to the left of $r$
		\State $d \gets \min\{ |x_i - r|, |x_j - r| \}$
		\State move servers $x_i$ and $x_j$ $d$ units toward $r$
	\EndIf
\end{algorithmic}
\end{algorithm}

\begin{theorem}{The Double Coverage Algorithm is k-competitive} 
\end{theorem}

Let $\sigma$ be a sequence of requests, $\DC(\sigma)$ be the cost the double coverage algorithm 
pays to service the requests and $\OPT(\sigma)$ be the cost that $\OPT$ pays to service the requests.

\textbf{Proof :} For any sequence of requests $\sigma$ we want to show,
\begin{equation}
\DC(\sigma) \leq k \OPT(\sigma) + C
\end{equation}

Ideally, we would show that for each request, $\DC$ pays at most $k\OPT$ to service the request. 
However this is unfortunately not true (imagine if $\OPT$ already has a server on the request).
Instead we use a potential argument and show that for each request $i$, the change in $\DC = \Delta\DC$ plus
the change in the potential, $\phi^{i} - \phi^{i-1} = \Delta\phi^i$ is less than $k$ times the change in $\OPT = k\Delta\OPT$, 
that is,

\begin{equation}
\Delta\DC^i(\sigma) + \Delta\phi^{i} \leq k \Delta\OPT^i(\sigma)
\label{eq:step}
\end{equation}

Where $\phi$ is the potential function. Specifically, for request $i$ if $x^i$ is the position of 
our servers and $y^i$ is the position of OPT's servers

$$
\phi^i = \phi(x^i, y^i) = \phi_1^i +  \phi_2^i = kM(x^i, y^i) + \sum_{h<l} d(x_h^i, x_l^i)
$$

Where $\phi_1 = kM(x,y)$ is $k$ times the minimum distance matching between $x$ and $y$, and $ \phi_2 = \sum_{h<l} d(x_h, x_l)$ 
is the pairwise distances all our servers $x$.

Note that if we sum \ref{eq:step} for each step the potentials cancel out and we are left with,

$$
\DC(\sigma) + \phi^{last} - \phi^{first} \leq k \OPT(\sigma) + C
$$
Since $\phi$ is bounded we get,

$$
\DC(\sigma) \leq k \OPT(\sigma) + C
$$

For ease of exposition we break each request into two stages, in stage one OPT moves, 
and stage two DC moves. Now consider the stage 1, where OPT moves distance $d$ to service the 
request, consider the changes of each

\begin{itemize}
		\item $\Delta\OPT = d$, since OPT moved $d$
		\item $\Delta\DC = 0$, since we didn't move yet
		\item $\Delta\phi_1 \leq kd$, since in the worse case, the cost of $M(x,y)$ increased by $d$
		\item $\Delta\phi_2 = 0$, since we didn't move yet
\end{itemize}

From this we get,

\begin{align*}
	\Delta\DC^i(\sigma) + \Delta\phi^{i} &\leq k \Delta\OPT^i(\sigma)\\
	0 + kd  + 0 &\leq kd\\
\end{align*}

Hence for stage one, 
$$
\Delta\DC^i(\sigma) + \Delta\phi^{i} \leq k \Delta\OPT^i(\sigma)
$$

For stage two we have two cases, 
\begin{enumerate}
		\item The request is outside the servers, in which case, we send one server 
		\item The request is between two servers, in which case, we send two servers
\end{enumerate}

For case one we have, suppose we move one server $d'$ to service the request,

\begin{itemize}
		\item $\Delta\OPT = 0$, since OPT didn't move, 
		\item $\Delta\DC = d'$, since we moved a single server $d$
		\item $\Delta\phi_1 = -kd'$, since we moved to the same point as one OPT's servers to service the request
		\item $\Delta\phi_2 = (k-1)d'$, since we moved one server away from all the others
\end{itemize}

Summing these terms we get,
\begin{align*}
	\Delta\DC^i(\sigma) + \Delta\phi^{i} &\leq k \Delta\OPT^i(\sigma)\\
	d' - kd'  + (k-1)d' &\leq 0\\
	0 &\leq 0\\
\end{align*}

For case two suppose again the we move both servers $d'$,

\begin{itemize}
		\item $\Delta\OPT = 0$, since OPT didn't move, 
		\item $\Delta\DC = 2d'$, since we moved two servers $d'$
		\item $\Delta\phi_1 \leq 0$, since we moved to the same point as one OPT's servers to service the request, but then moved another server possible away from the one it was matched to
		\item $\Delta\phi_2 = -2d'$, since two servers moved closer to each other and the distances to between the servers that moved and didn't move cancel out
\end{itemize}
Summing these terms we get,
\begin{align*}
	\Delta\DC^i(\sigma) + \Delta\phi^{i} &\leq k \Delta\OPT^i(\sigma)\\
	2d' - 2d'  + 0 &\leq 0\\
	0 &\leq 0\\
\end{align*}

Therefore, we have shown that at each step, 
$$
\Delta\DC^i(\sigma) + \Delta\phi^{i} \leq k \Delta\OPT^i(\sigma)
$$

Which implies,

$$
\DC(\sigma) \leq k \OPT(\sigma) + C
$$

Which by definition means that DC is $k$-competitive. $\square$

\section{Convex Body Chasing}

We now look at a problem related to the k-server problem called convex body chasing which is,

\begin{itemize}
		\item you have a single server
		\item requests arrive online and are convex bodies
		\item service each request by move the server into the convex body
		\item minimize the distance your server travels for the sequence of requests
\end{itemize}

For a long time whether there was any competitive algorithm was an open problem. The first 
result to demonstrate a competitive algorithm was for chasing half-spaces (i.e. each convex body was a 
half-space) where the authors showed an algorithm that was $2^d$-competitive, where $d$ is the dimension 
of the space.

\textbf{Proof : } no algorithm is better than $\sqrt{d}$ competitive

Consider the sequence of requests, first is a halfspace, the second is a point on the boundary 
of the first halfspace. Clearly, the optimal solution is to just go to the second point immediately 
(since it satisfies both requests). However for any algorithm, no matter which point we go to 
first, the second request can be placed somewhere else the boundary of the halfspace, meaning the algorithm 
takes an $L$ shaped path, which is at best $\sqrt{d}$-competitive with the direct (optimal) path.

More recently in 2019 it was shown that this problem can always be solved with competitive ratio
$d$ by solving the special case where each request is nested inside the previous one. Creating an algorithm 
for this special case can then be adapted to the general case. The algorithm to show this is simply

\begin{itemize}
		\item Move to the Steiner Point of the request
\end{itemize}

\begin{definition}[Steiner Point]
	Each vertex of a convex body has a cone where it is the minimum point inside of the convex body if you are moving 
	in a direction within the cone. The Steiner Point is the average of the vertexes of the convex body, weighted by the size 
	of it's cone. 
\end{definition}

Mathmatically, for a convex body $k$, the Steiner point $St(k)$ is,
\begin{align*}
	St(k) &= \int_{||\theta|| = 1} \nabla_k(\theta) d\theta, \quad \nabla_k(\theta) = argmax_{x \in k}\{x \text{ is the minimum point moving in direction } \theta\}
\end{align*}

Where $|| \theta|| = 1$ is all vectors on the unit sphere. This can be rewritten as, 
\begin{align*}
	St(k) &= d \int_{||\theta|| = 1} S_k(\theta) \theta d\theta, \quad S_k(\theta) = max_{x \in k}\{x \text{ is the minimum point moving in direction } \theta\}
\end{align*}

\textbf{Claim :} For nested convex bodies, moving to $St(k)$ is d-competitive.

\textbf{Proof :}

We begin with the sum of the cost for each step, for a sequence of requests $\sigma = \{k_1, ..., k_T\}$,

\begin{align*}
	ALG(\sigma) &= \sum_{i=1}^T ||St(k_i) - St(k_{i-1})|| \quad \text{ the cost of moving from each steiner point to the other}\\
		&= \sum_{i=1}^T ||d \int_{||\theta|| = 1} (S_{k_i}(\theta) - S_{k_{i-1}}(\theta)) \theta d\theta || \quad \text{ replace with definition from above}\\
		&= d   \int_{||\theta|| = 1}\sum_{i=1}^T ||S_{k_i}(\theta) - S_{k_{i-1}}(\theta)  ||\theta d\theta \quad \text{move sum inside}\\
		&= d   \int_{||\theta|| = 1}\sum_{i=1}^T ||S_{k_i}(\theta) - S_{k_{i-1}}(\theta) || d\theta\quad ||\theta|| = 1\\
		&= d   \int_{||\theta|| = 1}||S_{k_T}(\theta) - S_{k_{1}}(\theta) || d\theta\quad \text{ the sum telescopes, leaving on the first and last term} \\ 
\end{align*}

It can then be shown that $\int_{||\theta|| = 1} S_{k_T}(\theta) - S_{k_{1}}(\theta) \leq \OPT + C$, hence the algorithm is $d$-competitive.

\end{document}

### 2022-04-19 proj/paper/paper.tex
add new stuff

\documentclass[a4paper]{article}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{paralist}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{paralist}
\usepackage[svgname]{xcolor}
\usepackage{enumerate}
\usepackage{array}
\usepackage{times}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{environ}
\usepackage{times}
\usepackage{textcomp}
\usepackage{caption}
\usepackage[square,numbers]{natbib}

\urlstyle{rm}

\setlength\parindent{0pt} % Removes all indentation from paragraphs

% TO SHOW SOLUTIONS, include following (else comment out):
\newenvironment{soln}{
    \leavevmode\color{blue}\ignorespaces
}{}

\hypersetup{
%    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

\geometry{
  top=1in,            % <-- you want to adjust this
  inner=1in,
  outer=1in,
  bottom=1in,
  headheight=3em,       % <-- and this
  headsep=2em,          % <-- and this
  footskip=3em,
}

\title{Stochastic Gradient Descent for Numerical MAX-CSP} % Title
\author{
Derek Paulsen \\
}

\date{}

\newcommand{\ind}[1]{\mathds{1}[#1]}
\def \eps{\epsilon}
\begin{document}
\maketitle

\section{Abstract}
\section{Introduction}
% TODO
\section{Background}

\subsubsection{Constraint satisfaction problems}

Constraint satisfaction problems (CSPs) encompasses many different problems. Formally a 
constraint satisfaction problem (CSP) is defined as a triple $\langle X, D, C \rangle$, where 
% XXX this is from wikipedia https://en.wikipedia.org/wiki/Constraint_satisfaction_problem
\begin{align*}
	X &= \{X_1, ..., X_n\} \text{ the set of variables}\\
	D &= \{D_1, ..., D_n\} \text{ the domains of each respective variable}\\
	C &= \{C_1, ..., C_m\} \text{ the set of constraints}\\
\end{align*}

A classic example of CSP, is the 3-SAT problem. In this case
the boolean variables that appear in at least one clause would be $X$. The domain for 
each variable would be $D_i = \{true, false\}$ and the constraints would all be 
three variable disjunctive clauses. The 3-SAT problem of course known to be NP-Complete.
In this paper we focus on a subset of CSPs called Numerical MAX-CSP. 
In typical a CSP all constraints must be satisfied, however in MAX-CSP the goal is simply to 
satisfy as may constraints as possible, additionally the domains are the variables 
are numerical (e.g. the real numbers).

While this problem has certainly been considered before, we were only able to find one previous 
work that directly address this issue. 
%TODO Add citation here

In this paper the authors propose a solution to numerical MAX-CSP. In particular where 
$D_i = \mathds{R}$ and each constraint $C_j = a_j^Tx \leq b_j, a_j \in \mathds{R}^m, b_j \in \mathds{R}$. 
That is, given a set of linear inequalities, satisfy as many as possible. In the paper the authors 
propose a specific algorithm for this problem based on branch and bound techniques. Unfortunately, 
this algorithm is worst case exponential time. In fact, this problem is easy to formulate as a mixed integer 
program,

\begin{align*}
\min_{x,z}\quad &1^Tz\\
s.t. \quad Ax - \epsilon z &\leq 0\\
		w &\in \mathds{R}^n\\
		z &\in \{0,1\}^m\\
\end{align*}

Where $\epsilon$ is some large number. The general problem of mixed integer linear programming is 
of course NP-Hard in general (by reduction to 0-1 integer programming which is NP-Complete). Hence 
this numerical MAX-CSP is likely not to admit an efficient solution.

\subsubsection{Stochastic Gradient Descent}

Stochastic gradient descent (SGD) is an optimization technique which has gained much 
attention in recent years due to its use in training neural networks for machine learning 
tasks such as computer vision and speech recognition. Despite this focus on machine learning,
stochastic gradient descent is a general purpose optimization procedure, capable of optimizing arbitrary 
differentiable functions. Many variations have been proposed in recent years, however
each technique uses the same basic idea. Given an input $X$ and a desired output $y$, take a sample $X'$ and $y'$, compute $f(X')$ and 
measure the difference between $f(X')$ and $y'$ (e.g. $|f(X') - y'|$) . Next compute the gradient $\nambla(|f(X') - y'|)$ 
of update the internal state of by taking a step in a direction of $-\nambla(|f(X') - y'|)$. Repeat this 
process until some criteria in met.

While SGD is not guaranteed and, frequently doesn't, find global optimums it is very effective at finding 
local minima for complex functions which are not convex. This makes it an invaluable tool for 
optimizing functions which don't admit exact optimization techniques such as those used in 
linear programs and quadratic programs.

\section{Motivation}

Numerical MAX-CSPs have a wide variety to applications in such as debugging infeasible 
linear programs, however we are interested in one particular problem, which is that of 
tuning the weights for full search engines.

Full text search engines are used in a plethora of applications. These search engines (such as Apache Lucene)
are built for efficient retrieval of top-k documents based on a TF/IDF based scoring metric which is dot 
product between two sparse vectors $q^Td = score$[[1]][1][[2]][2][[3]][3]. While this default scoring gives decent results out of the box, 
it is frequently augmented by re-weighting the query $q$, with some weight vector $w$
changing to scoring to be $(q\odot w)^Td = score$. This allows for boosting of certain terms or fields
to improve the quality of search results while not changing the underlying search algorithm.

The problem we will address in this project is finding a good weight vector $w$. In particular our
problem setting is as follows. We are giving a set of query vectors $Q = \{q_1, ... q_n\}$. For each 
of these query vectors $q_i$ we are given a set of $k$ retrieved document vectors with labels $R_i = \{(d_{i,1}, l_{i,1}), ..., (d_{i, k}, l_{i,k})\}$, 
$l \in \{relevant, irrelevant\}$.
We wish to find a weight vector $w$ that minimizes the number of irrelevant documents that are examined 
before the relevant result is found. This problem can be formulated as a mixed integer linear program (MILP)
of the following form
\begin{align*}
\min_{w,z}\quad &1^Tz\\
s.t. \quad Aw - \epsilon z &\leq 0\\
		w &\geq 0\\
		z &\in \{0,1\}\\
\end{align*}

\section{Problem}

As mentioned above, we can formulate our numerical MAX-CSP as a MILP, which can then be 
feed into an off the shelf solver. While this is simple solution (at least programming-wise), 
it has a major issue, namely that depending on the input, the solver time could be exponential in 
the number of constraints. For our motivating example, we would ideally feed in many labeled queries, 
each producing tens of constraints, hence exponential runtime in the number of constraints greatly 
reduces the usefulness of the solution. Moreover, because we have continuous and binary variables, 
the problem doesn't easily admit an LP relaxation approximation.

\section{Our Solution}

To remedy problem of runtime we propose a new solution based on 
stochstic gradient descent.

\section{Experiments}
\section{Discussion}
\section{Conclusions}

%\bibliographystyle{abbrvnat}
%\bibliography{references}
\end{document}

### 2022-04-19 proj/graph.py
bug fix and graphing first go

@@ -1,0 +1,57 @@
+import pandas as pd
+import json
+from pathlib import Path
+import matplotlib.pyplot as plt
+import numpy as np
+
+DATA_DIR = Path('./exp_res')
+# json data
+MILP = DATA_DIR / 'fixed_MILP.json'
+LP = DATA_DIR / 'LP.json'
+TORCH = DATA_DIR / 'torch.json'
+
+
+DF = None
+
+def process_time_series(x):
+
+    s = pd.DataFrame(x).astype(np.int32).set_index('time')['obj_val']
+    s = s.groupby(s.index).last()
+    return s
+
+
+def read_data(f):
+    with open(f) as ifs:
+        df = pd.DataFrame(list(map(json.loads,ifs))).set_index('dataset')
+
+    method_name = df['method_name'].iloc[0]
+    df['time_series'] = [process_time_series(x).rename(method_name) for x in df['time_series'].values]
+
+    return df
+
+
+def graph(milp, torch):
+    for ds in milp.index:
+        ts = pd.concat([
+            milp.at[ds, 'time_series'],
+            torch.at[ds, 'time_series']
+        ], axis=1).fillna(method='ffill').sort_index()
+        print(ts)
+        ax = ts.plot()
+        ax.set_title(ds)
+        ax.legend()
+        plt.show()
+    
+
+def main():
+    milp = read_data(MILP)
+    torch = read_data(TORCH)
+    graph(milp, torch)
+
+
+if __name__ == '__main__':
+    main()
+
+
+
+

### 2022-04-19 proj/run.py

@@ -51,9 +51,9 @@ def main(args):
     if args.k < 1:
         raise ValueError(args.k)
 
-    opt = get_optimizer(args.method)
     with open(args.out, 'a') as ofs:
         for ds in datasets:
+            opt = get_optimizer(args.method)
             res = run(opt, ds, args.k)
             ofs.write(json.dumps(res)) 
             ofs.write('\n')

### 2022-04-19 proj/graph.py
update graph script

@@ -3,12 +3,16 @@ import json
 from pathlib import Path
 import matplotlib.pyplot as plt
 import numpy as np
+from argparse import ArgumentParser
+
 
 DATA_DIR = Path('./exp_res')
 # json data
 MILP = DATA_DIR / 'fixed_MILP.json'
 LP = DATA_DIR / 'LP.json'
 TORCH = DATA_DIR / 'torch.json'
+argp = ArgumentParser()
+argp.add_argument('--show_graph', action='store_true')
 
 
 DF = None
@@ -29,28 +33,37 @@ def read_data(f):
 
     return df
 
+def print_stats(df, ds):
 
-def graph(milp, torch):
+    t_overtake = df['MILP'].lt(df['torch']).idxmax()
+    print(ds)
+    print(f'time for MILP to overtake torch : {t_overtake}')
+    print(f'time for MILP to reach best value: {df.MILP.idxmin()}')
+    print(f'time for torch to reach best value: {df.torch.idxmin()}')
+    print()
+
+def graph(milp, torch, show):
     for ds in milp.index:
         ts = pd.concat([
             milp.at[ds, 'time_series'],
             torch.at[ds, 'time_series']
         ], axis=1).fillna(method='ffill').sort_index()
-        print(ts)
-        ax = ts.plot()
-        ax.set_title(ds)
-        ax.legend()
-        plt.show()
-    
+        print_stats(ts, ds)
+        if show:
+            ax = ts.plot()
+            ax.set_title(ds)
+            ax.legend()
+            plt.show()
+        
 
-def main():
+def main(args):
     milp = read_data(MILP)
     torch = read_data(TORCH)
-    graph(milp, torch)
+    graph(milp, torch, args.show_graph)
 
 
 if __name__ == '__main__':
-    main()
+    main(argp.parse_args())
 
 
 

### 2022-04-19 proj/torch_optimizer.py
add debugging

@@ -21,7 +21,7 @@ class PosLinear(nn.Module):
         nn.Module.__init__(self)
 
         self._dim = in_features
-        self.weights = nn.Parameter(torch.rand(self._dim, dtype=torch.float32, requires_grad=True))
+        self.weights = nn.Parameter(torch.rand(self._dim, dtype=torch.float32, requires_grad=True) + .01)
 
     def forward(self, X):
         return torch.matmul(X, nn.functional.relu(self.weights))
@@ -48,8 +48,8 @@ class BoostModel(nn.Module):
 
 class TorchOptimizer(Optimizer):
 
-    def __init__(self, iters=500, step_interval=10, timeout=100):
-        self._timeout = Optimizer.TIMEOUT
+    def __init__(self, iters=500, step_interval=10, timeout=Optimizer.TIMEOUT):
+        self._timeout = timeout
         self._iters = iters
         # number iterations per round (gradient updates)
         self._step_interval = step_interval
@@ -126,6 +126,6 @@ class TorchOptimizer(Optimizer):
             cvs[i] = cv
         
         best_idx = np.argmin(cvs)
-
+        #log.debug(f'best idx : {best_idx}, value : {cvs[best_idx]}')
 
         return weights[best_idx], cvs[best_idx]

### 2022-04-19 proj/utils.py

@@ -7,7 +7,7 @@ logging.basicConfig(
         format='[%(filename)s:%(lineno)s - %(funcName)s() ] %(asctime)-15s : %(message)s',
 )
 
-log_level = logging.INFO
+log_level = logging.DEBUG
 logger = logging.getLogger(__name__)
 
 class Timer:

### 2022-04-21 proj/graph.py
change y limit

@@ -52,6 +52,7 @@ def graph(milp, torch, show):
         if show:
             ax = ts.plot()
             ax.set_title(ds)
+            ax.set_ylim(bottom=0.0)
             ax.legend()
             plt.show()
         

### 2022-04-22 proj/graph.py
whoops rans exps not on main

@@ -6,7 +6,7 @@ import numpy as np
 from argparse import ArgumentParser
 
 
-DATA_DIR = Path('./exp_res')
+DATA_DIR = Path('./exp_res/k=25')
 # json data
 MILP = DATA_DIR / 'fixed_MILP.json'
 LP = DATA_DIR / 'LP.json'

### 2022-04-22 proj/run.py

@@ -17,7 +17,7 @@ datasets = list(DATA_DIR.glob('*.parquet'))
 argp = ArgumentParser()
 
 argp.add_argument('--method', required=True, choices=['torch', 'MILP', 'LP'])
-argp.add_argument('--k', required=False, default=25, type=int)
+argp.add_argument('--k', required=False, default=50, type=int)
 argp.add_argument('--out', default='out.json')
 
 
@@ -42,6 +42,8 @@ def run(opt, f, k):
     res['dataset'] = f.stem
     res['dataset_path'] = str(f.absolute())
     res['time_ran'] = str(datetime.now())
+    res['num_constraints'] = len(const)
+    res['topk'] = k
     log.info(f'\n{pformat(res)}')
 
     return res

### 2022-04-22 proj/torch_optimizer.py

@@ -21,7 +21,7 @@ class PosLinear(nn.Module):
         nn.Module.__init__(self)
 
         self._dim = in_features
-        self.weights = nn.Parameter(torch.rand(self._dim, dtype=torch.float32, requires_grad=True) + .01)
+        self.weights = nn.Parameter(torch.rand(self._dim, dtype=torch.float32, requires_grad=True) + .1)
 
     def forward(self, X):
         return torch.matmul(X, nn.functional.relu(self.weights))
@@ -48,7 +48,7 @@ class BoostModel(nn.Module):
 
 class TorchOptimizer(Optimizer):
 
-    def __init__(self, iters=500, step_interval=10, timeout=Optimizer.TIMEOUT):
+    def __init__(self, iters=500, step_interval=25, timeout=Optimizer.TIMEOUT):
         self._timeout = timeout
         self._iters = iters
         # number iterations per round (gradient updates)
@@ -100,7 +100,7 @@ class TorchOptimizer(Optimizer):
         
         model = BoostModel(len(columns))
         loss_fn = nn.L1Loss()
-        optimizer = torch.optim.Adam(model.parameters(), lr=.01, weight_decay=0.0)
+        optimizer = torch.optim.Adam(model.parameters(), lr=.5, weight_decay=0.0)
         lr_decay = torch.optim.lr_scheduler.ExponentialLR(optimizer, gamma=1)
         
 
@@ -121,11 +121,11 @@ class TorchOptimizer(Optimizer):
             lr_decay.step()
             p = torch.sum(pred).cpu().detach().numpy()
             cv = torch.count_nonzero(pred > -1.0).cpu().detach().numpy()
-            log.debug(f'epoch {i} : sum = {p}, total > 0 = {cv}')
+            #log.debug(f'epoch {i} : sum = {p}, total > 0 = {cv}')
             weights[i] = model.weights
             cvs[i] = cv
         
         best_idx = np.argmin(cvs)
-        #log.debug(f'best idx : {best_idx}, value : {cvs[best_idx]}')
+        log.debug(f'best idx : {best_idx}, value : {cvs[best_idx]}, {cvs}')
 
         return weights[best_idx], cvs[best_idx]

### 2022-04-22 proj/graph.py
save graph

@@ -6,9 +6,9 @@ import numpy as np
 from argparse import ArgumentParser
 
 
-DATA_DIR = Path('./exp_res/k=25')
+DATA_DIR = Path('./exp_res/k=50')
 # json data
-MILP = DATA_DIR / 'fixed_MILP.json'
+MILP = DATA_DIR / 'MILP.json'
 LP = DATA_DIR / 'LP.json'
 TORCH = DATA_DIR / 'torch.json'
 argp = ArgumentParser()

### 2022-04-26 proj/run.py
add results for music

@@ -55,10 +55,11 @@ def main(args):
 
     with open(args.out, 'a') as ofs:
         for ds in datasets:
-            opt = get_optimizer(args.method)
-            res = run(opt, ds, args.k)
-            ofs.write(json.dumps(res)) 
-            ofs.write('\n')
+            if 'Music' in ds.name:
+                opt = get_optimizer(args.method)
+                res = run(opt, ds, args.k)
+                ofs.write(json.dumps(res)) 
+                ofs.write('\n')
 
 
 if __name__ == '__main__':

### 2022-04-26 proj/linprog_optimizer.py
increase constant to possibly fix num errors in Music dataset

@@ -116,7 +116,7 @@ class LinProgOptimizer(Optimizer):
             # a_i'x - z_i <= -1
             # remove coeffecients = 0 
             const = const[const != 0]
-            problem += const.mul(boost_vars[const.index]).sum() - 1000 * constraint_indicator_vars.at[idx] <= -1.0
+            problem += const.mul(boost_vars[const.index]).sum() - 10000 * constraint_indicator_vars.at[idx] <= -1.0
         
         # add query costs
         for i, query_cost in query_cost_vars.items():

### 2022-04-26 proj/run.py
add electronics

@@ -12,8 +12,12 @@ from utils import get_logger
 log = get_logger(__name__)
 
 DATA_DIR = Path('./data')
-datasets = list(DATA_DIR.glob('*.parquet'))
-
+#datasets = list(DATA_DIR.glob('*.parquet'))
+datasets = [
+        './data/Music.parquet', 
+        './data/electronics.parquet', 
+        './data/home.parquet', 
+]
 argp = ArgumentParser()
 
 argp.add_argument('--method', required=True, choices=['torch', 'MILP', 'LP'])
@@ -55,11 +59,10 @@ def main(args):
 
     with open(args.out, 'a') as ofs:
         for ds in datasets:
-            if 'Music' in ds.name:
-                opt = get_optimizer(args.method)
-                res = run(opt, ds, args.k)
-                ofs.write(json.dumps(res)) 
-                ofs.write('\n')
+            opt = get_optimizer(args.method)
+            res = run(opt, ds, args.k)
+            ofs.write(json.dumps(res)) 
+            ofs.write('\n')
 
 
 if __name__ == '__main__':

### 2022-04-26 proj/run.py
bug fix

@@ -37,6 +37,7 @@ def get_optimizer(t):
         raise ValueError(t)
 
 def run(opt, f, k):
+    f = Path(f)
     log.info(f'running {f}')
     const = Optimizer.read_constraints(f)
     const = Optimizer.truncate_topk(const, k)

### 2022-04-26 proj/linprog_optimizer.py
add more logging

@@ -72,6 +72,7 @@ class LinProgOptimizer(Optimizer):
         return timer.get_interval()
 
     def  _opt(self, constraints, relaxed):
+        timer = Timer()
         # the variables corresponding to the boost values in the 
         # query spec, this is the the main optimization target
         boost_vars = pd.Series(data=[
@@ -129,7 +130,9 @@ class LinProgOptimizer(Optimizer):
 
         # add objective function
         problem += query_cost_vars.sum()
-        
+        self._results['setup_time'] = timer.get_total()
+
+        log.info('starting solver')
         self._results['opt_time'] = self._solve_problem(problem)
         # get values to boosting variables from solution
         boost_map = boost_vars.apply(pulp.value)
@@ -138,7 +141,7 @@ class LinProgOptimizer(Optimizer):
 
 
     def optimize(self, constraints):
-
+        log.info('starting optimization')
         boost_map = self._opt(constraints, self._relaxed)
         boost_map = self.post_process_boost_map(boost_map)
         

### 2022-04-26 proj/torch_optimizer.py

@@ -69,6 +69,7 @@ class TorchOptimizer(Optimizer):
         X = torch.Tensor(constraints.values)
         label = -torch.ones(len(constraints), dtype=torch.float32)
         cands = []
+        self._results['setup_time'] = timer.get_total()
         while timer.get_total() < self._timeout:
             w, obj_val = self._optimize(constraints.columns, X, label)
             cands.append((w, obj_val, timer.get_total()))

### 2022-04-28 proj/graph.py
clean up exp res

@@ -26,7 +26,7 @@ def process_time_series(x):
 
 def read_data(f):
     with open(f) as ifs:
-        df = pd.DataFrame(list(map(json.loads,ifs))).set_index('dataset')
+        df = pd.DataFrame([json.loads(x) for x in ifs if len(x) > 1]).set_index('dataset')
 
     method_name = df['method_name'].iloc[0]
     df['time_series'] = [process_time_series(x).rename(method_name) for x in df['time_series'].values]
@@ -34,7 +34,7 @@ def read_data(f):
     return df
 
 def print_stats(df, ds):
-    if df['MILP'].lt(df['torch']).any():
+    if 'MILP' in df.columns and df['MILP'].lt(df['torch']).any():
         t_overtake = df['MILP'].lt(df['torch']).idxmax()
     else:
         t_overtake = np.nan

### 2022-04-28 proj/paper/paper.tex
bug found results

@@
 
+\newcommand{\red}[1]{{\color{red}#1}}
 \newcommand{\ind}[1]{\mathds{1}[#1]}
@@
 
-\subsubsection{Stochastic Gradient Descent}
+\subsubsection{Gradient Descent}
 
-Stochastic gradient descent (SGD) is an optimization technique which has gained much 
-attention in recent years due to its use in training neural networks for machine learning 
+%TODO rewrite this
+
+Gradient descent is an optimization technique which has gained much 
+attention in recent years due to the stochastic variant being used in training neural networks for machine learning 
 tasks such as computer vision and speech recognition. Despite this focus on machine learning,
-stochastic gradient descent is a general purpose optimization procedure, capable of optimizing arbitrary 
+gradient descent is a general purpose optimization procedure, capable of optimizing arbitrary 
 differentiable functions. Many variations have been proposed in recent years, however
-each technique uses the same basic idea. Given an input $X$ and a desired output $y$, take a sample $X'$ and $y'$, compute $f(X')$ and 
-measure the difference between $f(X')$ and $y'$ (e.g. $|f(X') - y'|$) . Next compute the gradient $\nambla(|f(X') - y'|)$ 
-of update the internal state of by taking a step in a direction of $-\nambla(|f(X') - y'|)$. Repeat this 
-process until some criteria in met.
+each technique uses the same basic idea. Given a function $f$ and current point $x$, compute $f(x)$. 
+Next compute the gradient $\nabla(f(x))$ and update $x$ taking a step in a direction of $-\eta \nabla(f(x))$. Repeat this 
+process until the minima of the function is found.
 
-While SGD is not guaranteed and, frequently doesn't, find global optimums it is very effective at finding 
-local minima for complex functions which are not convex. This makes it an invaluable tool for 
+If $f(x)$ is convex, then gradient descent is guaranteed to find an optimal solution. If $f(x)$ is not convex, 
+the gradient descent will converge to a local minimum. Although in the non-convex case, gradient descent 
+is not guaranteed, and frequently won't, find the global minimum, it is useful for finding 
+minima of functions which are cannot be handled by typical solvers.  This makes it an invaluable tool for 
 optimizing functions which don't admit exact optimization techniques such as those used in 
@@
 \min_{w,z}\quad &1^Tz\\
-s.t. \quad Aw - \epsilon z &\leq 0\\
+s.t. \quad Aw - \epsilon z &\leq \gamma\\
 		w &\geq 0\\
@@
 
-\section{Problem}
+\section{Baseline Solutions}
 
-As mentioned above, we can formulate our numerical MAX-CSP as a MILP, which can then be 
-feed into an off the shelf solver. While this is simple solution (at least programming-wise), 
-it has a major issue, namely that depending on the input, the solver time could be exponential in 
-the number of constraints. For our motivating example, we would ideally feed in many labeled queries, 
+As metioned above we can formulate our problem as a mixed integer linear program 
+with binary constraints. In addition, to the mixed integer formulation, we also create an LP
+relaxation of the problem as an approximation of the MILP, namely,
+
+\begin{align*}
+\min_{w,z}\quad &1^Tz\\
+s.t. \quad Aw - \epsilon z &\leq \gamma\\
+		w &\geq 0\\
+		0 \leq z &\leq 1\\
+\end{align*}
+
+The rationale behind this is that the LP relaxation can be solved efficiently and provide a 
+deterministic approximation of the MILP. Although the solution could be arbitrarily bad,
+we find that the LP typically finds a reasonable solution in much less time 
+relative to the MILP.
+
+\section{Problem with Baseline Solutions}
+
+While modeling the MAX-CSP problem as a MILP is simple solution (at least programming-wise), 
+it has a few major issues, namely that depending on the input, the solver time could be exponential in 
+the number of constraints. Moreover, even commercial solvers can struggle with large problem sizes, 
+either taking large amounts of time to produce any feasible solution or having prohibitive memory requirements.
+For our motivating example, we would ideally feed in many labeled queries, 
 each producing tens of constraints, hence exponential runtime in the number of constraints greatly 
-reduces the usefulness of the solution. Moreover, because we have continuous and binary variables, 
-the problem doesn't easily admit an LP relaxation approximation.
+reduces the usefulness of the solution.
 
@@
 
-To remedy problem of runtime we propose a new solution based on 
-stochstic gradient descent.
+To remedy problem of runtime we propose a new solution based on gradient descent. The high level
+idea of the solution is simple, begin with a random weight feasible vector $w$ and then perform 
+gradient descent until the local minima is found. More specifically we have,
+
+$$
+f(x) = \text{HardSigmoid}(Aw - \gamma)
+$$
+Where, 
+$$
+x\in R, \quad \text{HardSigmoid}(x) = \begin{cases} 0 \text{ if } x \leq 0\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
+$$
+
+For gradient updates we start with $\eta = .5$ which decays exponentially every 25 gradient updates. We do a total 
+of 100 gradient updates and then back track to find the $w$ (i.e. the one that satisfies as many constraints as 
+possible). We repeat this process starting with random $w$ until we reach a preset time limit.
 
 \section{Experiments}
+
+We ran our experiments on a server with an AMD Ryzen Threadripper 2950X (16c/32t) processor with 
+64GB of RAM. Our MAX-CSP instances are generated from real world datasets, with the number of constraints 
+ranging from \red{TODO : max num const} to \red{TODO : max num const}. For the baseline solutions we 
+leverage Gurobi, which takes advantage of all threads available.
+
 \section{Discussion}

### 2022-04-28 proj/graph.py

@@ -14,13 +14,15 @@ TORCH = DATA_DIR / 'torch.json'
 argp = ArgumentParser()
 argp.add_argument('--show_graph', action='store_true')
 
-
+TIME_LIMIT = 2700
 DF = None
 
-def process_time_series(x):
+def process_time_series(row):
 
-    s = pd.DataFrame(x).astype(np.int32).set_index('time')['obj_val']
+    s = pd.DataFrame(row['time_series']).astype(np.int32).set_index('time')['obj_val']
     s = s.groupby(s.index).last()
+    s = s.loc[s.index < TIME_LIMIT]
+    s[TIME_LIMIT] = row['total_violated']
     return s
 
 
@@ -29,21 +31,21 @@ def read_data(f):
         df = pd.DataFrame([json.loads(x) for x in ifs if len(x) > 1]).set_index('dataset')
 
     method_name = df['method_name'].iloc[0]
-    df['time_series'] = [process_time_series(x).rename(method_name) for x in df['time_series'].values]
+    df['time_series'] = [process_time_series(x[1]).rename(method_name) for x in df.iterrows()]
 
     return df
 
-def print_stats(df, ds):
-    if 'MILP' in df.columns and df['MILP'].lt(df['torch']).any():
-        t_overtake = df['MILP'].lt(df['torch']).idxmax()
+def print_stats(ts, ds):
+    if 'MILP' in ts.columns and ts['MILP'].lt(ts['torch']).any():
+        t_overtake = ts['MILP'].lt(ts['torch']).idxmax()
     else:
         t_overtake = np.nan
 
     print(ds)
     print(f'time for MILP to overtake torch : {t_overtake}')
-    print(f'time for MILP to reach best value: {df.MILP.idxmin()}')
-    print(f'time for torch to reach best value: {df.torch.idxmin()}')
-    print()
+    print(f'time for MILP to reach best value: {ts.MILP.idxmin()}')
+    print(f'time for torch to reach best value: {ts.torch.idxmin()}')
+    print(ts)
 
 def graph(milp, torch, show):
     print(milp.index)

### 2022-04-28 proj/optimizer.py

@@ -17,7 +17,7 @@ class Optimizer(ABC):
     
     @staticmethod
     def create_results(constraints, weights, method_name , time_series):
-        violated = constraints.mul(weights).sum(axis=1).sub(1.0).gt(0.0)
+        violated = constraints.mul(weights).sum(axis=1).gt(-1.0)
         hist = violated.groupby(level=0).sum().value_counts()
         weights.index = [str(x) for x in weights.index]
         return {

### 2022-04-28 proj/fix_res.py
fix res again? might be incorrect

@@ -42,7 +42,7 @@ def fix_res(dat):
     const = Optimizer.read_constraints(f)
     const = Optimizer.truncate_topk(const, dat['topk'])
 
-    dat['total_violated'] = Optimizer.create_results(const, weights, dat['method_name'], pd.DataFrame(columns=['time', 'obj_val']))
+    dat['total_violated'] = Optimizer.create_results(const, weights, dat['method_name'], pd.DataFrame(columns=['time', 'obj_val']))['total_violated']
 
     return dat
 

### 2022-04-28 proj/linprog_optimizer.py

@@ -93,20 +93,6 @@ class LinProgOptimizer(Optimizer):
             ],
             index=constraints.index
         )
-        # the costs for violating a given constraint
-        constraint_costs = constraint_indicator_vars
-        
-        # the cost for a given matching tuple in a query
-        sub_query_index = pd.MultiIndex.from_frame(constraints.index.to_frame(index=None).iloc[:, 0:2].drop_duplicates())
-        sub_query_cost_vars = pd.Series(
-                [pulp.LpVariable(f'y_{i}_{j}', lowBound=0) for i,j in sub_query_index],
-                index = sub_query_index
-        ).sort_index()
-        # the total cost for a query
-        query_cost_vars = pd.Series(
-                [pulp.LpVariable(f'y_{i}', lowBound=0) for i in constraints.index.get_level_values(0).unique()],
-                index = constraints.index.get_level_values(0).unique()
-        )
         
         problem = pulp.LpProblem('boosting_weights', pulp.LpMinimize)
 
@@ -119,17 +105,9 @@ class LinProgOptimizer(Optimizer):
             const = const[const != 0]
             problem += const.mul(boost_vars[const.index]).sum() - 10000 * constraint_indicator_vars.at[idx] <= -1.0
         
-        # add query costs
-        for i, query_cost in query_cost_vars.items():
-            query_cost = query_cost_vars[i]
-            # add constriants to compute the sub query costs
-            for j, sub_query_cost in sub_query_cost_vars.loc[i].items():
-                problem += constraint_costs.loc[(i,j,)].sum() - sub_query_cost <= 0
-            
-            problem += sub_query_cost_vars.loc[i].sum() - query_cost <= 0.0
 
         # add objective function
-        problem += query_cost_vars.sum()
+        problem += constraint_indicator_vars.sum()
         self._results['setup_time'] = timer.get_total()
 
         log.info('starting solver')

### 2022-04-28 proj/torch_optimizer.py

@@ -120,7 +120,7 @@ class TorchOptimizer(Optimizer):
                 optimizer.step()
 
             lr_decay.step()
-            p = torch.sum(pred).cpu().detach().numpy()
+            #p = torch.sum(pred).cpu().detach().numpy()
             cv = torch.count_nonzero(pred > -1.0).cpu().detach().numpy()
             #log.debug(f'epoch {i} : sum = {p}, total > 0 = {cv}')
             weights[i] = model.weights

### 2022-04-28 proj/optimizer.py
fix gurobi precision issue

@@ -5,7 +5,7 @@ import pandas as pd
 
 class Optimizer(ABC):
 
-    TIMEOUT=2700 # 24hrs of cpu time
+    TIMEOUT=60 # 24hrs of cpu time
     
     @abstractproperty
     def results(self):
@@ -18,6 +18,8 @@ class Optimizer(ABC):
     @staticmethod
     def create_results(constraints, weights, method_name , time_series):
         violated = constraints.mul(weights).sum(axis=1).gt(-1.0)
+        # correct for float inprecision 
+        time_series['obj_val'] = np.maximum(violated.sum(), time_series['obj_val'].values)
         hist = violated.groupby(level=0).sum().value_counts()
         weights.index = [str(x) for x in weights.index]
         return {
@@ -63,4 +65,4 @@ class Optimizer(ABC):
     @staticmethod
     def post_process_boost_map(boost_map):
         boost_map = boost_map.where(boost_map > 0).dropna()
-        return boost_map / boost_map.min()
+        return boost_map 

### 2022-04-28 proj/run.py

@@ -12,12 +12,7 @@ from utils import get_logger
 log = get_logger(__name__)
 
 DATA_DIR = Path('./data')
-#datasets = list(DATA_DIR.glob('*.parquet'))
-datasets = [
-        './data/Music.parquet', 
-        './data/electronics.parquet', 
-        './data/home.parquet', 
-]
+datasets = list(DATA_DIR.glob('*.parquet'))
 argp = ArgumentParser()
 
 argp.add_argument('--method', required=True, choices=['torch', 'MILP', 'LP'])
@@ -59,7 +54,7 @@ def main(args):
         raise ValueError(args.k)
 
     with open(args.out, 'a') as ofs:
-        for ds in datasets:
+        for ds in datasets[1:]:
             opt = get_optimizer(args.method)
             res = run(opt, ds, args.k)
             ofs.write(json.dumps(res)) 

### 2022-04-28 proj/torch_optimizer.py

@@ -48,7 +48,7 @@ class BoostModel(nn.Module):
 
 class TorchOptimizer(Optimizer):
 
-    def __init__(self, iters=100, step_interval=25, timeout=Optimizer.TIMEOUT):
+    def __init__(self, iters=500, step_interval=25, timeout=Optimizer.TIMEOUT):
         self._timeout = timeout
         self._iters = iters
         # number iterations per round (gradient updates)
@@ -67,7 +67,7 @@ class TorchOptimizer(Optimizer):
     def optimize(self, constraints):
         timer = Timer()
         X = torch.Tensor(constraints.values)
-        label = -torch.ones(len(constraints), dtype=torch.float32)
+        label = -torch.ones(len(constraints), dtype=torch.float64)
         cands = []
         self._results['setup_time'] = timer.get_total()
         while timer.get_total() < self._timeout:

### 2022-04-28 proj/linprog_optimizer.py
change timeout back and decrease the gamma to 1000 to try and avoid numerical issues with gurobi

@@ -103,7 +103,7 @@ class LinProgOptimizer(Optimizer):
             # a_i'x - z_i <= -1
             # remove coeffecients = 0 
             const = const[const != 0]
-            problem += const.mul(boost_vars[const.index]).sum() - 10000 * constraint_indicator_vars.at[idx] <= -1.0
+            problem += const.mul(boost_vars[const.index]).sum() - 1000 * constraint_indicator_vars.at[idx] <= -1.0
         
 
         # add objective function

### 2022-04-28 proj/optimizer.py

@@ -5,7 +5,7 @@ import pandas as pd
 
 class Optimizer(ABC):
 
-    TIMEOUT=60 # 24hrs of cpu time
+    TIMEOUT=2700 # 24hrs of cpu time
     
     @abstractproperty
     def results(self):

### 2022-04-28 proj/optimizer.py
change to float 32

@@ -33,7 +33,7 @@ class Optimizer(ABC):
 
     @staticmethod
     def read_constraints(file):
-        constraints = pd.read_parquet(file)
+        constraints = pd.read_parquet(file).astype(np.float32)
         # begins as (id2, matching tuple, id1)
         constraints.index = pd.MultiIndex.from_tuples(list(map(eval, constraints.index)))
         constraints.columns = pd.MultiIndex.from_tuples(list(map(eval, constraints.columns)))

### 2022-04-28 proj/torch_optimizer.py

@@ -67,7 +67,7 @@ class TorchOptimizer(Optimizer):
     def optimize(self, constraints):
         timer = Timer()
         X = torch.Tensor(constraints.values)
-        label = -torch.ones(len(constraints), dtype=torch.float64)
+        label = -torch.ones(len(constraints), dtype=torch.float32)
         cands = []
         self._results['setup_time'] = timer.get_total()
         while timer.get_total() < self._timeout:

### 2022-04-28 proj/optimizer.py
more refinements

@@ -17,7 +17,8 @@ class Optimizer(ABC):
     
     @staticmethod
     def create_results(constraints, weights, method_name , time_series):
-        violated = constraints.mul(weights).sum(axis=1).gt(-1.0)
+        # we can always scale values to make the margin as wide as we want
+        violated = constraints.mul(weights).sum(axis=1).ge(0)
         # correct for float inprecision 
         time_series['obj_val'] = np.maximum(violated.sum(), time_series['obj_val'].values)
         hist = violated.groupby(level=0).sum().value_counts()
@@ -65,4 +66,4 @@ class Optimizer(ABC):
     @staticmethod
     def post_process_boost_map(boost_map):
         boost_map = boost_map.where(boost_map > 0).dropna()
-        return boost_map 
+        return boost_map

### 2022-04-28 proj/torch_optimizer.py

@@ -121,7 +121,7 @@ class TorchOptimizer(Optimizer):
 
             lr_decay.step()
             #p = torch.sum(pred).cpu().detach().numpy()
-            cv = torch.count_nonzero(pred > -1.0).cpu().detach().numpy()
+            cv = torch.count_nonzero(pred >= 0).cpu().detach().numpy()
             #log.debug(f'epoch {i} : sum = {p}, total > 0 = {cv}')
             weights[i] = model.weights
             cvs[i] = cv

### 2022-04-30 proj/paper/paper.tex
change opt

@@
 \usepackage{amsmath}
+\usepackage{booktabs}
 \usepackage{dsfont}
@@
 
-\setlength\parindent{0pt} % Removes all indentation from paragraphs
-
 % TO SHOW SOLUTIONS, include following (else comment out):
@@
 
-\title{Stochastic Gradient Descent for Numerical MAX-CSP} % Title
+\title{Gradient Descent for Numerical MAX-CSP} % Title
 \author{
@@
 \section{Introduction}
+
 % TODO
+
+Many real world problems can be modeled as constraint 
+satisfaction problems. The problems range from logic puzzles 
+like sudoku to tuning search engines. Briefly, a constraint satisfaction 
+problem is simply the task of finding a valid assignment of variables 
+which satisfy as set of constraints. Many classic problems in theoretical computer 
+science fit this model, such as the NP-Complete 3-SAT and circuit SAT problems. In this paper 
+look variant of CSP, called numerical MAX-CSP. In this variant each variable is a real number
+and all of the constraints are linear inequalities, the goal then is to satisfy as 
+many linear inequalities as possible.
+
+The rest of this paper is then structured as follows. First, we give background on the problem 
+and the basis for our solution. Next we motivate solving this problem by tying it to 
+tuning a keyword search engine with labeled data. We then move on to describe a baseline 
+solution by modeling the problem as mixed integer linear program and discuss the issues with 
+this solution. Next, we present our algorithm and compare it to the baseline solution. Finally, 
+we discuss the experimental results and give directions for future work.
+
 \section{Background}
 
-\subsubsection{Constraint satisfaction problems}
+\subsection{Constraint satisfaction problems}
 
@@
 \end{align*}
-
-A classic example of CSP, is the 3-SAT problem. In this case
+A classic example of a CSP, is the 3-SAT problem. In this case
 the boolean variables that appear in at least one clause would be $X$. The domain for 
@@
 While this problem has certainly been considered before, we were only able to find one previous 
-work that directly address this issue. 
+work that directly addresses this issue. 
 %TODO Add citation here
@@
 \min_{x,z}\quad &1^Tz\\
-s.t. \quad Ax - \epsilon z &\leq 0\\
+s.t. \quad Ax - \epsilon z &\leq b\\
 		w &\in \mathds{R}^n\\
@@
 
-\subsubsection{Gradient Descent}
-
-%TODO rewrite this
+\subsection{Gradient Descent}
 
@@
 
+Where each row, $a$, of $A$ comes corresponds to 
+$$
+a = (d_{i,y} - d_{i,x}) \odot q_i  \quad i \in \{1,...n\} \quad x,y \in \{1,...,k\}, l_{i,x} = relevant, l_{i,y} = irrelevant
+$$
+
+That is, the pairwise difference between the components of a matching and non-matching document, multiplied by the 
+relevant query. We note that with this setup, 
+$$
+a^Tw < 0 \implies (q\odot w)^Td_{i,x} > (q\odot w)^Td_{i,y}
+$$
+That is, the relevant document will be scored higher than the non-relevant document for the query. Solving this 
+MILP then corresponds to minimizing the number of irrelevant documents that need to be 
+examined in order to find the relevant document for each query.
+
 \section{Baseline Solutions}
@@
 $$
-Where, 
+Where $ \gamma = -1$ and, 
 $$
@@
 For gradient updates we start with $\eta = .5$ which decays exponentially every 25 gradient updates. We do a total 
-of 100 gradient updates and then back track to find the $w$ (i.e. the one that satisfies as many constraints as 
-possible). We repeat this process starting with random $w$ until we reach a preset time limit.
+of 500 gradient updates and then back track to find the best $w$ (i.e. the one that satisfies as many constraints as 
+possible). We then repeat this process starting with random $w$ until we reach a preset time limit.
 
@@
 64GB of RAM. Our MAX-CSP instances are generated from real world datasets, with the number of constraints 
-ranging from \red{TODO : max num const} to \red{TODO : max num const}. For the baseline solutions we 
-leverage Gurobi, which takes advantage of all threads available.
+ranging from 7769 to 5994778. For the baseline solutions we 
+leverage Gurobi, a commercial solver with an academic license. We choose gurobi because it is consistently 
+one of the top performing available solvers \cite{benchmark}.
 
-\section{Discussion}
-\section{Conclusions}
+\begin{tabular}{|l|r|r|r|}
+\toprule
+Dataset &  \# of Constraints &  \# of Columns &  \% Non-Zero \\
+\midrule
+         $D_0$ &                   7769 &                 14 &         66.24 \\
+         $D_1$ &                  19350 &                 18 &         61.46 \\
+         $D_2$ &                 111974 &                 18 &         72.83 \\
+         $D_3$ &                 196213 &                 30 &         51.62 \\
+         $D_4$ &                3711208 &                124 &         25.09 \\
+         $D_5$ &                5994778 &                 22 &         68.58 \\
+\bottomrule
+\end{tabular}
 
-%\bibliographystyle{abbrvnat}
-%\bibliography{references}
+\bibliographystyle{abbrvnat}
+\bibliography{references}
 \end{document}

### 2022-04-30 proj/optimizer.py

@@ -17,7 +17,7 @@ class Optimizer(ABC):
     
     @staticmethod
     def create_results(constraints, weights, method_name , time_series):
-        violated = constraints.mul(weights).sum(axis=1).gt(-1.0)
+        violated = constraints.mul(weights).sum(axis=1).ge(0)
         hist = violated.groupby(level=0).sum().value_counts()
         weights.index = [str(x) for x in weights.index]
         return {

### 2022-05-01 proj/exp_res/old/k=25/fix_milp.py
add updated exp results

@@ -1,0 +1,20 @@
+import json 
+import pandas as pd
+
+res = list(reversed(list(map(json.loads, open('./MILP.json')))))
+
+for i in range(1, len(res)):
+    prev = pd.DataFrame(res[i-1]['time_series'])
+    curr = pd.DataFrame(res[i]['time_series'])
+    res[i-1]['time_series'] = prev.iloc[len(curr):].to_dict(orient='list')
+
+with open('fixed_MILP.json', 'w') as ofs:
+    for r in reversed(res):
+        ofs.write(json.dumps(r))
+        ofs.write('\n')
+
+
+
+
+
+

### 2022-05-01 proj/graph.py

@@ -6,10 +6,10 @@ import numpy as np
 from argparse import ArgumentParser
 
 
-DATA_DIR = Path('./exp_res/k=50')
+DATA_DIR = Path('./exp_res/')
 # json data
-MILP = DATA_DIR / 'MILP_fixed.json'
-LP = DATA_DIR / 'LP_fixed.json'
+MILP = DATA_DIR / 'MILP.json'
+LP = DATA_DIR / 'LP.json'
 TORCH = DATA_DIR / 'torch_fixed.json'
 argp = ArgumentParser()
 argp.add_argument('--show_graph', action='store_true')

### 2022-05-01 proj/run.py

@@ -54,7 +54,7 @@ def main(args):
         raise ValueError(args.k)
 
     with open(args.out, 'a') as ofs:
-        for ds in datasets[1:]:
+        for ds in datasets[:1]:
             opt = get_optimizer(args.method)
             res = run(opt, ds, args.k)
             ofs.write(json.dumps(res)) 

### 2022-05-01 proj/paper/paper.tex
save paper

@@
 
-% TODO
+% TODO clean up
 
@@
 
-\section{Background}
+\section{Preliminaries}
+
+In this section we will give an overview of constraint satisfaction problems,
+and the specific flavor that we are attempting to tackle in this paper. We then 
+give a brief overview gradient descent which is the basis of our proposed solution.
 
@@
 Constraint satisfaction problems (CSPs) encompasses many different problems. Formally a 
-constraint satisfaction problem (CSP) is defined as a triple $\langle X, D, C \rangle$, where 
-% XXX this is from wikipedia https://en.wikipedia.org/wiki/Constraint_satisfaction_problem
+constraint satisfaction problem (CSP) is defined as a triple $\langle X, D, C \rangle$ \cite{wiki_CSP}, where 
 \begin{align*}
@@
 each variable would be $D_i = \{true, false\}$ and the constraints would all be 
-three variable disjunctive clauses. The 3-SAT problem of course known to be NP-Complete.
+three variable disjunctive clauses. 
 In this paper we focus on a subset of CSPs called Numerical MAX-CSP. 
@@
 While this problem has certainly been considered before, we were only able to find one previous 
-work that directly addresses this issue. 
-%TODO Add citation here
-
-In this paper the authors propose a solution to numerical MAX-CSP. In particular where 
+work that directly addresses this issue \cite{num_max_csp_paper}.
+In this paper the authors propose a exact solution to numerical MAX-CSP. In particular where 
 $D_i = \mathds{R}$ and each constraint $C_j = a_j^Tx \leq b_j, a_j \in \mathds{R}^m, b_j \in \mathds{R}$. 
 That is, given a set of linear inequalities, satisfy as many as possible. In the paper the authors 
-propose a specific algorithm for this problem based on branch and bound techniques. Unfortunately, 
-this algorithm is worst case exponential time. In fact, this problem is easy to formulate as a mixed integer 
-program,
+propose a specific algorithm for this problem based on branch and bound techniques. While the algorithm 
+does produce optimal solutions, the algorithm is worst case exponential time.
+In fact, this problem is readily to formulated as a mixed integer linear program (MILP),
 
@@
 Where $\epsilon$ is some large number. The general problem of mixed integer linear programming is 
-of course NP-Hard in general (by reduction to 0-1 integer programming which is NP-Complete). Hence 
-this numerical MAX-CSP is likely not to admit an efficient solution.
+NP-Hard in general (by reduction to 0-1 integer programming which is NP-Complete \cite{karp_1972}). Hence 
+numerical MAX-CSP is likely not to admit an efficient solution.
 
@@
 are built for efficient retrieval of top-k documents based on a TF/IDF based scoring metric which is dot 
-product between two sparse vectors $q^Td = score$[[1]][1][[2]][2][[3]][3]. While this default scoring gives decent results out of the box, 
+product between two sparse vectors $q^Td = score$\cite{lucene_history} \cite{WAND_paper} \cite{block_max_WAND_paper}.
+While this default scoring gives decent results out of the box, 
 it is frequently augmented by re-weighting the query $q$, with some weight vector $w$
@@
 before the relevant result is found. This problem can be formulated as a mixed integer linear program (MILP)
-of the following form
+as above but with fixed $\gamma < 0$ instead of $b$, in particular,
 \begin{align*}
@@
 with binary constraints. In addition, to the mixed integer formulation, we also create an LP
-relaxation of the problem as an approximation of the MILP, namely,
-
-\begin{align*}
-\min_{w,z}\quad &1^Tz\\
-s.t. \quad Aw - \epsilon z &\leq \gamma\\
-		w &\geq 0\\
-		0 \leq z &\leq 1\\
-\end{align*}
-
+relaxation of the problem as an approximation of the MILP.
 The rationale behind this is that the LP relaxation can be solved efficiently and provide a 
@@
 $$
-f(x) = \text{HardSigmoid}(Aw - \gamma)
+f(w) = \text{HardTanh}(Aw - \gamma)
 $$
@@
 $$
-x\in R, \quad \text{HardSigmoid}(x) = \begin{cases} 0 \text{ if } x \leq 0\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
+x\in R, \quad \text{HardTanh}(x) = \begin{cases} -1 \text{ if } x \leq -1\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
 $$
@@
 
+\section{Discussion}
+\section{Future Work}
+
 \bibliographystyle{abbrvnat}

### 2022-05-01 proj/graph.py
add warning message to graph.py

@@ -10,7 +10,7 @@ DATA_DIR = Path('./exp_res/')
 # json data
 MILP = DATA_DIR / 'MILP.json'
 LP = DATA_DIR / 'LP.json'
-TORCH = DATA_DIR / 'torch_fixed.json'
+TORCH = DATA_DIR / 'torch.json'
 argp = ArgumentParser()
 argp.add_argument('--show_graph', action='store_true')
 
@@ -50,6 +50,10 @@ def print_stats(ts, ds):
 def graph(milp, torch, show):
     print(milp.index)
     for ds in milp.index:
+        if ds not in torch.index:
+            print(f'SKIPPING {ds}')
+            continue
+
         ts = pd.concat([
             milp.at[ds, 'time_series'],
             torch.at[ds, 'time_series']

### 2022-05-01 proj/create_dataset_table.py
add script

@@ -1,0 +1,45 @@
+import pandas as pd
+from pathlib import Path
+
+files = Path('./data/').glob('*.parquet')
+
+#'Walmart-Amazon' 
+
+NAME_TO_ID = dict(
+    zip(['Abt-Buy',
+    'Amazon-Google', 
+    'DBLP-GoogleScholar', 
+    'uwhealth', 
+    'electronics', 
+    'Music'],
+
+    range(10))
+)
+
+print(NAME_TO_ID)
+
+
+def make_row(f):
+    df = pd.read_parquet(f)
+
+    return {
+            'Dataset Name' : f.stem,
+            'Number of Constraints' : len(df),
+            'Number of Columns' : len(df.columns),
+            'Percent Non-Zero' : df.ne(0).mean().mean() * 100
+    }
+
+table = pd.DataFrame(list(map(make_row, files)))
+
+print(table)
+
+table = table.loc[table['Dataset Name'].isin(NAME_TO_ID)]
+table.index = table['Dataset Name'].apply(NAME_TO_ID.__getitem__).values
+table = table.sort_index()
+
+table['Dataset Name'] = table['Dataset Name'].apply(lambda x : f'D_{NAME_TO_ID[x]}')
+print(table)
+
+print(table.to_latex(index=False, column_format='|l|r|r|r|'))
+
+

### 2022-05-02 proj/create_dataset_table.py
save

@@ -1,23 +1,11 @@
 import pandas as pd
 from pathlib import Path
+from utils import NAME_TO_ID
 
 files = Path('./data/').glob('*.parquet')
 
 #'Walmart-Amazon' 
-
-NAME_TO_ID = dict(
-    zip(['Abt-Buy',
-    'Amazon-Google', 
-    'DBLP-GoogleScholar', 
-    'uwhealth', 
-    'electronics', 
-    'Music'],
-
-    range(10))
-)
-
 print(NAME_TO_ID)
-
 
 def make_row(f):
     df = pd.read_parquet(f)

### 2022-05-02 proj/graph.py

@@ -4,6 +4,7 @@ from pathlib import Path
 import matplotlib.pyplot as plt
 import numpy as np
 from argparse import ArgumentParser
+from utils import NAME_TO_ID
 
 
 DATA_DIR = Path('./exp_res/')
@@ -28,7 +29,10 @@ def process_time_series(row):
 
 def read_data(f):
     with open(f) as ifs:
-        df = pd.DataFrame([json.loads(x) for x in ifs if len(x) > 1]).set_index('dataset')
+        df = pd.DataFrame([json.loads(x) for x in ifs if len(x) > 1])
+
+    df['dataset_id'] = df['dataset'].apply(lambda x : NAME_TO_ID.get(x))
+    df = df.loc[df['dataset_id'].notnull()].set_index('dataset_id')
 
     method_name = df['method_name'].iloc[0]
     df['time_series'] = [process_time_series(x[1]).rename(method_name) for x in df.iterrows()]
@@ -47,21 +51,22 @@ def print_stats(ts, ds):
     print(f'time for torch to reach best value: {ts.torch.idxmin()}')
     print(ts)
 
+
 def graph(milp, torch, show):
     print(milp.index)
-    for ds in milp.index:
-        if ds not in torch.index:
-            print(f'SKIPPING {ds}')
+    for name, i in NAME_TO_ID.items():
+        if i not in torch.index or i not in milp.index:
+            print(f'SKIPPING {name}')
             continue
 
         ts = pd.concat([
-            milp.at[ds, 'time_series'],
-            torch.at[ds, 'time_series']
+            milp.at[i, 'time_series'],
+            torch.at[i, 'time_series']
         ], axis=1).fillna(method='ffill').sort_index()
-        print_stats(ts, ds)
+        print_stats(ts, name)
         if show:
             ax = ts.plot()
-            ax.set_title(ds)
+            ax.set_title(name)
             ax.set_ylim(bottom=0.0)
             ax.legend()
             plt.show()

### 2022-05-02 proj/utils.py

@@ -2,6 +2,17 @@ import sys
 import logging
 import time
 
+NAME_TO_ID = dict(
+    zip(['Abt-Buy',
+    'Amazon-Google', 
+    'DBLP-GoogleScholar', 
+    'uwhealth', 
+    'electronics', 
+    'Music'],
+
+    range(10))
+)
+
 logging.basicConfig(
         stream=sys.stderr,
         format='[%(filename)s:%(lineno)s - %(funcName)s() ] %(asctime)-15s : %(message)s',

### 2022-05-02 proj/graph.py
new graph

@@ -2,11 +2,15 @@ import pandas as pd
 import json
 from pathlib import Path
 import matplotlib.pyplot as plt
+import matplotlib
 import numpy as np
 from argparse import ArgumentParser
 from utils import NAME_TO_ID
+from optimizer import Optimizer
+#matplotlib.rc('text', usetex=True)
+#matplotlib.rc('font', family='serif')
 
-
+TIME_LIMIT = Optimizer.TIMEOUT
 DATA_DIR = Path('./exp_res/')
 # json data
 MILP = DATA_DIR / 'MILP.json'
@@ -15,16 +19,17 @@ TORCH = DATA_DIR / 'torch.json'
 argp = ArgumentParser()
 argp.add_argument('--show_graph', action='store_true')
 
-TIME_LIMIT = 2700
 DF = None
 
 def process_time_series(row):
 
     s = pd.DataFrame(row['time_series']).astype(np.int32).set_index('time')['obj_val']
     s = s.groupby(s.index).last()
-    s = s.loc[s.index < TIME_LIMIT]
-    s[TIME_LIMIT] = row['total_violated']
-    return s
+    time = min(TIME_LIMIT, int(row['opt_time']))
+    s = s.loc[s.index < time]
+    s[time] = row['total_violated']
+
+    return s / row['num_constraints'] * 100
 
 
 def read_data(f):
@@ -52,30 +57,58 @@ def print_stats(ts, ds):
     print(ts)
 
 
-def graph(milp, torch, show):
-    print(milp.index)
+def iter_time_series(milp, torch):
     for name, i in NAME_TO_ID.items():
         if i not in torch.index or i not in milp.index:
-            print(f'SKIPPING {name}')
             continue
-
         ts = pd.concat([
             milp.at[i, 'time_series'],
             torch.at[i, 'time_series']
-        ], axis=1).fillna(method='ffill').sort_index()
+        ], axis=1).sort_index()
+        idx = ts['MILP'].last_valid_index()
+        ts['MILP'].loc[:idx] = ts['MILP'].loc[:idx].fillna(method='ffill')
+        ts['torch']= ts['torch'].fillna(method='ffill')
+
+        yield name, ts
+
+
+def graph(milp, torch, show):
+    print(milp.index)
+    for name, ts in iter_time_series(milp, torch):
         print_stats(ts, name)
         if show:
             ax = ts.plot()
             ax.set_title(name)
-            ax.set_ylim(bottom=0.0)
[…]

### 2022-05-02 proj/paper/paper.tex
new graph
save paper

@@
 \usepackage{geometry}
+\usepackage{algpseudocode}
+\usepackage{algorithm}
+\usepackage{algorithmicx}
 \usepackage{graphicx}
@@
 \usepackage{fancyvrb}
-\usepackage{algorithm}
-\usepackage{algorithmic}
 \usepackage{float}
@@
 
-\section{Baseline Solutions}
+\section{Baseline Solution}
 
-As metioned above we can formulate our problem as a mixed integer linear program 
-with binary constraints. In addition, to the mixed integer formulation, we also create an LP
-relaxation of the problem as an approximation of the MILP.
-The rationale behind this is that the LP relaxation can be solved efficiently and provide a 
-deterministic approximation of the MILP. Although the solution could be arbitrarily bad,
-we find that the LP typically finds a reasonable solution in much less time 
-relative to the MILP.
-
-\section{Problem with Baseline Solutions}
-
-While modeling the MAX-CSP problem as a MILP is simple solution (at least programming-wise), 
-it has a few major issues, namely that depending on the input, the solver time could be exponential in 
-the number of constraints. Moreover, even commercial solvers can struggle with large problem sizes, 
-either taking large amounts of time to produce any feasible solution or having prohibitive memory requirements.
-For our motivating example, we would ideally feed in many labeled queries, 
-each producing tens of constraints, hence exponential runtime in the number of constraints greatly 
-reduces the usefulness of the solution.
+As motioned above we can formulate our problem as an  MILP with binary
+constraints. This MILP can then be feed into any supported off the shelf solver
+to produce an exact optimal answer.  This solution is appealing because,
+assuming that the solver is correct, it will give an optimal solution, however
+it suffers from a few major drawbacks.  First and foremast, depending on the
+input, the solver time could be exponential in the number of constraints.
+Moreover we found that even commercial solvers can struggle with large problem
+sizes, either taking large amounts of time to produce any feasible solution or
+having prohibitive memory requirements.  For our motivating use case, we would
+ideally feed in many labeled queries, each producing tens of constraints, hence
+exponential runtime in the number of constraints greatly reduces the usefulness
+of the solution.
 
@@
 
-To remedy problem of runtime we propose a new solution based on gradient descent. The high level
-idea of the solution is simple, begin with a random weight feasible vector $w$ and then perform 
-gradient descent until the local minima is found. More specifically we have,
+To remedy problem of runtime we propose a new solution based on gradient
+descent. The high level idea of the solution is simple, begin with a random
+weight feasible vector $w$ and then perform gradient descent until the local
+minima is found. More specifically we minimize the following function,
 
 $$
-f(w) = \text{HardTanh}(Aw - \gamma)
-$$
-Where $ \gamma = -1$ and, 
-$$
-x\in R, \quad \text{HardTanh}(x) = \begin{cases} -1 \text{ if } x \leq -1\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
+f(w) = \text{HardTanh}(Aw - \gamma)^T\overrightarrow{1}
 $$
 
-For gradient updates we start with $\eta = .5$ which decays exponentially every 25 gradient updates. We do a total 
-of 500 gradient updates and then back track to find the best $w$ (i.e. the one that satisfies as many constraints as 
-possible). We then repeat this process starting with random $w$ until we reach a preset time limit.
+$$
+\text{Where }  \gamma = -1 \text{ and for } x\in R,  \text{HardTanh}(x) = \begin{cases} -1 \text{ if } x \leq -1\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
+$$
 
+\algrenewcommand\textproc{}
+\algrenewcommand{\algorithmiccomment}[1]{\hfill// #1}
+
+\begin{algorithm}[H]
+\caption{OptimizeGD($f$, $T$)}
+\textbf{Input : } The function $f$ to optimize, a time limit $T$
+\textbf{Output : } A local minima $w^*$
+\begin{algorithmic}[1]
+	\State $w^* \gets \overrightarrow{0}$ 
+	\While{current time $ < T$}
+		\State $w \gets w \sim \mathds{U}_{[.1,1.1]}^n$ \Comment{Initialize $w$ to a uniform random vector}
+		\For{$i =  1,...20$}
+			\For{$j =  1,...25$}
+				\State $w \gets w - \text{ADAM}(\nabla f(w))$
+			\EndFor
+			\If{ $f(w) < f(w^*)$}
+				\State $w^* \gets w$
+			\EndIf
+		\EndFor
+	\EndWhile
+	\\\Return $w^*$
+\end{algorithmic}
+\end{algorithm}
+
+We use the ADAM optimizer \cite{ADAM_paper} because we emprically found it to perform well,
+while giving minimal overhead. 
+%TODO add pseduo code?
 \section{Experiments}
@@
 64GB of RAM. Our MAX-CSP instances are generated from real world datasets, with the number of constraints 
-ranging from 7769 to 5994778. For the baseline solutions we 
+ranging from 7.7K to 6.0M. For the baseline solutions we 
 leverage Gurobi, a commercial solver with an academic license. We choose gurobi because it is consistently 
-one of the top performing available solvers \cite{benchmark}.
-
+one of the top performing available solvers \cite{benchmark}. Both solutions where given a timeout of 2700 seconds 
+(corresponding to 24 hours of CPU time).
+\begin{table}[ht!]
+\caption{Dataset statistics}
 \begin{tabular}{|l|r|r|r|}
@@
 \midrule
-         $D_0$ &                   7769 &                 14 &         66.24 \\
-         $D_1$ &                  19350 &                 18 &         61.46 \\
-         $D_2$ &                 111974 &                 18 &         72.83 \\
-         $D_3$ &                 196213 &                 30 &         51.62 \\
-         $D_4$ &                3711208 &                124 &         25.09 \\
-         $D_5$ &                5994778 &                 22 &         68.58 \\
+         $D_0$ &                   7.7K &                 14 &         66.24 \\
+         $D_1$ &                  19.3K &                 18 &         61.46 \\
+         $D_2$ &                 111.9K &                 18 &         72.83 \\
+         $D_3$ &                 196.2K &                 30 &         51.62 \\
+         $D_4$ &                3.7M    &                124 &         25.09 \\
+         $D_5$ &                6.0M    &                 22 &         68.58 \\
 \bottomrule
 \end{tabular}
+\end{table}
+
+\begin{figure}[ht!]
+	\caption{Comparison of MILP vs. Gradient Descent (GD)}
+	\includegraphics[width=\textwidth]{./time_series.png}
+\end{figure}
 
 \section{Discussion}
+
+In this section we will discuss the experimental results. There are
+two major points. We first discuss the quality of solutions produced. 
+Next, we discuss scalabilty of solutions.
+
+\subsection{Quality of Solutions}
+
+Much to our surprise, the quality of solutions produced by gradient descent
+(GD) was very similar if not better than the MILP solution. We attribute this
+to the fact that only one dataset was solved to optimality ($D_2$), in fact, we
+let $D_0$ run for over 12 hours and MILP still hadn't found an optimal
+solution. This was rather suprising because, $D_0$ has less an order of
+magnitude fewer constraints than $D_2$. This underlines up a key trade off
+between the solutions, namely, quality of solutions vs. runtime predictability.
+While the MILP solution will produce optimal solutions given enough time, it is
+very hard to predict how long it will take to produce an optimal solution. On
+the otherhand, there are no guarantees that can be made about GD in terms of
+the quality of solutions but its runtime is very predictable, scaling roughly
+linearly with the number of constraints times the number of columns.
+
+\subsection{Scalability}
+
+This then brings us to our second point which is the scalability of the 
+solutions. To begin, from $D_4$ and $D_5$, we can clearly see that 
+GD has superior scaling properties when compared to MILP in terms of 
+runtime. MILP timed out on both datasets before even begining to refine the 
+solutions produced. We attribute this again to the vastly different complexity of the 
+two solutions. While GD has linear complexity in the number of constraints, 
+MILP has best case polynomial complexity in the number of constraints.  For small 
+problems this not an issue, but as the problem size increases MILP quickly becomes
+impractical for even approximating the optimal solution.
+
 \section{Future Work}
+
+We identify multiple directions for future work, which fall into 
+two categories, optimization techniques, and problem applications.
+
+\subsection{Optimization Techniques}
+
+In this paper we have only taken a small look at the possible 
+optimization techniques that could be applied. The most obvious 
+variation that could tried is to do stochastic gradient descent. This 
+variation could have two major potential benefits. First, it can 
+decrease both runtime and memory requirements of the solution further by no longer requiring the 
+entire dataset be processed on each gradient update. Second, the stochastic 
+version may be able to find better solutions as it is much more likely that it 
+could escape worse local minima, instead of relying mostly on having a good starting point.
+
+\subsection{Problem Applications}
+
+We restricted our problem setting to numerical MAX-CSPs with linear
+constraints, however in principle there is no reason why this technique could
+not be applied to arbitrary constraints. In particular, non-convex constraints
+would potentially be solved by applying similar techniques, which current
+solvers are not able to handle in an capacity. We believe that it is likely
+that modifying the optimzation procedure, such as using stochastic gradient
+descent, will likely be required to provide good approximations of such
+functions.
+
+\section{Conclusion}
+
+In this work we have taken brief look at a particular kind of CSP, namely, 
+numerical MAX-CSP's. We motivated this problem with a real world use case for 
+tuning full text search engines. We then proposed a solution based on gradient 
+descent and gave preliminary evidence to show that it has some key advantages 
+over the baseline MILP based solution of the problem, especially for large problem 
+sizes where commercial solvers are too slow to practical. We believe that our solution 
+shows promise for finding approximate solutions to problems which were not previously tractable
+with traditional MILP based solutions. Future work should continue to explore the 
+possible variations and possibly apply similar techniques to related problems.
 

### 2022-05-02 proj/graph.py

@@ -68,7 +68,8 @@ def iter_time_series(milp, torch):
         idx = ts['MILP'].last_valid_index()
         ts['MILP'].loc[:idx] = ts['MILP'].loc[:idx].fillna(method='ffill')
         ts['torch']= ts['torch'].fillna(method='ffill')
-
+        
+        ts = ts.rename(columns={'torch' : 'GD'})
         yield name, ts
 
 
@@ -89,11 +90,22 @@ def graph_six(milp, torch):
     
     for ax, ts_data, i in zip(axes.flatten(), time_series, range(len(time_series))):
         name, ts = ts_data
-        ts.plot(ax=ax)
+        ts.plot(ax=ax, logx=False)
         ax.set_title(f'$D_{i}$')
         ax.set_xlabel('Time (seconds)')
         ax.set_ylabel('Percent of Constraints Violated')
         ax.legend()
+        text = "Time to First Solution"
+        for c in ts.columns:
+            u = ts[c].drop_duplicates()
+            ax.scatter(u.index, u.values)
+            # first soln
+            x = u.first_valid_index()
+            text += f"\n{c} = {x} seconds"
+
+        ax.annotate(text, (.5, .5), xycoords='axes fraction')
+            
+            
     
     fig.tight_layout()
     fig.savefig('./paper/time_series.png')

### 2022-05-04 proj/paper/paper.tex
first edit of paper done

@@
 
-% TO SHOW SOLUTIONS, include following (else comment out):
-\newenvironment{soln}{
-    \leavevmode\color{blue}\ignorespaces
-}{}
-
 \hypersetup{
@@
 
-\section{Abstract}
+%\section{Abstract}
 \section{Introduction}
 
-% TODO clean up
+% TODO clean up, still clean up
 
-Many real world problems can be modeled as constraint 
-satisfaction problems. The problems range from logic puzzles 
-like sudoku to tuning search engines. Briefly, a constraint satisfaction 
-problem is simply the task of finding a valid assignment of variables 
-which satisfy as set of constraints. Many classic problems in theoretical computer 
-science fit this model, such as the NP-Complete 3-SAT and circuit SAT problems. In this paper 
-look variant of CSP, called numerical MAX-CSP. In this variant each variable is a real number
-and all of the constraints are linear inequalities, the goal then is to satisfy as 
-many linear inequalities as possible.
+Many real world problems can be modeled as constraint satisfaction problems (CSPs),
+ranging from logic puzzles like sudoku to machine learning. Briefly, a
+constraint satisfaction problem is the task of finding a valid assignment of
+variables which satisfy a set of constraints. Many classic problems in
+theoretical computer science fit this model, such as the 3-SAT and circuit SAT
+problems. As such, large amounts of work have been
+devoted to studying CSPs and how to solve them, leading to many exact and
+approximation algorithms.  In this paper we examine a lesser studied variant of
+CSPs, called numerical MAX-CSPs. In this variant each variable is a real number
+and all of the constraints are linear inequalities, the goal then is to satisfy
+as many linear inequalities as possible. We find that while it is less
+studied than other CSPs, numerical MAX-CSPs have direct applications to 
+a variety real world problems, such as solving over-constrained LP's and 
+selecting weights for full text search engines.
 
-The rest of this paper is then structured as follows. First, we give background on the problem 
-and the basis for our solution. Next we motivate solving this problem by tying it to 
-tuning a keyword search engine with labeled data. We then move on to describe a baseline 
-solution by modeling the problem as mixed integer linear program and discuss the issues with 
-this solution. Next, we present our algorithm and compare it to the baseline solution. Finally, 
-we discuss the experimental results and give directions for future work.
+The rest of this paper is structured as follows. We begin with background
+on the problem and the basis for our proposed solution. Next, we motivate our particular
+problem by tying it to tuning a keyword search engine with labeled data. We
+then move on to describe a baseline solution by modeling the problem as mixed
+integer linear program and discuss the issues with this solution. Next, we
+present our algorithm and compare it to the baseline solution. Finally, we
+discuss the experimental results and give directions for future work.
 
-\section{Preliminaries}
+\section{Background}
 
-In this section we will give an overview of constraint satisfaction problems,
-and the specific flavor that we are attempting to tackle in this paper. We then 
-give a brief overview gradient descent which is the basis of our proposed solution.
-
-\subsection{Constraint satisfaction problems}
-
-Constraint satisfaction problems (CSPs) encompasses many different problems. Formally a 
-constraint satisfaction problem (CSP) is defined as a triple $\langle X, D, C \rangle$ \cite{wiki_CSP}, where 
-\begin{align*}
-	X &= \{X_1, ..., X_n\} \text{ the set of variables}\\
-	D &= \{D_1, ..., D_n\} \text{ the domains of each respective variable}\\
-	C &= \{C_1, ..., C_m\} \text{ the set of constraints}\\
-\end{align*}
-A classic example of a CSP, is the 3-SAT problem. In this case
-the boolean variables that appear in at least one clause would be $X$. The domain for 
-each variable would be $D_i = \{true, false\}$ and the constraints would all be 
-three variable disjunctive clauses. 
-In this paper we focus on a subset of CSPs called Numerical MAX-CSP. 
-In typical a CSP all constraints must be satisfied, however in MAX-CSP the goal is simply to 
-satisfy as may constraints as possible, additionally the domains are the variables 
-are numerical (e.g. the real numbers).
-
-While this problem has certainly been considered before, we were only able to find one previous 
-work that directly addresses this issue \cite{num_max_csp_paper}.
-In this paper the authors propose a exact solution to numerical MAX-CSP. In particular where 
-$D_i = \mathds{R}$ and each constraint $C_j = a_j^Tx \leq b_j, a_j \in \mathds{R}^m, b_j \in \mathds{R}$. 
-That is, given a set of linear inequalities, satisfy as many as possible. In the paper the authors 
-propose a specific algorithm for this problem based on branch and bound techniques. While the algorithm 
-does produce optimal solutions, the algorithm is worst case exponential time.
-In fact, this problem is readily to formulated as a mixed integer linear program (MILP),
-
-\begin{align*}
-\min_{x,z}\quad &1^Tz\\
-s.t. \quad Ax - \epsilon z &\leq b\\
-		w &\in \mathds{R}^n\\
-		z &\in \{0,1\}^m\\
-\end{align*}
-
-Where $\epsilon$ is some large number. The general problem of mixed integer linear programming is 
-NP-Hard in general (by reduction to 0-1 integer programming which is NP-Complete \cite{karp_1972}). Hence 
-numerical MAX-CSP is likely not to admit an efficient solution.
+In this section we give a brief overview gradient descent which
+is the basis of our proposed solution. We then give
+an overview of constraint satisfaction problems,
+and the specific variant that we address in this paper.
 
@@
 
-Gradient descent is an optimization technique which has gained much 
-attention in recent years due to the stochastic variant being used in training neural networks for machine learning 
-tasks such as computer vision and speech recognition. Despite this focus on machine learning,
-gradient descent is a general purpose optimization procedure, capable of optimizing arbitrary 
-differentiable functions. Many variations have been proposed in recent years, however
-each technique uses the same basic idea. Given a function $f$ and current point $x$, compute $f(x)$. 
-Next compute the gradient $\nabla(f(x))$ and update $x$ taking a step in a direction of $-\eta \nabla(f(x))$. Repeat this 
-process until the minima of the function is found.
+Gradient descent is an optimization technique which has gained much attention
+in recent years due to  its use in training neural
+networks for machine learning tasks such as computer vision and speech
+recognition. Despite the focus on machine learning, gradient descent is a
+general purpose optimization procedure, capable of optimizing arbitrary
+differentiable functions. Many variations have been proposed in recent years,
+however each technique uses the same core idea. Given a function $f$ to minimize and
+current point $x$, compute the gradient $\nabla(f(x))$
+and update $x$ taking a step in a direction of $-\eta \nabla(f(x))$. Repeat
+this process until the minimum of the function is found.
 
-If $f(x)$ is convex, then gradient descent is guaranteed to find an optimal solution. If $f(x)$ is not convex, 
-the gradient descent will converge to a local minimum. Although in the non-convex case, gradient descent 
-is not guaranteed, and frequently won't, find the global minimum, it is useful for finding 
-minima of functions which are cannot be handled by typical solvers.  This makes it an invaluable tool for 
-optimizing functions which don't admit exact optimization techniques such as those used in 
-linear programs and quadratic programs.
+If $f(x)$ is convex, then gradient descent is guaranteed to find an optimal
+solution. If $f(x)$ is not convex, the gradient descent will converge to a
+local minimum. Although in the non-convex case, gradient descent is not
+guaranteed, and frequently won't, find the global minimum, it is still useful for
+approximating the global minimum of functions which are cannot be handled by solvers.
+This makes it an invaluable tool for optimizing functions which don't admit
+exact optimization techniques such as those used in linear programs and
+quadratic programs.
 
-\section{Motivation}
+\subsection{CSPs and MAX-CSPs}
 
-Numerical MAX-CSPs have a wide variety to applications in such as debugging infeasible 
-linear programs, however we are interested in one particular problem, which is that of 
-tuning the weights for full search engines.
+Constraint satisfaction problems (CSPs) encompasses many different problems. Formally, a 
+CSP is defined as a triple $\langle X, D, C \rangle$ \cite{wiki_CSP}, where 
+\begin{align*}
+	X &= \{X_1, ..., X_n\} \text{ the set of variables}\\
+	D &= \{D_1, ..., D_n\} \text{ the domains of each variable}\\
+	C &= \{C_1, ..., C_m\} \text{ the set of constraints}\\
+\end{align*}
 
-Full text search engines are used in a plethora of applications. These search engines (such as Apache Lucene)
-are built for efficient retrieval of top-k documents based on a TF/IDF based scoring metric which is dot 
-product between two sparse vectors $q^Td = score$\cite{lucene_history} \cite{WAND_paper} \cite{block_max_WAND_paper}.
-While this default scoring gives decent results out of the box, 
-it is frequently augmented by re-weighting the query $q$, with some weight vector $w$
-changing to scoring to be $(q\odot w)^Td = score$. This allows for boosting of certain terms or fields
-to improve the quality of search results while not changing the underlying search algorithm.
+A classic example of a CSP, is the 3-SAT problem. In this case, 
+$X_i$ corresponds to a boolean variable, which has domain $D_i = \{true, false\}$
+and the constraints are all three variable disjunctive clauses. If we change the objective to satisfying as 
+many constraints as possible we get the MAX-3-SAT problem which is a type of a 
+MAX-CSP. More generally, a MAX-CSP is simply a CSP where the objective is to 
+satisfy as many constraints as possible instead of being required to satisfy all 
+of them.
 
-The problem we will address in this project is finding a good weight vector $w$. In particular our
-problem setting is as follows. We are giving a set of query vectors $Q = \{q_1, ... q_n\}$. For each 
+\section{Related Work}
+
+In this paper we focus on a subset of MAX-CSPs called numerical MAX-CSPs.  While
+MAX-CSPs have been studied extensively and there exist many approximation
+algorithms \cite{approx_for_csp_paper} for a variety of problems, most work has
+been focus on MAX-CSPs with finite domains (e.g. boolean assignment of
+variables).  In contrast, we were only 
+able to find one previous work that directly addresses numerical MAX-CSPs
+\cite{num_max_csp_paper}.  In this paper the authors propose an exact algorithm
+for solving numerical MAX-CSP instances. In particular, where $D_i =
+\mathds{R}$ and each constraint $C_j = a_j^Tx \leq b_j, a_j \in \mathds{R}^n,
+b_j \in \mathds{R}$.  That is, given a set of linear inequalities, satisfy as
+many as possible. While the algorithm does
+produce optimal solutions, the algorithm is worst case exponential time.
+
+We note that this problem can be is readily formulated as a mixed integer
+linear program (MILP) with binary indicator variables. The general problem of
+mixed integer linear programming is NP-Hard (by reduction to 0-1
+integer programming which is NP-Complete \cite{karp_1972}). While there has
+been little work directly addressing numerical MAX-CSPs, there has been a
+large body of literature aiming to improve MILP solvers. In particular, there
+has been a lot of attention given to improving the branch and bound algorithms
+used by many MILP solvers. These improvements typically aim to provide better
+branching heuristics for branch and bound algorithms, such as strong branching
+\cite{strong_branching_paper}. More recent work has examined machine learning
+for better branching heuristics \cite{bengio_lodi_prouvost_2021}
+\cite{liberto_kadioglu_leo_malitsky_2016}, either to approximate the strong
+branching using machine learning (to reduce the computation cost) or to learn
+ completely new branching heuristics.
+
+\section{Problem and Motivation}
+
+Numerical MAX-CSPs have a wide variety to applications from debugging
+infeasible linear programs to machine learning classification, however we are
+interested in one particular problem, which is tuning the weights for
+full text search engines.  Full text search engines are used in a plethora of
+applications. These search engines (such as Apache Lucene) are built for
+efficient retrieval of top-k documents based on a TF/IDF based scoring metric
+which is dot product between two sparse vectors $q^Td =
+score$\cite{lucene_history} \cite{WAND_paper} \cite{block_max_WAND_paper}.
+While this default scoring gives decent results out of the box, it is
+frequently augmented by re-weighting the query $q$, with some weight vector $w$
+changing to scoring to be $(q\odot w)^Td = score$ (where $\odot$ is element-wise multiplication).
+This allows for boosting of certain terms or fields to improve the quality of
+search results while not changing the underlying search algorithm.
+
+The problem is then to find a good non-negative weight vector $w$. In particular our
+problem setting is as follows. We are giving a set of query vectors $Q = \{q_1, ... q_n\}, q_i \in \mathds{R}^n$. For each 
 of these query vectors $q_i$ we are given a set of $k$ retrieved document vectors with labels $R_i = \{(d_{i,1}, l_{i,1}), ..., (d_{i, k}, l_{i,k})\}$, 
-$l \in \{relevant, irrelevant\}$.
+where, $d_i \in \mathds{R}^n$ and $l \in \{relevant, irrelevant\}$.
 We wish to find a weight vector $w$ that minimizes the number of irrelevant documents that are examined 
-before the relevant result is found. This problem can be formulated as a mixed integer linear program (MILP)
-as above but with fixed $\gamma < 0$ instead of $b$, in particular,
+before the relevant document is found. This problem can be formulated as a mixed integer linear program (MILP) as follows,
 \begin{align*}
@@
 
-Where each row, $a$, of $A$ comes corresponds to 
+Where $\gamma$ is a negative fixed constant, $\epsilon$ is a large positive constant for 
+indicating constraint violation, and each row, $a$, of $A$ comes corresponds to 
 $$
@@
 $$
-That is, the relevant document will be scored higher than the non-relevant document for the query. Solving this 
-MILP then corresponds to minimizing the number of irrelevant documents that need to be 
-examined in order to find the relevant document for each query.
+In plain English, the relevant document will be scored higher than the
+non-relevant document for the query. Solving this MILP then corresponds to
+minimizing the number of irrelevant documents that need to be examined in order
+to find the relevant document for each query.
 
@@
 
-As motioned above we can formulate our problem as an  MILP with binary
+As mentioned above, we can formulate our problem as an MILP with binary
 constraints. This MILP can then be feed into any supported off the shelf solver
-to produce an exact optimal answer.  This solution is appealing because,
-assuming that the solver is correct, it will give an optimal solution, however
-it suffers from a few major drawbacks.  First and foremast, depending on the
-input, the solver time could be exponential in the number of constraints.
-Moreover we found that even commercial solvers can struggle with large problem
-sizes, either taking large amounts of time to produce any feasible solution or
-having prohibitive memory requirements.  For our motivating use case, we would
-ideally feed in many labeled queries, each producing tens of constraints, hence
-exponential runtime in the number of constraints greatly reduces the usefulness
-of the solution.
+to produce an optimal answer.  This solution to our problem, which we will
+refer to as MILP when there is no ambiguity, is appealing because, assuming that the solver is
+correct, it will give an optimal solution, however it suffers from a few major
+drawbacks.  First and foremost, depending on the input, the runtime time is worst case
+exponential in the number of constraints.  Moreover, we found that even
+commercial solvers can struggle with large problem sizes, either taking large
+amounts of time to produce any feasible solution or simply not terminating in a
+reasonable amount of time. For our motivating use case, we would ideally feed
+in many labeled queries, each producing tens of constraints, hence exponential
+runtime in the number of constraints greatly reduces the usefulness of the
+solution.
 
@@
 
-To remedy problem of runtime we propose a new solution based on gradient
+To address the problems of the MILP algorithm, we propose a new algorithm based on gradient
 descent. The high level idea of the solution is simple, begin with a random
-weight feasible vector $w$ and then perform gradient descent until the local
+feasible weight vector $w$ and then perform gradient descent until the local
 minima is found. More specifically we minimize the following function,
@@
 $$
-f(w) = \text{HardTanh}(Aw - \gamma)^T\overrightarrow{1}
+f(w) = \langle \text{HardTanh}(Aw - \gamma), \overrightarrow{1}\rangle 
 $$
@@
 $$
-\text{Where }  \gamma = -1 \text{ and for } x\in R,  \text{HardTanh}(x) = \begin{cases} -1 \text{ if } x \leq -1\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
+\text{Where }  \gamma = -1 \text{ and }  \text{HardTanh}(x) = \begin{cases} -1 \text{ if } x \leq -1\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
 $$
@@
 \caption{OptimizeGD($f$, $T$)}
-\textbf{Input : } The function $f$ to optimize, a time limit $T$
-\textbf{Output : } A local minima $w^*$
+\textbf{Input : } The function $f$ to optimize, the time limit $T$\\
+\textbf{Output : } A local minimum $w^*$
 \begin{algorithmic}[1]
@@
 				\State $w \gets w - \text{ADAM}(\nabla f(w))$ 
+				\State $w \gets (\max\{0, w_1\},..., \max\{0, w_n\})$ \Comment{Make all elements of $w$ non-negative}
 			\EndFor
-			\If{ $f(w) < f(w^*)$}
-				\State $w^* \gets w$
+			\If{$f(w) < f(w^*)$}
+				\State $w^* \gets w$ \Comment{Update best solution}
 			\EndIf
@@
 
-We use the ADAM optimizer \cite{ADAM_paper} because we emprically found it to perform well,
-while giving minimal overhead. 
-%TODO add pseduo code?
+For our implementation we use PyTorch \cite{pytorch} for its high performance
+and flexibility. For gradient updates we use the ADAM gradient descent algorithm with default parameters
+\cite{ADAM_paper} because we empirically found it to perform well relative to
+other available gradient descent algorithms, while giving minimal overhead.
+
 \section{Experiments}
 
-We ran our experiments on a server with an AMD Ryzen Threadripper 2950X (16c/32t) processor with 
-64GB of RAM. Our MAX-CSP instances are generated from real world datasets, with the number of constraints 
-ranging from 7.7K to 6.0M. For the baseline solutions we 
-leverage Gurobi, a commercial solver with an academic license. We choose gurobi because it is consistently 
-one of the top performing available solvers \cite{benchmark}. Both solutions where given a timeout of 2700 seconds 
-(corresponding to 24 hours of CPU time).
+We ran our experiments on a server with an AMD Ryzen Threadripper 2950X
+(16c/32t) processor with 64GB of RAM. Our MAX-CSP instances are generated from
+real world datasets, with the number of constraints ranging from 7.7K to 6.0M.
+For MILP we leverage Gurobi, a commercial solver with an
+academic license. We choose Gurobi because it is consistently one of the top
+performing solvers \cite{benchmark} with an academic license. Both solutions where given a
+timeout of 2700 seconds (corresponding to 24 hours of CPU time with 32 threads).
+
 \begin{table}[ht!]
@@
 
-In this section we will discuss the experimental results. There are
-two major points. We first discuss the quality of solutions produced. 
-Next, we discuss scalabilty of solutions.
+In this section we discuss the two major takeaways from the experimental results,
+the quality of solutions produced and the scalability of the algorithms.
 
@@
 Much to our surprise, the quality of solutions produced by gradient descent
-(GD) was very similar if not better than the MILP solution. We attribute this
-to the fact that only one dataset was solved to optimality ($D_2$), in fact, we
-let $D_0$ run for over 12 hours and MILP still hadn't found an optimal
-solution. This was rather suprising because, $D_0$ has less an order of
-magnitude fewer constraints than $D_2$. This underlines up a key trade off
+(GD) were very similar, and in some cases even better, than MILP. We attribute this
+to the fact that only a single dataset was solved to optimality in the 
+allotted time ($D_2$) with MILP.  In fact, despite having an order of magnitude 
+fewer constraints then $D_2$, MILP was unable to produce an optimal solution to $D_0$ 
+in 12 hours.  This underlines up a key trade off
 between the solutions, namely, quality of solutions vs. runtime predictability.
-While the MILP solution will produce optimal solutions given enough time, it is
-very hard to predict how long it will take to produce an optimal solution. On
-the otherhand, there are no guarantees that can be made about GD in terms of
-the quality of solutions but its runtime is very predictable, scaling roughly
+While MILP will produce an optimal solution given enough time and memory, it is
+very hard to predict how long it will take to produce an optimal solution. 
+In contrast, there are no guarantees that can be made about GD in terms of
+the quality of solutions but its runtime is very predictable, scaling
 linearly with the number of constraints times the number of columns.
@@
 
-This then brings us to our second point which is the scalability of the 
-solutions. To begin, from $D_4$ and $D_5$, we can clearly see that 
+This brings us to our second major point which is the scalability of the 
+solutions. To begin, from examining the results for $D_4$ and $D_5$, we can clearly see that 
 GD has superior scaling properties when compared to MILP in terms of 
-runtime. MILP timed out on both datasets before even begining to refine the 
-solutions produced. We attribute this again to the vastly different complexity of the 
-two solutions. While GD has linear complexity in the number of constraints, 
-MILP has best case polynomial complexity in the number of constraints.  For small 
+runtime. On both datasets, MILP timed out on both datasets before even beginning to refine the 
+solutions produced, where GD produced a feasible solution in less than two minutes.
+We attribute this disparity to the vastly different computational complexity of the 
+two algorithms. GD has linear complexity in the number of constraints in contrast to
+MILP which has best case polynomial complexity in the number of constraints.  For small 
 problems this not an issue, but as the problem size increases MILP quickly becomes
-impractical for even approximating the optimal solution.
+impractical even for approximating the optimal solution. Additionally, in order to 
+make the comparison fair, we ran both solutions on the CPU, however there is nothing 
+to prevent GD from being run on a GPU. Leveraging a GPU has the potential to greatly decrease the 
+iteration time on large datasets, which would decrease the time to produce a feasible solution 
+and likely improve the quality of results.
 
@@
 
-In this paper we have only taken a small look at the possible 
-optimization techniques that could be applied. The most obvious 
-variation that could tried is to do stochastic gradient descent. This 
-variation could have two major potential benefits. First, it can 
-decrease both runtime and memory requirements of the solution further by no longer requiring the 
-entire dataset be processed on each gradient update. Second, the stochastic 
-version may be able to find better solutions as it is much more likely that it 
-could escape worse local minima, instead of relying mostly on having a good starting point.
+In this paper we have only taken a small look at the possible optimization
+techniques that could be applied. The most obvious variation that could experimented with
+is stochastic gradient descent. This variation has the potential for two major
+potential benefits. First, it can decrease both runtime and memory requirements
+of the solution further by no longer requiring the entire dataset be processed
+on each gradient update. Second, the stochastic version may be able to find
+better solutions as it is much more likely that it could escape worse local
+minima, instead of relying mostly on having a good starting point.
 
@@
 would potentially be solved by applying similar techniques, which current
-solvers are not able to handle in an capacity. We believe that it is likely
-that modifying the optimzation procedure, such as using stochastic gradient
+solvers are not able to handle in any capacity. We believe that it is likely
+that modifying the optimization procedure, such as using stochastic gradient
 descent, will likely be required to provide good approximations of such
@@
 
-In this work we have taken brief look at a particular kind of CSP, namely, 
-numerical MAX-CSP's. We motivated this problem with a real world use case for 
-tuning full text search engines. We then proposed a solution based on gradient 
-descent and gave preliminary evidence to show that it has some key advantages 
-over the baseline MILP based solution of the problem, especially for large problem 
-sizes where commercial solvers are too slow to practical. We believe that our solution 
-shows promise for finding approximate solutions to problems which were not previously tractable
-with traditional MILP based solutions. Future work should continue to explore the 
-possible variations and possibly apply similar techniques to related problems.
+In this work we have taken brief look at numerical MAX-CSPs, motivated by a
+real world use case for tuning full text search engines with labeled data. We
+then proposed an algorithm based on gradient descent and ran preliminary
+experiments that show it has some key advantages over the baseline MILP based
+solution, especially for large problem instances, where commercial solvers are
+too slow to be practical. We believe that our solution shows promise for
+finding approximate solutions to problem instances which were not previously
+tractable with traditional optimization techniques. Future work should continue
+to explore the possible variations of the gradient descent algorithm and
+explore applying the technique to related problems.
 

### 2022-05-04 proj/graph.py

@@ -18,6 +18,7 @@ LP = DATA_DIR / 'LP.json'
 TORCH = DATA_DIR / 'torch.json'
 argp = ArgumentParser()
 argp.add_argument('--show_graph', action='store_true')
+argp.add_argument('--graph3', action='store_true')
 
 DF = None
 
@@ -112,6 +113,35 @@ def graph_six(milp, torch):
 
     plt.show()
 
+def graph_three(milp, torch):
+    time_series = list(iter_time_series(milp, torch))[::2]
+    fig, axes = plt.subplots(1, 3, figsize=(22, 5))
+    sizes = ['7.7k', '111.9k', '3.7M']
+
+    
+    for ax, ts_data, i, sz in zip(axes.flatten(), time_series, range(0, 10, 2), sizes):
+        name, ts = ts_data
+        ts.plot(ax=ax, logx=False)
+        ax.set_title(f'$D_{i}$')
+        ax.set_xlabel('Time (seconds)')
+        ax.set_ylabel('Percent of Constraints Violated')
+        ax.legend()
+        text = "Time to First Solution"
+        for c in ts.columns:
+            u = ts[c].drop_duplicates()
+            ax.scatter(u.index, u.values)
+            # first soln
+            x = u.first_valid_index()
+            text += f"\n{c} = {x} seconds"
+        text += f'\n\nNumber of Constraints = {sz}'
+        ax.annotate(text, (.5, .5), xycoords='axes fraction')
+            
+            
+    
+    fig.tight_layout()
+    fig.savefig('/home/derek/slides_fig.png')
+
+    plt.show()
         
 
 def main(args):
@@ -120,8 +150,11 @@ def main(args):
     if args.show_graph:
         graph_six(milp, torch)
     else:
-        graph(milp, torch, args.show_graph)
-
+        pass
+        #graph(milp, torch, args.show_graph)
+    
+    if args.graph3:
+        graph_three(milp, torch)
 
 if __name__ == '__main__':
     main(argp.parse_args())

### 2022-05-05 proj/paper/paper.tex
save second edit

@@
 
-\title{Gradient Descent for Numerical MAX-CSP} % Title
+\title{Gradient Descent for Numerical MAX-CSPs} % Title
 \author{
@@
 problems. As such, large amounts of work have been
-devoted to studying CSPs and how to solve them, leading to many exact and
+devoted to studying CSPs and how to solve them, yielding many exact and
 approximation algorithms.  In this paper we examine a lesser studied variant of
-CSPs, called numerical MAX-CSPs. In this variant each variable is a real number
+CSP, called numerical MAX-CSPs. In this variant each variable is a real number
 and all of the constraints are linear inequalities, the goal then is to satisfy
@@
 a variety real world problems, such as solving over-constrained LP's and 
-selecting weights for full text search engines.
+tuning full text search engines.
 
@@
 on the problem and the basis for our proposed solution. Next, we motivate our particular
-problem by tying it to tuning a keyword search engine with labeled data. We
+problem by tying it to an application of tuning a keyword search engine with labeled data. We
 then move on to describe a baseline solution by modeling the problem as mixed
@@
 In this section we give a brief overview gradient descent which
-is the basis of our proposed solution. We then give
+is the basis of our proposed algorithm. We then give
 an overview of constraint satisfaction problems,
@@
 If $f(x)$ is convex, then gradient descent is guaranteed to find an optimal
-solution. If $f(x)$ is not convex, the gradient descent will converge to a
+solution (given that $\eta$ is step to an appropriate constant). If $f(x)$ is not convex, the gradient descent will converge to a
 local minimum. Although in the non-convex case, gradient descent is not
@@
 This makes it an invaluable tool for optimizing functions which don't admit
-exact optimization techniques such as those used in linear programs and
+exact optimization techniques such as those used for linear and
 quadratic programs.
@@
 
-Constraint satisfaction problems (CSPs) encompasses many different problems. Formally, a 
+Constraint satisfaction problems (CSPs) at a high level are problems that involve
+finding a valid assignment of variables which satisfy a set of constraints.  Formally, a 
 CSP is defined as a triple $\langle X, D, C \rangle$ \cite{wiki_CSP}, where 
@@
 
-A classic example of a CSP, is the 3-SAT problem. In this case, 
+A classic example of a CSP, is the 3-SAT problem. In this problem, 
 $X_i$ corresponds to a boolean variable, which has domain $D_i = \{true, false\}$
@@
 MAX-CSP. More generally, a MAX-CSP is simply a CSP where the objective is to 
-satisfy as many constraints as possible instead of being required to satisfy all 
-of them.
+satisfy as many constraints as possible as opposed to being required to satisfy all 
+constraints.
 
@@
 many as possible. While the algorithm does
-produce optimal solutions, the algorithm is worst case exponential time.
+produce optimal solutions, the algorithm is worst case exponential time, making
+it infeasible for large problems.
 
@@
 used by many MILP solvers. These improvements typically aim to provide better
-branching heuristics for branch and bound algorithms, such as strong branching
+branching heuristics, such as strong branching
 \cite{strong_branching_paper}. More recent work has examined machine learning
@@
 branching using machine learning (to reduce the computation cost) or to learn
- completely new branching heuristics.
+ completely new branching heuristics which will reduce the number of nodes 
+ explored to find an optimal solution.
 
@@
 interested in one particular problem, which is tuning the weights for
-full text search engines.  Full text search engines are used in a plethora of
-applications. These search engines (such as Apache Lucene) are built for
+full text search engines.  Full text search engines are the backbone of 
+many data management and web applications and ergo are a ubiquitous in 
+everyday life. These search engines (such as Apache Lucene) are built for
 efficient retrieval of top-k documents based on a TF/IDF based scoring metric
@@
 where, $d_i \in \mathds{R}^n$ and $l \in \{relevant, irrelevant\}$.
-We wish to find a weight vector $w$ that minimizes the number of irrelevant documents that are examined 
+We wish to find a weight vector $w$ that minimizes the number of irrelevant documents (search results) that are examined 
 before the relevant document is found. This problem can be formulated as a mixed integer linear program (MILP) as follows,
@@
 to produce an optimal answer.  This solution to our problem, which we will
-refer to as MILP when there is no ambiguity, is appealing because, assuming that the solver is
-correct, it will give an optimal solution, however it suffers from a few major
-drawbacks.  First and foremost, depending on the input, the runtime time is worst case
-exponential in the number of constraints.  Moreover, we found that even
-commercial solvers can struggle with large problem sizes, either taking large
-amounts of time to produce any feasible solution or simply not terminating in a
-reasonable amount of time. For our motivating use case, we would ideally feed
-in many labeled queries, each producing tens of constraints, hence exponential
-runtime in the number of constraints greatly reduces the usefulness of the
-solution.
+refer to as MILP when there is no ambiguity, is appealing because it will give
+an optimal solution, however it suffers from a few major drawbacks.  First and
+foremost, depending on the input, the runtime time is worst case exponential in
+the number of constraints.  Moreover, we found that even commercial solvers can
+struggle with large problem sizes, either taking large amounts of time to
+produce any feasible solution or simply not terminating in a reasonable amount
+of time. For our motivating use case, we would ideally feed in many labeled
+queries, each producing tens of constraints, hence exponential runtime in the
+number of constraints greatly reduces the usefulness of the solution.
 
@@
 $$
+
+By applying HardTanh, we change the 0-1 loss from the MILP formulation into a continuous differentiable
+function, which allows for gradient descent to be applied to the function.
 
@@
 			\If{$f(w) < f(w^*)$}
-				\State $w^* \gets w$ \Comment{Update best solution}
+				\State $w^* \gets w$ \Comment{Update best solution every 25 gradient updates}
 			\EndIf
@@
 We ran our experiments on a server with an AMD Ryzen Threadripper 2950X
-(16c/32t) processor with 64GB of RAM. Our MAX-CSP instances are generated from
-real world datasets, with the number of constraints ranging from 7.7K to 6.0M.
-For MILP we leverage Gurobi, a commercial solver with an
-academic license. We choose Gurobi because it is consistently one of the top
-performing solvers \cite{benchmark} with an academic license. Both solutions where given a
-timeout of 2700 seconds (corresponding to 24 hours of CPU time with 32 threads).
+(16c/32t) processor with 64GB of RAM running pop\_os 18.04. Our numerical MAX-CSP
+instances are generated from real world datasets, with the number of
+constraints ranging from 7.7K to 6.0M.  For MILP we leverage Gurobi, a
+commercial solver with an academic license. We choose Gurobi because it is
+consistently one of the top performing solvers \cite{benchmark} with an
+academic or free license. Both solutions where given a timeout of 2700 seconds,
+corresponding to 24 hours of CPU time with 32 threads.
 
@@
 This brings us to our second major point which is the scalability of the
-solutions. To begin, from examining the results for $D_4$ and $D_5$, we can clearly see that 
-GD has superior scaling properties when compared to MILP in terms of 
-runtime. On both datasets, MILP timed out on both datasets before even beginning to refine the 
-solutions produced, where GD produced a feasible solution in less than two minutes.
-We attribute this disparity to the vastly different computational complexity of the 
-two algorithms. GD has linear complexity in the number of constraints in contrast to
-MILP which has best case polynomial complexity in the number of constraints.  For small 
-problems this not an issue, but as the problem size increases MILP quickly becomes
-impractical even for approximating the optimal solution. Additionally, in order to 
-make the comparison fair, we ran both solutions on the CPU, however there is nothing 
-to prevent GD from being run on a GPU. Leveraging a GPU has the potential to greatly decrease the 
-iteration time on large datasets, which would decrease the time to produce a feasible solution 
-and likely improve the quality of results.
+solutions. Examination of the results for $D_4$ and $D_5$, we can clearly see
+that GD has superior scaling properties when compared to MILP in terms of
+runtime. On both datasets, MILP timed out on both datasets before even
+beginning to refine the solution produced, where GD produced a feasible
+solution in less than two minutes.  We attribute this disparity to the vastly
+different computational complexity of the two algorithms. GD has linear
+complexity in the number of constraints in contrast to MILP which has best case
+polynomial complexity in the number of constraints. For small problems this not
+an issue, but as the problem size increases MILP quickly becomes impractical
+even for approximating the optimal solution. Additionally, in order to make the
+comparison fair, we ran both solutions on the CPU, however there is nothing to
+prevent GD from being run on a GPU. Leveraging GPU compute has the potential to
+greatly decrease the iteration time on larger datasets, which would decrease
+the time to produce a feasible solution and likely improve the quality of
+results.
 
@@
 is stochastic gradient descent. This variation has the potential for two major
-potential benefits. First, it can decrease both runtime and memory requirements
+benefits. First, it can decrease both runtime and memory requirements
 of the solution further by no longer requiring the entire dataset be processed
@@
 constraints, however in principle there is no reason why this technique could
-not be applied to arbitrary constraints. In particular, non-convex constraints
+not be applied to more complex constraints. In particular, non-convex constraints
+(e.g. polynomials of degree greater than 2)
 would potentially be solved by applying similar techniques, which current

### 2022-05-06 proj/paper/paper.tex
remove homeworks

@@
 devoted to studying CSPs and how to solve them, yielding many exact and
-approximation algorithms.  In this paper we examine a lesser studied variant of
+approximation algorithms \cite{approx_for_csp_paper}. In this paper we examine a lesser studied variant of
 CSP, called numerical MAX-CSPs. In this variant each variable is a real number
@@
 recognition. Despite the focus on machine learning, gradient descent is a
-general purpose optimization procedure, capable of optimizing arbitrary
+general purpose optimization procedure capable of optimizing arbitrary
 differentiable functions. Many variations have been proposed in recent years,
@@
 current point $x$, compute the gradient $\nabla(f(x))$
-and update $x$ taking a step in a direction of $-\eta \nabla(f(x))$. Repeat
+and update $x$ by taking a step in a direction of $-\eta \nabla(f(x))$. Repeat
 this process until the minimum of the function is found.
 
-If $f(x)$ is convex, then gradient descent is guaranteed to find an optimal
-solution (given that $\eta$ is step to an appropriate constant). If $f(x)$ is not convex, the gradient descent will converge to a
+If $f(x)$ is convex, then gradient descent is guaranteed to find the global minimum of the function
+(given that $\eta$ is set to an appropriate constant). If $f(x)$ is not convex, then gradient descent will converge to a
 local minimum. Although in the non-convex case, gradient descent is not
@@
 for better branching heuristics \cite{bengio_lodi_prouvost_2021}
-\cite{liberto_kadioglu_leo_malitsky_2016}, either to approximate the strong
+\cite{liberto_kadioglu_leo_malitsky_2016}, either to approximate strong
 branching using machine learning (to reduce the computation cost) or to learn
@@
 full text search engines.  Full text search engines are the backbone of 
-many data management and web applications and ergo are a ubiquitous in 
+many data management and web applications and ergo are ubiquitous in 
 everyday life. These search engines (such as Apache Lucene) are built for
@@
 $$
-f(w) = \langle \text{HardTanh}(Aw - \gamma), \overrightarrow{1}\rangle 
+f(w) = \langle \text{HardTanh}(Aw), \overrightarrow{1}\rangle 
 $$
@@
 corresponding to 24 hours of CPU time with 32 threads.
-
 \begin{table}[ht!]
+	\centering
 \caption{Dataset statistics}
@@
          $D_0$ &                   7.7K &                 14 &         66.24 \\
+		 \hline
          $D_1$ &                  19.3K &                 18 &         61.46 \\
+		 \hline
          $D_2$ &                 111.9K &                 18 &         72.83 \\
+		 \hline
          $D_3$ &                 196.2K &                 30 &         51.62 \\
+		 \hline
          $D_4$ &                3.7M    &                124 &         25.09 \\
+		 \hline
          $D_5$ &                6.0M    &                 22 &         68.58 \\
@@
 benefits. First, it can decrease both runtime and memory requirements
-of the solution further by no longer requiring the entire dataset be processed
+of the solution by no longer requiring the entire dataset be processed
 on each gradient update. Second, the stochastic version may be able to find
